### 2013-09-26 Paper/paper.tex
Working rely/guarantee calculus for infinite traces (I think)

\documentclass{llncs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{graphicx,color}
\usepackage{multicol}
\usepackage{bussproofs}
\usepackage{float}

\begin{document}

\title{TACAS Paper}

\author{Alasdair Armstrong \and Victor B.~F.~Gomes \and Georg Struth}

\institute{Department of Computer Science, University of Sheffield, UK\\
\email{$\{$a.armstrong,v.gomes,g.struth$\}$@dcs.shef.ac.uk}}

\maketitle

\begin{abstract}
  TODO
\end{abstract}

\pagestyle{plain}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{align}
1&\le r\\
r\|r &= r\\
r\|xy &= rx\|ry\\
(r\|x)^+ &= r\|x^+
\end{align}

$r, g \vdash \{p\}x\{q\} \longleftrightarrow p(r\|x)\le q \land x \le g$

\begin{prooftree}
\LeftLabel{Weakening}
\AxiomC{$r' \le r$}
\AxiomC{$g \le g'$}
\AxiomC{$p \le p'$}
\AxiomC{$r', g' \vdash \{p'\}x\{q'\}$}
\AxiomC{$q' \le q$}
\QuinaryInfC{$r, g \vdash \{p\}x\{q\}$}
\end{prooftree}

\begin{prooftree}
\LeftLabel{Sequential}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{q\}y\{s\}$}
\BinaryInfC{$r, g \vdash \{q\}xy\{s\}$}
\end{prooftree}

\begin{prooftree}
\LeftLabel{Parallel}
\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
\AxiomC{$g_1 \le r_2$}
\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
\AxiomC{$g_2 \le r_1$}
\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
\end{prooftree}

\begin{prooftree}
\LeftLabel{Choice}
\AxiomC{$r, g \vdash \{p\}x\{q\}$}
\AxiomC{$r, g \vdash \{p\}y\{q\}$}
\BinaryInfC{$r, g \vdash \{p\}x + y\{q\}$}
\end{prooftree}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}

\end{document}

### 2013-10-02 Paper/paper.tex
Added more proofs to Pointwise.thy
Added title to paper
Added counterexample
Added rg section to paper

@@
 \usepackage{float}
+\usepackage{caption}
+\usepackage{subcaption}
+
+\usetikzlibrary{arrows,shapes,calc}
+
+\newtheorem{lem}{Lemma}
 
@@
 
-\title{TACAS Paper}
+\tikzstyle{elem} = [circle]
+\tikzstyle{line} = [draw,thick, -latex']
+
+\title{Algebraic Principles for Concurrency Verification Tools}
 
@@
 \begin{abstract}
-  TODO
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \section{Introduction}
 
+\cite{hoare_concurrent_2011}
+\cite{kozen_completeness_1994}
+\cite{kozen_kleene_1997}
+\cite{armstrong_kleene_2013}
+\cite{nipkow_isabelle/hol:_2002}
+
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Kleene Algebra}
 
-\begin{align}
-1&\le r\\
-r\|r &= r\\
-r\|xy &= rx\|ry\\
-(r\|x)^+ &= r\|x^+
-\end{align}
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Rely/Guarantee}
 
@@
 
+\newpage
+\section{Algebra for Rely/Guarantee}
+
+Define a rely/guarantee algebra as a structure
+$(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
+lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
+$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
+subset of relys/guarantees which satisfy the following axioms
+\begin{align}
+r\|r &\le r, \label{rg1}\\
+r &\le r\|r', \label{rg2}\\
+r\|xy &= (r\|x)(r\|y), \label{rg3}\\
+r\|x^+ &\le (r\|x)^+ \label{rg4}.
+\end{align}
+By convention, we use $r$ and $g$ to refer to elements of $RG$, and
+$x,y,z$ for arbitrary elements of $K$. Some elementary consequences of these rules are as follows
+\begin{align*}
+1 &\le r,\\
+r^\star = rr &= r = r\|r,\\
+r\|x^+ &= (r\|x)^+.
+\end{align*}
+
+Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
+show this, we use Isabelle's \emph{Nitpick} counterexample generator
+to construct models which violate each of these axioms, yet satisfy
+all others. The models thus constructed for (\ref{rg1}) and
+(\ref{rg3}) are shown in figures \ref{fig:rg1} and \ref{fig:rg3}
+respectively. The model constructed for (\ref{rg2}) is just the two
+element rely/guarantee algebra with $0$ and $1$.
+
+\begin{figure}[H]
+\centering
+\begin{subfigure}{0.24\textwidth}
+\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
+  \node (center) {};
+  \node [elem] (r1) at (90:1) {$r_1$};
+  \node [elem] (one) at (200:1) {$1$};
+  \node [elem] (r2) at (20:1) {$r_2$};
+  \node [elem] (zero) at (-90:1) {$0$};
+
+  \path [line] (r1) -- (r2);
+  \path [line] (r1) -- (one);
+  \path [line] (r1) -- (zero);
+  \path [line] (r2) -- (one);
+  \path [line] (r2) -- (zero);
+  \path [line] (one) -- (zero);
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}{0.24\textwidth}
+\begin{align*}
+r_1 \| r_1 &= r_1\\
+r_1 \| r_2 &= r_1\\
+r_2 \| r_1 &= r_1\\
+r_2 \| r_2 &= r_1
+\end{align*}
+\end{subfigure}
+\begin{subfigure}{0.24\textwidth}
+\begin{align*}
+r_1r_1 &= r_1\\
+r_1r_2 &= r_1\\
+r_2r_1 &= r_1\\
+r_2r_2 &= r_2
+\end{align*}
+\end{subfigure}
+\begin{subfigure}{0.24\textwidth}
+\begin{align*}
+r_1^\star &= r_1\\
+r_2^\star &= r_2
+\end{align*}
+\end{subfigure}
+\caption{4 element counterexample for $r \in RG \implies r\|r \le r$}
+\label{fig:rg1}
+\end{figure}
+
+Axiom (\ref{rg4}) can be derived from (\ref{rg3}) in all finite
+models. This is because any finite $K$ is a weak left quantale with
+respect to both sequential and parallel composition. In a quantale,
+fixpoint fusion laws can be used to prove that
+\begin{align*}
+r\|x^+ = r\|(\mu y.\; x + xy) = \mu y.\; r\|x + r\|xy = (r\|x)^+.
+\end{align*}
+Thus it is impossible to construct a model which demonstrates
+that (\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg2}).
+
+\begin{figure}
+\centering
+\begin{subfigure}{0.24\textwidth}
+\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
+  \node (center) {};
+  \node [elem] (r1) at (90:1) {$r_1$};
+  \node [elem] (one) at (0:1) {$1$};
+  \node [elem] (zero) at (-90:1) {$0$};
+
+  \path [line] (r1) -- (one);
+  \path [line] (r1) -- (zero);
+  \path [line] (r1) -- (zero);
+  \path [line] (one) -- (zero);
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}{0.24\textwidth}
+\begin{align*}
+r_1 \| r_1 &= r_1\\
+r_1 r_1 &= r_1\\
+r_1^\star &= r_1
+\end{align*}
+\end{subfigure}
+\caption{3 element counterexample for $r \in RG \implies r\|xy = (r\|x)(r\|y)$}
+\label{fig:rg3}
+\end{figure}
+
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Language Model}
 
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Example}
+
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusion}
 
+\bibliography{paper}{}
+\bibliographystyle{plain}
+
 \end{document}

### 2013-10-03 Paper/paper.tex
Started on language model section

@@
 
+\DeclareFontFamily{U}{wncy}{}
+\DeclareFontShape{U}{wncy}{m}{n}{<->wncyss10}{}
+\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
+\DeclareMathSymbol{\sh}{\mathord}{mcy}{"78} 
+\DeclareMathOperator{\sha}{\sh}
+\DeclareMathOperator{\tr}{Tr}
+
+\newcommand{\udagger}{\text{$\rotatebox[origin=c]{180}{$\dagger$}$}}
+
 \newtheorem{lem}{Lemma}
@@
 \tikzstyle{line} = [draw,thick, -latex']
+\tikzstyle{rel} = [draw,thin,dashed, -latex']
 
@@
 
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+related work?
+
 \newpage
@@
 \section{Kleene Algebra}
+\label{sec:KA}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Rely/Guarantee}
-
-$r, g \vdash \{p\}x\{q\} \longleftrightarrow p(r\|x)\le q \land x \le g$
-
-\begin{prooftree}
-\LeftLabel{Weakening}
-\AxiomC{$r' \le r$}
-\AxiomC{$g \le g'$}
-\AxiomC{$p \le p'$}
-\AxiomC{$r', g' \vdash \{p'\}x\{q'\}$}
-\AxiomC{$q' \le q$}
-\QuinaryInfC{$r, g \vdash \{p\}x\{q\}$}
-\end{prooftree}
-
-\begin{prooftree}
-\LeftLabel{Sequential}
-\AxiomC{$r, g \vdash \{p\}x\{q\}$}
-\AxiomC{$r, g \vdash \{q\}y\{s\}$}
-\BinaryInfC{$r, g \vdash \{q\}xy\{s\}$}
-\end{prooftree}
-
-\begin{prooftree}
-\LeftLabel{Parallel}
-\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
-\AxiomC{$g_1 \le r_2$}
-\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
-\AxiomC{$g_2 \le r_1$}
-\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
-\end{prooftree}
-
-\begin{prooftree}
-\LeftLabel{Choice}
-\AxiomC{$r, g \vdash \{p\}x\{q\}$}
-\AxiomC{$r, g \vdash \{p\}y\{q\}$}
-\BinaryInfC{$r, g \vdash \{p\}x + y\{q\}$}
-\end{prooftree}
-
-\newpage
 \section{Algebra for Rely/Guarantee}
+\label{sec:RG}
 
@@
 Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
-show this, we use Isabelle's \emph{Nitpick} counterexample generator
+show this, we use Isabelle's
+\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample generator
 to construct models which violate each of these axioms, yet satisfy
 all others. The models thus constructed for (\ref{rg1}) and
-(\ref{rg3}) are shown in figures \ref{fig:rg1} and \ref{fig:rg3}
+(\ref{rg3}) are shown in Figures \ref{fig:rg1} and \ref{fig:rg3}
 respectively. The model constructed for (\ref{rg2}) is just the two
@@
 respect to both sequential and parallel composition. In a quantale,
-fixpoint fusion laws can be used to prove that
+fixpoint fusion and induction laws can be used to prove that
 \begin{align*}
-r\|x^+ = r\|(\mu y.\; x + xy) = \mu y.\; r\|x + r\|xy = (r\|x)^+.
+r\|x^+ = r\|(\mu y.\; x + xy) = \mu y.\; r\|x + r\|xy \le (r\|x)^+.
 \end{align*}
-Thus it is impossible to construct a model which demonstrates
+Thus it is impossible to construct a finite model demonstrating
 that (\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg2}).
@@
 
+Jones quintuples can be encodeded in this setting as
+\begin{align}
+r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g.
+\end{align}
+Using this encoding we can derive all the standard rules for
+rely/guarantee, shown in Figure \ref{fig:rgrules}.
+
+We can expand our rely/guarantee algebras to have residuals
+axiomatised by Galois connections for both parallel and sequential
+composition, i.e.
+\begin{align}
+xy \le z &\iff y \le x \rightarrow z, \label{res1}\\
+xy \le z &\iff x \le z \leftarrow y, \label{res2}\\
+x\|y \le z &\iff y \le x/z \label{res3}.
+\end{align}
+For sequential composition, (\ref{res1}) and (\ref{res2}) essentially
+provide a link between weakest-precondition style semantics and hoare
+logic. As mentioned in Section \ref{sec:KA}, the triple $\{p\} c
+\{q\}$ can be encoded as $pc \le q$, while $q \leftarrow c$ represents
+$wlp(c,q)$. Equation (\ref{res2}) can thus be written as $\{p\} c
+\{q\} \iff p \le wlp(c,q)$. The parallel residual $x/z$ can be
+understood similarly as follows; it is the weakest program such that
+when placed in parallel with $x$, the composition behaves as $z$.
+
+By using residuals quintuples can be encoded in the following
+equivalent way,
+\begin{align}
+r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
+\end{align}
+This encoding allows us to think in terms of program
+refinement, as $r/(p \rightarrow q) \sqcap g$ defines the weakest
+program that when placed in parallel with interference from $r$, and
+guaranteeing interference at most $g$, goes from $p$ to $q$.
+
+\begin{figure}[tbh]
+\centering
+\begin{prooftree}
+\RightLabel{Skip}
+\AxiomC{$pr \le p$}
+\UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
+\end{prooftree}
+
+\begin{prooftree}
+\RightLabel{Weakening}
+\AxiomC{$r' \le r$}
+\AxiomC{$g \le g'$}
+\AxiomC{$p \le p'$}
+\AxiomC{$r', g' \vdash \{p'\}x\{q'\}$}
+\AxiomC{$q' \le q$}
+\QuinaryInfC{$r, g \vdash \{p\}x\{q\}$}
+\end{prooftree}
+
+\begin{prooftree}
+\RightLabel{Sequential}
+\AxiomC{$r, g \vdash \{p\}x\{q\}$}
+\AxiomC{$r, g \vdash \{q\}y\{s\}$}
+\BinaryInfC{$r, g \vdash \{q\}xy\{s\}$}
+\end{prooftree}
+
+\begin{prooftree}
+\RightLabel{Parallel}
+\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
+\AxiomC{$g_1 \le r_2$}
+\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
+\AxiomC{$g_2 \le r_1$}
+\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
+\end{prooftree}
+
+\begin{prooftree}
+\RightLabel{Choice}
+\AxiomC{$r, g \vdash \{p\}x\{q\}$}
+\AxiomC{$r, g \vdash \{p\}y\{q\}$}
+\BinaryInfC{$r, g \vdash \{p\}x + y\{q\}$}
+\end{prooftree}
+
+\begin{prooftree}
+\RightLabel{Star}
+\AxiomC{$pr \le p$}
+\AxiomC{$r, g \vdash \{p\}x\{p\}$}
+\BinaryInfC{$r, g \vdash \{p\}x^\star\{p\}$}
+\end{prooftree}
+\caption{Rely/guarantee inference rules}
+\label{fig:rgrules}
+\end{figure}
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
+
+x
 \newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Language Model}
+\section{Concurrent Kleene Algebra with Tests}
+
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Finite Language Model}
+
+In this section, we construct a finite language model satisfying the
+axioms in Section \ref{sec:RG}. Restricting our attention to finite
+languages means that we do not need to concern ourselves with
+termination side-conditions.
+
+The shuffle product $\sha$ on a pair of finite words is defined inductively as
+\begin{align*}
+\epsilon \sha s &= \{s\},\\
+s \sha \epsilon &= \{s\},\\
+as \sha bt &= a(s \sha bt) \cup b(as \sha t).
+\end{align*}
+Shuffle is associative, commutative, and distributes over arbitrary
+joins.
+
+\begin{lem}
+$(\mathcal{P}((\Sigma\times\Sigma)^*), \cup, \cdot, \sha, \emptyset, \{\epsilon\})$ is a weak trioid.
+\end{lem}
+
+Our alphabet comprises relations ($R,S,T,\dots$) over a state space
+$\Sigma$. An example of a word in
+$(\{\sigma_1,\sigma_2,\sigma_3\}^2)^*$ is shown below. This word only
+has a single valid \emph{trace}, which is
+$\sigma_1\rightarrow\sigma_1\rightarrow\sigma_2\rightarrow\sigma_3\rightarrow\sigma_2$.
+We define an operator $\tr(X)$ which returns all the traces in a language
+$X$.
+
+\begin{center}
+\begin{tikzpicture}[x=1cm,auto]
+  \node (center) {};
+  \node [elem] (s12) {$\sigma_2$};
+  \node [elem, above of=s12] (s11) {$\sigma_1$};
+  \node [elem, below of=s12] (s13) {$\sigma_3$};
+  \node [right of=s13, node distance=0.75cm] (r1) {};
+  \node [below of=r1, node distance=0.5cm] (t4) {R};
+
+  \node [elem, right of=s12, node distance=1.5cm] (s22) {$\sigma_2$};
+  \node [elem, above of=s22] (s21) {$\sigma_1$};
+  \node [elem, below of=s22] (s23) {$\sigma_3$};
+  \node [right of=s23, node distance=0.75cm] (r2) {};
+  \node [below of=r2, node distance=0.5cm] (t4) {S};
+
+  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
+  \node [elem, above of=s32] (s31) {$\sigma_1$};
+  \node [elem, below of=s32] (s33) {$\sigma_3$};
+  \node [right of=s33, node distance=0.75cm] (r3) {};
+  \node [below of=r3, node distance=0.5cm] (t4) {T};
+
+  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
+  \node [elem, above of=s42] (s41) {$\sigma_1$};
+  \node [elem, below of=s42] (s43) {$\sigma_3$};
+  \node [right of=s43, node distance=0.75cm] (r4) {};
+  \node [below of=r4, node distance=0.5cm] (t4) {U};
+
+  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
+  \node [elem, above of=s52] (s51) {$\sigma_1$};
+  \node [elem, below of=s52] (s53) {$\sigma_3$};
+
+  \path [line] (s11) -- (s21);
+  \path [line] (s21) -- (s32);
+  \path [line] (s32) -- (s43);
+  \path [line] (s43) -- (s52);
+
+  \path [rel] (s11) -- (s22);
+  \path [rel] (s12) -- (s22);
+  \path [rel] (s23) -- (s32);
+  \path [rel] (s32) -- (s42);
+  \path [rel] (s42) -- (s51);
+  \path [rel] (s41) -- (s52);
+\end{tikzpicture}
+\end{center}
+
+Following~\cite{brookes_full_1993} and~\cite{dingel_refinement_2002},
+we inductively generate the stutter/mumble language $w^\dagger$ for a
+word $w$ in a language over $\Sigma^2$ as follows. Assume $R,S,T \in
+\Sigma^2$ and $u,v,w \in (\Sigma^2)^*$. First, $Iw \in
+w^\dagger$. Secondly, if $uRv \in x^\dagger$ then $uIRv \in
+x^\dagger$ and $uRIv \in x^\dagger$ (\emph{stuttering}). Thirdly, if
+$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$
+(\emph{mumbling}). The stutter/mumble closure of a language is
+defined by lifting this operation to languages in the obvious way.
+
+Next, we define the following language for a word $w = R_0R_1\dots
+R_n$. The language $w^\udagger$ contains all strings $S_0S_1\dots S_n$
+such that for all $m \le n$, $S_m \subseteq R_m$. Again this operation
+is lifted to languages in the same way as before. Both these closure
+operators are combined then into a single closure operator,
+$X^\ddagger$, which satisfies the properties
+\begin{align*}
+(X^\ddagger \sha Y)^\ddagger &= (X \sha Y)^\ddagger,\\
+(X^\ddagger Y)^\ddagger = (&XY)^\ddagger = (XY^\ddagger)^\ddagger,\\
+(X \cup Y)^\ddagger &= (X^\ddagger \cup Y^\ddagger),\\
+(X^\ddagger \cap Y^\ddagger)^\ddagger &= (X^\ddagger \cap Y^\ddagger),\\
+\tr(X^\ddagger) &\subseteq X^\ddagger.
+\end{align*}
+For all binary operators $\bullet$, we can therefore define
+$\bullet_\ddagger$ as $\lambda X Y. (X^\ddagger\bullet
+Y^\ddagger)^\ddagger$ and show that
+$(\mathcal{P}((\Sigma^2)^*)^\ddagger,\cup_\ddagger,\cdot_\ddagger,\sha_\ddagger,\emptyset,{\epsilon}^\ddagger)$
+is a weak trioid.
 

### 2013-10-03 Paper/paper.tex
More paper bits

@@
 \DeclareMathOperator{\tr}{Tr}
+\DeclareMathOperator{\fin}{finite}
 
@@
 
-The shuffle product $\sha$ on a pair of finite words is defined inductively as
+As usual, the product of two languages X and Y over an alphabet
+$\Sigma$ is defined as $XY = \{x\frown y.\; x \in X \land y \in Y\}$.
+The language product is associative, and distributes over arbitrary
+joins from both the left and the right. The shuffle $\sha$ on a pair
+of finite words is defined inductively as
 \begin{align*}
@@
 s \sha \epsilon &= \{s\},\\
-as \sha bt &= a(s \sha bt) \cup b(as \sha t).
+as \sha bt &= a(s \sha bt) \cup b(as \sha t),
 \end{align*}
-Shuffle is associative, commutative, and distributes over arbitrary
-joins.
+which is then lifted to languages X and Y as the shuffle product,
+\begin{align*}
+X \sha Y = \bigcup\{xs \sha ys.\; x \in X \land x \in Y\}.
+\end{align*}
+The shuffle product is associative, commutative, and distributes over
+arbitrary joins. Both products share the same unit, $\{\epsilon\}$ and
+zero, $\emptyset$. Therefore, $(\mathcal{P}(\Sigma)^*), \cup, \cdot,
+\sha, \emptyset, \{\epsilon\})$ forms a weak trioid.
 
 \begin{lem}
-$(\mathcal{P}((\Sigma\times\Sigma)^*), \cup, \cdot, \sha, \emptyset, \{\epsilon\})$ is a weak trioid.
+$(\mathcal{P}(\Sigma)^*), \cup, \cdot, \sha, \emptyset, \{\epsilon\})$ is a weak trioid.
 \end{lem}
@@
 
+\begin{center}
+\begin{tikzpicture}[x=1cm,auto]
+  \node [elem] (s12) {$\sigma_2$};
+  \node [elem, above of=s12] (s11) {$\sigma_1$};
+  \node [elem, below of=s12] (s13) {$\sigma_3$};
+  \node [right of=s13, node distance=0.75cm] (r1) {};
+  \node [below of=r1, node distance=0.5cm] (t4) {R};
+
+  \node [elem, right of=s12, node distance=1.5cm] (i2) {$\sigma_2$};
+  \node [elem, above of=i2] (i1) {$\sigma_1$};
+  \node [elem, below of=i2] (i3) {$\sigma_3$};
+  \node [right of=i3, node distance=0.75cm] (ri) {};
+  \node [below of=ri, node distance=0.5cm] (ti) {V};
+
+  \node [elem, right of=i2, node distance=1.5cm] (s22) {$\sigma_2$};
+  \node [elem, above of=s22] (s21) {$\sigma_1$};
+  \node [elem, below of=s22] (s23) {$\sigma_3$};
+  \node [right of=s23, node distance=0.75cm] (r2) {};
+  \node [below of=r2, node distance=0.5cm] (t4) {S};
+
+  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
+  \node [elem, above of=s32] (s31) {$\sigma_1$};
+  \node [elem, below of=s32] (s33) {$\sigma_3$};
+  \node [right of=s33, node distance=0.75cm] (r3) {};
+  \node [below of=r3, node distance=0.5cm] (t4) {T};
+
+  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
+  \node [elem, above of=s42] (s41) {$\sigma_1$};
+  \node [elem, below of=s42] (s43) {$\sigma_3$};
+  \node [right of=s43, node distance=0.75cm] (r4) {};
+  \node [below of=r4, node distance=0.5cm] (t4) {U};
+
+  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
+  \node [elem, above of=s52] (s51) {$\sigma_1$};
+  \node [elem, below of=s52] (s53) {$\sigma_3$};
+
+  \path [rel] (s11) -- (i1);
+  \path [rel] (s21) -- (s32);
+  \path [line] (s32) -- (s43);
+  \path [line] (s43) -- (s52);
+  \path [line] (i2) -- (s23);
+
+  \path [rel] (s11) -- (i2);
+  \path [line] (s12) -- (i2);
+  \path [line] (s23) -- (s32);
+  \path [line] (s32) -- (s42);
+  \path [line] (s42) -- (s51);
+  \path [rel] (s41) -- (s52);
+\end{tikzpicture}
+\end{center}
+
 \newpage

### 2013-10-04 Paper/paper.tex
Added some KA background and abstract sentence

@@
 \begin{abstract}
+We investigate the laws required for the derivation of the
+rely/guarantee and owicki-gries inference rules in a minimalistic
+algebraic setting based on Kleene algebra...
 \end{abstract}
@@
 \label{sec:KA}
+
+A \emph{dioid} or \emph{idempotent semiring} is an structure
+$(D,+,\cdot,0,1)$ such that $(D,+,0)$ is a join semilattice with zero,
+$(D,\cdot,1)$ is a monoid, $+$ distributes over $\cdot$ from both the
+right and left, and $0$ is a right and left annihilator for
+$\cdot$. The natural partial order $\le$ is defined as $x \le y $ iff
+$x + y = y$.
+
+A Kleene algebra extends a dioid with a star operation satisfying the
+following laws.
+\begin{align*}
+  1 + xx^* &\le x^* &\qquad z + xy \le y &\implies x^*z \le y\\
+  1 + x^*x &\le x^* &\qquad z + yx \le y &\implies zx^* \le y
+\end{align*}
+
+In the context of programs, one typically thinks of $\cdot$ as being
+sequential composition, $+$ as non-deterministic choice, $0$ as the
+abortive action, $1$ as skip and $1$ as finite iteration. For
+concurrent software, it is natural to extend the notion of a dioid to
+that of a \emph{trioid} with operators for both sequential and
+parallel composition. One can also axiomatise a parallel star
+operation, but this seems to be a less useful operation. A
+dioid or trioid is \emph{weak} if it does not satisfy $x0 = 0$.
+
+Formally, a trioid is a structure $(D,+,\cdot,\|,0,1)$, such that
+both $(D,+,\cdot,0,1)$ and $(D,+,\|,0,1)$ are dioids. Previous work in
+this area has focused on identifying laws linking the sequential and
+parallel composition operators, notably, the exchange law
+\begin{align*}
+  (a \| b) \cdot (c \| d) \le (a \cdot d) \| (b \cdot c).
+\end{align*}
+The exchange law is appealing as it provides a way to derive many
+properties of parallel composition from a single law. For example, one
+can show that $\|$ must be commutative, that $x \cdot y \le x \| y$
 

### 2013-10-04 Paper/paper.tex
Updates to paper abstract

@@
 
-\author{Alasdair Armstrong \and Victor B.~F.~Gomes \and Georg Struth}
+\author{Alasdair Armstrong \and Victor B.~F.~Gomes\thanks{Sponsored by CNPq, Brazil} \and Georg Struth}
 
@@
 
-\begin{abstract}
-We investigate the laws required for the derivation of the
-rely/guarantee and owicki-gries inference rules in a minimalistic
-algebraic setting based on Kleene algebra...
-\end{abstract}
+We investigate equational laws for the derivation of rely-guarantee
+style, Owicki-Gries style and separation logic style inference and
+refinement rules in the minimalistic setting of semirings. The
+algebraic level is linked with concrete programming models based on
+languages and execution traces. The approach has been implemented in
+Isabelle/HOL where the link between algebras and models supports
+reasoning about both the control and data flow of concurrent
+programs. This is illustrated by simple program verification examples.
 
@@
 x
-
-related work?
 

### 2013-10-06 Paper/paper.tex
Added intuitive explanations of rules

--- 2013-10-04 Paper/paper.tex
@@
 \usepackage{float}
-\usepackage{caption}
-\usepackage{subcaption}
 
@@
 We investigate the laws required for the derivation of the
-rely/guarantee and owicki-gries inference rules in a minimalistic
+rely-guarantee and owicki-gries inference rules in a minimalistic
 algebraic setting based on Kleene algebra...
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Algebra for Rely/Guarantee}
+\section{Algebra for Rely-Guarantee}
 \label{sec:RG}
 
-Define a rely/guarantee algebra as a structure
+Define a rely-guarantee algebra as a structure
 $(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
@@
 
+%% Could use better explanations.
+
+Axiom (\ref{rg1}) states that interference from the a rely condition
+being run twice in parallel, is no different from just the
+interference from that rely condition. Axiom (\ref{rg2}) states that
+interference from a single rely condition is less than interference
+from itself and another rely condition. Axiom (\ref{rg3}) allows a
+rely or guarantee to be split across sequential programs. Axiom
+(\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent, except it deals
+with finite iteration.
+
 Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
@@
 respectively. The model constructed for (\ref{rg2}) is just the two
-element rely/guarantee algebra with $0$ and $1$.
+element rely-guarantee algebra with $0$ and $1$.
 
@@
 \centering
-\begin{subfigure}{0.24\textwidth}
+\begin{minipage}{0.24\textwidth}
 \begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
@@
 \end{tikzpicture}
-\end{subfigure}
-\begin{subfigure}{0.24\textwidth}
+\end{minipage}
+\begin{minipage}{0.24\textwidth}
 \begin{align*}
@@
 \end{align*}
-\end{subfigure}
-\begin{subfigure}{0.24\textwidth}
+\end{minipage}
+\begin{minipage}{0.24\textwidth}
 \begin{align*}
@@
 \end{align*}
-\end{subfigure}
+\end{minipage}
 \caption{4 element counterexample for $r \in RG \implies r\|r \le r$}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
-\begin{subfigure}{0.24\textwidth}
+\begin{minipage}{0.24\textwidth}
 \begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
@@
 \end{align*}
-\end{subfigure}
+\end{minipage}
 \caption{3 element counterexample for $r \in RG \implies r\|xy = (r\|x)(r\|y)$}
@@
 Using this encoding we can derive all the standard rules for
-rely/guarantee, shown in Figure \ref{fig:rgrules}.
+rely-guarantee, shown in Figure \ref{fig:rgrules}.
 
-We can expand our rely/guarantee algebras to have residuals
+We can expand our rely-guarantee algebras to have residuals
 axiomatised by Galois connections for both parallel and sequential
@@
 \end{prooftree}
-\caption{Rely/guarantee inference rules}
+\caption{Rely-guarantee inference rules}
 \label{fig:rgrules}

### 2013-10-06 Paper/paper.tex
Updated abstract

@@
 \begin{abstract}
-We investigate the laws required for the derivation of the
-rely-guarantee and owicki-gries inference rules in a minimalistic
-algebraic setting based on Kleene algebra...
+We investigate equational laws for the derivation of rely-guarantee
+style, Owicki-Gries style and separation logic style inference and
+refinement rules in the minimalistic setting of semirings. The
+algebraic level is linked with concrete programming models based on
+languages and execution traces. The approach has been implemented in
+Isabelle/HOL where the link between algebras and models supports
+reasoning about both the control and data flow of concurrent
+programs. This is illustrated by simple program verification examples.
 \end{abstract}

### 2013-10-07 Paper/paper.tex
CNPq
More on model. Owicki-Gries doesn't work

@@
 \usepackage{float}
+\usepackage{centernot}
 
@@
 \DeclareMathOperator{\fin}{finite}
+\DeclareMathOperator{\rely}{rely}
+\DeclareMathOperator{\guar}{guar}
 
@@
 
-\author{Alasdair Armstrong \and Victor B.~F.~Gomes \and Georg Struth}
+\author{Alasdair Armstrong \and Victor B.~F.~Gomes\thanks{Sponsored by CNPq, Brazil}  \and Georg Struth}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Kleene Algebra}
+\section{Semirings and Kleene Algebra}
 \label{sec:KA}
 
-A \emph{dioid} or \emph{idempotent semiring} is an structure
-$(D,+,\cdot,0,1)$ such that $(D,+,0)$ is a join semilattice with zero,
-$(D,\cdot,1)$ is a monoid, $+$ distributes over $\cdot$ from both the
-right and left, and $0$ is a right and left annihilator for
-$\cdot$. The natural partial order $\le$ is defined as $x \le y $ iff
-$x + y = y$.
-
-A Kleene algebra extends a dioid with a star operation satisfying the
-following laws.
+A \emph{semiring} is an structure
+$(D,+,\cdot,0,1)$ satisfying the following laws,
 \begin{align*}
-  1 + xx^* &\le x^* &\qquad z + xy \le y &\implies x^*z \le y\\
-  1 + x^*x &\le x^* &\qquad z + yx \le y &\implies zx^* \le y
+  x(y + z) &= xy + xz, &\qquad (x + y)z &= xz + yz,\\
+  (x + y) + z &= x + (y + z), &\qquad (xy)z &= x(yz),\\
+  x + y &= y + x, &\qquad 0 + x &= x,\\
+  1x &= x, &\qquad x1 &= x,\\
+  0x &= 0, &\qquad x0 &= 0.
+\end{align*}
+A semiring is \emph{idempotent} if $x + x = x$, and \emph{weak} if $x0
+= 0$ does not hold. The natural partial order $\le$ is defined as $x
+\le y $ iff $x + y = y$. Idempotent semirings are also called
+\emph{dioids}. In the context of programs, one typically thinks of
+$\cdot$ as being sequential composition, $+$ as non-deterministic
+choice, $0$ as the abortive action and $1$ as skip. For concurrent
+software, it is natural to extend the notion of a dioid to that of a
+\emph{trioid} with operators for both sequential and parallel
+composition. Formally, a trioid is a structure $(D,+,\cdot,\|,0,1)$
+where $(D,+,\cdot,0,1)$ is a dioid and,
+\begin{align*}
+  x\|(y + z) &= x\|y + x\|z, &\qquad x\|y &= y\|x,\\
+  (x\|y)\|z &= x\|(y\|z), &\qquad x\|1 &= x,\\
+  0\|x &= x.
 \end{align*}
 
-In the context of programs, one typically thinks of $\cdot$ as being
-sequential composition, $+$ as non-deterministic choice, $0$ as the
-abortive action, $1$ as skip and $1$ as finite iteration. For
-concurrent software, it is natural to extend the notion of a dioid to
-that of a \emph{trioid} with operators for both sequential and
-parallel composition. One can also axiomatise a parallel star
-operation, but this seems to be a less useful operation. A
-dioid or trioid is \emph{weak} if it does not satisfy $x0 = 0$.
+In a trioid there are no laws governing the interaction between
+sequential and parallel composition. One law which has been considered
+in previous papers is the \emph{exchange} law
+\begin{align*}
+  (w \| x) \cdot (y \| z) \le (w \cdot z) \| (x \cdot y).
+\end{align*}
 
-Formally, a trioid is a structure $(D,+,\cdot,\|,0,1)$, such that
-both $(D,+,\cdot,0,1)$ and $(D,+,\|,0,1)$ are dioids. Previous work in
-this area has focused on identifying laws linking the sequential and
-parallel composition operators, notably, the exchange law
+The exchange law is initially appealing as it provides a way to derive
+many properties of parallel composition from a single law. For example
+one can show that $\|$ must be commutative, or that $x\cdot y \le
+x\|y$. However, there are many interesting models which do not satisfy
+this law. First, one could consider fair parallel composition
+$x\parallel_f y$ in a model with infinite, or non-terminating
+programs. In this model, $x \cdot y \not\leq x \parallel_f y$, in the
+case where $x$ is non-terminating. A second example would be
+predicate-transformer style models for separation logic, where $\|$ is
+interpreted as separating conjunction. In such models, $\|$ and
+$\cdot$ have different units, which, in the presence of the exchange
+law implies that $\cdot$ is equal to $\|$.
+
+A Kleene algebra expands a dioid with a star operation satisfying the laws
 \begin{align*}
-  (a \| b) \cdot (c \| d) \le (a \cdot d) \| (b \cdot c).
+  1 + xx^* &\le x^*, &\qquad z + xy \le y &\implies x^*z \le y,\\
+  1 + x^*x &\le x^*, &\qquad z + yx \le y &\implies zx^* \le y.
 \end{align*}
-The exchange law is appealing as it provides a way to derive many
-properties of parallel composition from a single law. For example, one
-can show that $\|$ must be commutative, that $x \cdot y \le x \| y$
+
+For programs, the star represents finite iteration. One can similarly
+axiomatise a parallel star operation, but in practice this seems to be
+a less useful operation, although one could see it as an operator that
+spawns an arbitrary but finite number of processes.
 
@@
 $(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
-subset of relys/guarantees which satisfy the following axioms
+subset of relys and guarantees which satisfy the following axioms
 \begin{align}
@@
 Axiom (\ref{rg1}) states that interference from the a rely condition
-being run twice in parallel, is no different from just the
+being run twice in parallel is no different from just the
 interference from that rely condition. Axiom (\ref{rg2}) states that
@@
 Axiom (\ref{rg4}) can be derived from (\ref{rg3}) in all finite
-models. This is because any finite $K$ is a weak left quantale with
-respect to both sequential and parallel composition. In a quantale,
-fixpoint fusion and induction laws can be used to prove that
-\begin{align*}
-r\|x^+ = r\|(\mu y.\; x + xy) = \mu y.\; r\|x + r\|xy \le (r\|x)^+.
-\end{align*}
-Thus it is impossible to construct a finite model demonstrating
-that (\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg2}).
+models. This is because any finite $K$ is complete, and in a complete
+setting fixpoint laws can be used to show $r\|x^+ \le (r\|x)^+$. Thus
+it is impossible to construct a finite model demonstrating that
+(\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg2}).
 
@@
 \end{align}
-Using this encoding we can derive all the standard rules for
-rely-guarantee, shown in Figure \ref{fig:rgrules}.
+With this encoding we can derive the standard rely-guarantee inference rules, as
+shown in Figure \ref{fig:rgrules}.
 
@@
 \end{figure}
-x
 
-x
+% Or should this be weak?
 
-x
+While this setting is adequate for deriving the standard inference
+rules, the equality is too strong to capture many interesting
+statements about concurrent programs. Consider the statement,
+$x = y \implies x\|z = y\|z$,
+which holds in our algebra and can be read as follows; if $x$ and $y$ are
+equal, then they must be equal under all possible interference from
+any arbitrary $z$. This would seem to preclude any reasoning about
+interference in a fine-grained way, but we can work around this by
+using an \emph{interior operator}. In a rely-guarantee algebra an
+interior operator is any function $f : K \to K$ which satisfies
+\begin{align*}
+f(f(x)) &= f(x) \tag{$f$ is idempotent}\\
+x \le y \implies f(x) &\le f(y) \tag{$f$ is isotone}\\
+f(x) &\le x \tag{$f$ is coextensive}
+\end{align*}
 
-x
+We extend our rely-guarantee algebra to
+$(K,RG,+,\cdot,\|,^\star,[\--],0,1)$ by adding a single interior
+operator $[\--]$, which additionally satisfies
+\begin{align}
+[xy] &= [x][y], \label{rgimult}\\
+[x + y] &= [x] + [y],\\
+[x \sqcap y] &= [x] \sqcap [y],\\
+[x^\star] &= [x]^\star,\\
+[0] &= 0 \label{rgi0}
+\end{align}
+Intuitively $[x]$ can be thought of as the \emph{execution} of a
+program $x$. While $x$ contains all information about what the process
+or program \emph{could} do, $[x]$ only contains information about what
+it \emph{actually} does. Note that (\ref{rgimult}) -- (\ref{rgi0}) do
+not mention parallel composition, such that in general $[x] = [y]
+\centernot\implies [x\|z] = [y\|z]$. We amend our jones-quintuples to
+involve this operator as shown below.
+\begin{align*}
+r,g \vdash \{p\} c \{q\} \iff p[r\|c] \le q \land c \le g.
+\end{align*}
+All the rules in Figure \ref{fig:rgrules} can be derived with this
+encoding, and moreover their proofs remain the same, mutatis
+mutandis.
 
-x
-
-x
-
-x
-
-x
-
-x
-
-x
 \newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Concurrent Kleene Algebra with Tests}
+\section{Algebra for Owicki-Gries}
+\label{sec:OG}
+
+This algebra is really unsound!
+
+$px = xp$ is provable!
+
+Following in the style of Kozen's \emph{Kleene algebra with tests}
+(KAT), we extend a concurrent Kleene algebra in the same way, yielding
+\emph{concurrent Kleene algebra with tests} (CKAT). In this setting we
+can encode hoare triples not as $\{x\}y\{z\} \iff xy \le z$, as in
+Section \ref{sec:RG}, but instead as
+\begin{align*}
+\{p\}x\{q\} \iff px \le xq,
+\end{align*}
+where $p$ and $q$ are KAT-style tests, and not just arbitrary
+processes/programs.
+
+Formally, a concurrent Kleene algebra with tests is a structure
+\begin{align*}
+(K,B,+,\cdot,\|,\overline{\phantom{x}},\phantom{|}^\star,0,1),
+\end{align*}
+where $(K,+,\cdot,\|,\phantom{|}^\star,0,1)$ is a concurrent Kleene
+algebra and $(B,+,\cdot,\overline{\phantom{x}})$ is a Boolean subalgebra such that
+\begin{align*}
+(p\|q)(x\|y) &= px\|qy,\\ % THESE RULES MAKE NO SENSE!
+(x\|y)(p\|q) &= xp\|yq,\\
+p\|q &= pq,\\
+pp &= p,\\
+p &\le 1.
+\end{align*}
+Just as in KAT, negation is only defined for elements of the boolean
+subalgebra.
+
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Algebra for Separation Logic}
 
@@
 \section{Finite Language Model}
+\label{sec:Model}
 
@@
 axioms in Section \ref{sec:RG}. Restricting our attention to finite
-languages means that we do not need to concern ourselves with
-termination side-conditions.
+languages means we do not need to concern ourselves with non-algebraic
+termination side-conditions, nor do we need to worry about additional
+restrictions on parallel composition, e.g. fairness. However, all the
+results in this section can easily be adapted to potentially
+infinite languages, and our Isabelle/HOL formalisation includes these more
+general definitions by using coinductively defined lazy lists to
+represent words, and having a weakly-fair shuffle operator for such
+infinite languages.
 
@@
 The language product is associative, and distributes over arbitrary
-joins from both the left and the right. The shuffle $\sha$ on a pair
+joins from both the left and the right, i.e.
+\begin{align*}
+X(\bigcup Y) = \bigcup\{XZ.\; Z \in Y\} \qquad \text{and} \qquad (\bigcup Y)X = \bigcup\{ZX.\; Z \in Y\}.
+\end{align*}
+The shuffle $\sha$ on a pair
 of finite words is defined inductively as
 \begin{align*}
-\epsilon \sha s &= \{s\},\\
-s \sha \epsilon &= \{s\},\\
-as \sha bt &= a(s \sha bt) \cup b(as \sha t),
+\epsilon \sha s = \{s\}, \qquad s \sha \epsilon = \{s\},\\
+as \sha bt = a(s \sha bt) \cup b(as \sha t),
 \end{align*}
@@
 arbitrary joins. Both products share the same unit, $\{\epsilon\}$ and
-zero, $\emptyset$. Therefore, $(\mathcal{P}(\Sigma)^*), \cup, \cdot,
-\sha, \emptyset, \{\epsilon\})$ forms a weak trioid.
+zero, $\emptyset$. Proving properties of shuffle is suprisingly
+tricky (especially if one considers infinite words). For a in-depth
+treatment of the shuffle product see~\cite{shufflethings}. The
+above properties are enough to show that $(\mathcal{P}(\Sigma)^*),
+\cup, \cdot,\sha, \emptyset, \{\epsilon\})$ forms a weak trioid.
 
-\begin{lem}
-$(\mathcal{P}(\Sigma)^*), \cup, \cdot, \sha, \emptyset, \{\epsilon\})$ is a weak trioid.
-\end{lem}
+The rely-guarantee elements in this model are sets containing all the
+words which can be built from some set of actions in $\Sigma$. We can
+consider a function $\rely : \mathcal{P}(\Sigma) \to
+\mathcal{P}(\Sigma)^*$ which takes a set of actions and returns all
+the words that can be built from them. This function is an upper
+adjoint for $\guar$, the function which returns all the letters in a
+language.
+\begin{align*}
+\guar(X) \le Y \iff X \le \rely(Y)
+\end{align*}
 
-Our alphabet comprises relations ($R,S,T,\dots$) over a state space
-$\Sigma$. An example of a word in
-$(\{\sigma_1,\sigma_2,\sigma_3\}^2)^*$ is shown below. This word only
-has a single valid \emph{trace}, which is
-$\sigma_1\rightarrow\sigma_1\rightarrow\sigma_2\rightarrow\sigma_3\rightarrow\sigma_2$.
-We define an operator $\tr(X)$ which returns all the traces in a language
-$X$.
+Properties (\ref{rg1}) and (\ref{rg2}) from Section \ref{sec:RG} can
+mostly be derived from basic properties of $\rely$. Namely,
+\begin{align*}
+\rely(X \cup Y) &= \rely(X) \| \rely(Y),\\
+\qquad \rely(X \cap Y) &= \rely(X) \cap \rely(Y).
+\end{align*}
+The proofs of (\ref{rg3}) and (\ref{rg4}) are somewhat more involved,
+but ultimately rather mechanical, so we do not show them here. We
+define the set $RG$ of rely-guarantee elements as the range of the
+$\rely$ function and therefore are able to show that $(\mathcal{P}(\Sigma)^*), RG, \cup,
+\cdot,\sha, \emptyset, \{\epsilon\})$ is a rely-guarantee algebra.
+
+To make our model more concrete we consider an alphabet comprising
+relations ($R,S,T,\dots$) over a state space $\Sigma$. A single
+relation represents a single atomic action of a program. Assuming
+$\Sigma = \{\sigma_1,\sigma_2,\sigma_3\}$, an example of a word in
+$(\Sigma^2)^*$ is shown below.
 
@@
 
-Following~\cite{brookes_full_1993} and~\cite{dingel_refinement_2002},
-we inductively generate the stutter/mumble language $w^\dagger$ for a
-word $w$ in a language over $\Sigma^2$ as follows. Assume $R,S,T \in
-\Sigma^2$ and $u,v,w \in (\Sigma^2)^*$. First, $Iw \in
-w^\dagger$. Secondly, if $uRv \in x^\dagger$ then $uIRv \in
-x^\dagger$ and $uRIv \in x^\dagger$ (\emph{stuttering}). Thirdly, if
-$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$
-(\emph{mumbling}). The stutter/mumble closure of a language is
-defined by lifting this operation to languages in the obvious way.
+This word only has a single valid \emph{trace}, which is
+$\sigma_1\rightarrow\sigma_1\rightarrow\sigma_2\rightarrow\sigma_3\rightarrow\sigma_2$.
+Such state transition traces are commonly called \emph{Aczel
+  traces}~\cite{Aczel}. Note that unlike much of the literature involving
+Aczel-style traces for concurrency, we omit any kind of process or
+environment labelling from our transitions. We define an operator
+$\tr(X)$ which returns all such traces in a language $X$. $\tr$ is
+both idempotent and isotone.
 
-Next, we define the following language for a word $w = R_0R_1\dots
-R_n$. The language $w^\udagger$ contains all strings $S_0S_1\dots S_n$
-such that for all $m \le n$, $S_m \subseteq R_m$. Again this operation
-is lifted to languages in the same way as before. Both these closure
-operators are combined then into a single closure operator,
-$X^\ddagger$, which satisfies the properties
-\begin{align*}
-(X^\ddagger \sha Y)^\ddagger &= (X \sha Y)^\ddagger,\\
-(X^\ddagger Y)^\ddagger = (&XY)^\ddagger = (XY^\ddagger)^\ddagger,\\
-(X \cup Y)^\ddagger &= (X^\ddagger \cup Y^\ddagger),\\
-(X^\ddagger \cap Y^\ddagger)^\ddagger &= (X^\ddagger \cap Y^\ddagger),\\
-\tr(X^\ddagger) &\subseteq X^\ddagger.
-\end{align*}
-For all binary operators $\bullet$, we can therefore define
-$\bullet_\ddagger$ as $\lambda X Y. (X^\ddagger\bullet
-Y^\ddagger)^\ddagger$ and show that
-$(\mathcal{P}((\Sigma^2)^*)^\ddagger,\cup_\ddagger,\cdot_\ddagger,\sha_\ddagger,\emptyset,{\epsilon}^\ddagger)$
-is a weak trioid.
+Note that we cannot use the shuffle operator on directly traces. For
+example, imagine we shuffled the single transition $\sigma_2 \to
+\sigma_3$ into the above word, we could end up with a word as below.
 
@@
 
+By performing this shuffle action, we no longer have a trace from
+$\sigma_1$ to $\sigma_2$, but instead two traces from $\sigma_2$ to
+$\sigma_2$ and $\sigma_3$. This subtle interplay between shuffle and
+$\tr$ mimics how Aczel-traces are typically generated from an
+operational semantics in much existing rely-guarantee literature. This
+operational approach is not desireable from an algebraic perspective
+though, as we would rather see programs as mathematical objects in
+their own right---a more denotational approach.
+
+Following~\cite{brookes_full_1993} and~\cite{dingel_refinement_2002},
+we inductively generate the stutter/mumble language $w^\dagger$ for a
+word $w$ in a language over $\Sigma^2$ as follows: Assume $R,S,T \in
+\Sigma^2$ and $u,v,w \in (\Sigma^2)^*$. First, $Iw \in
+w^\dagger$. Secondly, if $uRv \in x^\dagger$ then $uIRv \in
+x^\dagger$ and $uRIv \in x^\dagger$ (\emph{stuttering}). Thirdly, if
+$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$
+(\emph{mumbling}). The stutter/mumble closure of a language is
+thus defined by lifting this operation to languages.
+\begin{align*}
+X^\dagger = \bigcup\{x^\dagger.\; x \in X\}
+\end{align*}
+
+Next, we define the following language for a word $w = R_0R_1\dots
+R_n$. The language $w^\udagger$ contains all strings $S_0S_1\dots S_n$
+such that for all $m \le n$, $S_m \subseteq R_m$. Effectively this
+gives us all the words which are less that $w$ pointwise. Again this
+operation is lifted to languages in the same way as before. We define
+the set of \emph{$smp$-closed} languages by combining both these
+closure operators into a single closure operator, $X^\ddagger$, which
+satisfies the properties
+\begin{align}
+(X^\ddagger \sha Y)^\ddagger &= (X \sha Y)^\ddagger,\\
+(X^\ddagger Y)^\ddagger = (&XY)^\ddagger = (XY^\ddagger)^\ddagger,\\
+(X \cup Y)^\ddagger &= (X^\ddagger \cup Y^\ddagger),\\
+(X^\ddagger \cap Y^\ddagger)^\ddagger &= (X^\ddagger \cap Y^\ddagger),\\
+\tr(X^\ddagger) &\subseteq X^\ddagger. \label{trcoext}
+\end{align}
+These five properties essentially allow us to `lift' all our operators
+from above to act on $smp$-closed languages in such a way that they
+still form a weak trioid. Henceforth, we will assume all our languages are
+$smp$-closed, and all operators have been appropriately lifted.
+
+Equation (\ref{trcoext}) shows that $\tr$ is now an interior operator
+on the set of $smp$-closed languages.
+
+We now consider how tests are represented in this model, we define
+$\langle R\rangle$ as the singleton language containing just the
+relation $R$. Tests are defined as traces of the form $\langle
+R\rangle^\ddagger$ where $R \subseteq I$. We denote the set of all
+tests as $B$, and let it be ranged over by $P,Q,S$. From the
+definition of $smp$-closure $\langle I\rangle^\ddagger =
+\{\epsilon\}^\ddagger$, thus the greatest test is $1$, while the least
+test is $\langle \emptyset\rangle^\ddagger$, $0$. As such our tests
+form a boolean algebra between $1$ and $0$, much they would in
+Kozen's Kleene algebra with tests.
+
+\begin{center}
+\begin{tikzpicture}[x=1cm,auto]
+  \node (center) {};
+  \node [elem] (s12) {$\sigma_2$};
+  \node [elem, above of=s12] (s11) {$\sigma_1$};
+  \node [elem, below of=s12] (s13) {$\sigma_3$};
+  \node [right of=s13, node distance=0.75cm] (r1) {};
+  \node [below of=r1, node distance=0.5cm] (t4) {$R$};
+
+  \node [elem, right of=s12, node distance=1.5cm] (s22) {$\sigma_2$};
+  \node [elem, above of=s22] (s21) {$\sigma_1$};
+  \node [elem, below of=s22] (s23) {$\sigma_3$};
+  \node [right of=s23, node distance=0.75cm] (r2) {};
+  \node [below of=r2, node distance=0.5cm] (t4) {$R$};
+
+  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
+  \node [elem, above of=s32] (s31) {$\sigma_1$};
+  \node [elem, below of=s32] (s33) {$\sigma_3$};
+
+  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
+  \node [elem, above of=s42] (s41) {$\sigma_1$};
+  \node [elem, below of=s42] (s43) {$\sigma_3$};
+  \node [right of=s43, node distance=0.75cm] (r4) {};
+  \node [below of=r4, node distance=0.5cm] (t4) {$R$};
+
+  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
+  \node [elem, above of=s52] (s51) {$\sigma_1$};
+  \node [elem, below of=s52] (s53) {$\sigma_3$};
+  \node [right of=s53, node distance=0.75cm] (r5) {};
+  \node [below of=r5, node distance=0.5cm] (t5) {$\lnot R$};
+
+  \node [elem, right of=s52, node distance=1.5cm] (s62) {$\sigma_2$};
+  \node [elem, above of=s62] (s61) {$\sigma_1$};
+  \node [elem, below of=s62] (s63) {$\sigma_3$};
+
+  \path [line] (s11) -- (s21);
+  \path [line] (s12) -- (s22);
+  \path [line] (s21) -- (s31);
+  \path [line] (s22) -- (s32);
+
+  \path [rel] (s42) -- (s52);
+  \path [rel] (s41) -- (s51);
+  \path [rel] (s53) -- (s63);
+\end{tikzpicture}
+\end{center}
+
+A key difference however is that tests only behave as tests when under
+$\tr$. This is illustrated above---$\tr(\langle R\rangle^\ddagger\langle
+R\rangle^\ddagger) = \tr(\langle R\rangle^\ddagger)$ while $\tr(\langle
+R\rangle^\ddagger\langle \lnot R\rangle^\ddagger) = 0$.
+
 \newpage

### 2013-10-08 Paper/paper.tex
Figured out rules for RG algebra w/ operator

@@
 
-Axiom (\ref{rg1}) states that interference from the a rely condition
-being run twice in parallel is no different from just the
-interference from that rely condition. Axiom (\ref{rg2}) states that
-interference from a single rely condition is less than interference
-from itself and another rely condition. Axiom (\ref{rg3}) allows a
-rely or guarantee to be split across sequential programs. Axiom
-(\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent, except it deals
-with finite iteration.
+Axiom (\ref{rg1}) states that interference from a rely condition being
+run twice in parallel is no different from just the interference from
+that rely condition begin run once in parallel. Axiom (\ref{rg2})
+states that interference from a single rely condition is less than
+interference from itself and another rely condition. Axiom (\ref{rg3})
+allows a rely or guarantee to be split across sequential
+programs. Axiom (\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent,
+except it deals with finite iteration.
 
@@
 \begin{align}
-r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g.
+r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g. \label{quin}
 \end{align}
-With this encoding we can derive the standard rely-guarantee inference rules, as
-shown in Figure \ref{fig:rgrules}.
+With this encoding we can derive the standard rely-guarantee inference
+rules, as shown in Figure \ref{fig:rgrules}. Thus equations
+(\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
+rules, somehow represent a minimal set of axioms from which these
+inference rules can be derived.
 
@@
 
-By using residuals quintuples can be encoded in the following
-equivalent way,
+By using residuals quintuples can be encoded in the following way,
+which is equivalent to the encoding in Equation (\ref{quin}).
 \begin{align}
@@
 \end{align}
-This encoding allows us to think in terms of program
-refinement, as $r/(p \rightarrow q) \sqcap g$ defines the weakest
-program that when placed in parallel with interference from $r$, and
-guaranteeing interference at most $g$, goes from $p$ to $q$.
+This encoding allows us to think in terms of program refinement, as
+$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
+placed in parallel with interference from $r$, and guaranteeing
+interference at most $g$, goes from $p$ to $q$---which is essentially
+a generic specification for a concurrent program.
 
@@
 
-While this setting is adequate for deriving the standard inference
-rules, the equality is too strong to capture many interesting
-statements about concurrent programs. Consider the statement,
-$x = y \implies x\|z = y\|z$,
-which holds in our algebra and can be read as follows; if $x$ and $y$ are
-equal, then they must be equal under all possible interference from
-any arbitrary $z$. This would seem to preclude any reasoning about
-interference in a fine-grained way, but we can work around this by
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Breaking Compositionality}
+\label{sec:INT}
+
+While the algebra in section \ref{sec:RG} is adequate for deriving the
+standard inference rules, the equality is too strong to capture many
+interesting statements about concurrent programs. Consider the
+congruence rule for parallel composition, which is inherent in the
+algebraic approach:
+\begin{align*}
+x = y \implies x\|z = y\|z.
+\end{align*}
+This can be read as follows; if $x$ and $y$ are equal, then they must
+be equal under all possible interference from any arbitrary $z$. At
+first glance this might seem to preclude any reasoning about
+interference in a fine-grained way using only algebra. This is not the
+case, but breaking such inherent compositionality in just the right way to
+capture interesting properties of interference requires some extra work.
+
+The most obvious way we can acheive this is by simply expanding our
+rely-guarantee algebra with an arbitrary function $f$. And redefining our quintuples
+as,
+\begin{align*}
+r, g \vdash \{p\} c \{q\} \iff f(p(r\|c)) \le f(q) \land c \le g.
+\end{align*}
+
+ but we can work around this by
 using an \emph{interior operator}. In a rely-guarantee algebra an

### 2013-10-14 Paper/paper.tex

@@
 \usepackage{centernot}
+\usepackage{marginnote}
 
@@
 \DeclareMathOperator{\guar}{guar}
+\DeclareMathOperator{\wlp}{wlp}
 
@@
 \section{Introduction}
-
-\cite{hoare_concurrent_2011}
-\cite{kozen_completeness_1994}
-\cite{kozen_kleene_1997}
-\cite{armstrong_kleene_2013}
-\cite{nipkow_isabelle/hol:_2002}
 
@@
 = 0$ does not hold. The natural partial order $\le$ is defined as $x
-\le y $ iff $x + y = y$. Idempotent semirings are also called
-\emph{dioids}. In the context of programs, one typically thinks of
-$\cdot$ as being sequential composition, $+$ as non-deterministic
-choice, $0$ as the abortive action and $1$ as skip. For concurrent
-software, it is natural to extend the notion of a dioid to that of a
-\emph{trioid} with operators for both sequential and parallel
-composition. Formally, a trioid is a structure $(D,+,\cdot,\|,0,1)$
-where $(D,+,\cdot,0,1)$ is a dioid and,
+\le y $ iff $x + y = y$. In an idempotent semiring $(D,+,0)$ therefore
+forms a join semilattice with least element $0$. Idempotent semirings
+are also called \emph{dioids}. In some contexts, it is also useful to
+add a meet operation $\sqcap$, such that $(D,+,\sqcap)$ is a lattice.
+
+In the context of programs, one typically thinks of $\cdot$ as being
+sequential composition, $+$ as non-deterministic choice, $0$ as the
+abortive action and $1$ as skip. For concurrent software, it is
+natural to extend the notion of a dioid to that of a \emph{trioid}
+with operators for both sequential and parallel composition. Formally,
+a trioid is a structure $(D,+,\cdot,\|,0,1)$ where $(D,+,\cdot,0,1)$
+is a dioid and,
 \begin{align*}
@@
 sequential and parallel composition. One law which has been considered
-in previous papers is the \emph{exchange} law
+in previous papers is the \emph{exchange} law~\cite{hoare_concurrent_2011,hoare_locality_2011}
 \begin{align*}
@@
 $x\parallel_f y$ in a model with infinite, or non-terminating
-programs. In this model, $x \cdot y \not\leq x \parallel_f y$, in the
+programs. In this model, $x \cdot y \not\leq x \parallel_f y$ in the
 case where $x$ is non-terminating. A second example would be
@@
 
-A Kleene algebra expands a dioid with a star operation satisfying the laws
+A Kleene algebra~\cite{kozen_completeness_1994} expands a dioid with a star operation satisfying the laws
 \begin{align*}
@@
 
-For programs, the star represents finite iteration. One can similarly
-axiomatise a parallel star operation, but in practice this seems to be
-a less useful operation, although one could see it as an operator that
-spawns an arbitrary but finite number of processes.
+For programs, the star represents finite iteration of a program $x$
+zero or more times. For iteration where $x$ must be executed at least
+once, we write $x^+$. One can similarly axiomatise a parallel star
+operation, but in practice this seems to be a less useful operation,
+although one could see it as an operator that spawns an arbitrary but
+finite number of processes.
 
-\newpage
+A Kleene algebra expanded with a parallel composition operator such
+that $(K,+,\cdot,\|,0,1)$ is a trioid in which the exchange law holds
+is called a \emph{concurrent Kleene algebra}~\cite{hoare_concurrent_2011}.
+
+In a (concurrent) Kleene algebra Hoare triples can be encoded \`a la
+Tarlecki as $\{x\}y\{z\} \iff xy \le z$. Assume residuals axiomatised
+by Galois connections
+\begin{align*}
+ x \le z \leftarrow y \iff xy \le z \iff y \le x \rightarrow z,
+\end{align*}
+exist, as in Pratt's \emph{action
+  algebra}~\cite{pratt_action_1990}. Then the Hoare triple
+$\{x\}y\{z\}$ is equivalent to $x \le z \leftarrow y$, where $z
+\leftarrow y$ can be seen as $\wlp(y,z)$.
+
+For sequential programs, Hoare logic can also be encoded algebraically by
+expanding a Kleene algebra with a Boolean test subalgebra. A
+\emph{Kleene algebra with tests} (KAT)~\cite{kozen_kleene_1997} is structure
+$(K,B,+,\cdot,\phantom{|}^\star,\overline{\phantom{x}},0,1)$ where
+$(K,+,\cdot,\phantom{|}^\star,0,1)$ is a Kleene algebra and
+$(B,+,\cdot,\overline{\phantom{x}},0,1)$ is a Boolean algebra such
+that $B \subseteq K$. The $\overline{\phantom{x}}$ operator is only
+defined on elements of the Boolean subalgebra. We denote arbitrary
+elements of $K$ by $x,y,z$ and tests by $p,q,r$.
+
+In KAT Hoare triples are defined in the following (equivalent) ways
+\begin{align*}
+\{p\}x\{q\} \iff px \le xq \iff px = pxq \iff px\overline{q} = 0.
+\end{align*}
+Using these encodings all the rules of Hoare logic excluding the
+assignment rule can be derived.
+
+One might be tempted to add a test subalgbra to concurrent Kleene
+algebra in the style of KAT. In such an algebra, can the Kozen-style
+encoding of hoare triples above can be used to derive Owicki-Gries
+style concurrency rules? The answer seems to be no---an algebraic
+approach imposes strong compositionality requirements (a point we will
+return to later), so it should come as no suprise that a
+non-compositional method such as Owicki-Gries cannot be so easily be
+done algebraically.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 mutandis.
-
-\newpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Algebra for Owicki-Gries}
-\label{sec:OG}
-
-This algebra is really unsound!
-
-$px = xp$ is provable!
-
-Following in the style of Kozen's \emph{Kleene algebra with tests}
-(KAT), we extend a concurrent Kleene algebra in the same way, yielding
-\emph{concurrent Kleene algebra with tests} (CKAT). In this setting we
-can encode hoare triples not as $\{x\}y\{z\} \iff xy \le z$, as in
-Section \ref{sec:RG}, but instead as
-\begin{align*}
-\{p\}x\{q\} \iff px \le xq,
-\end{align*}
-where $p$ and $q$ are KAT-style tests, and not just arbitrary
-processes/programs.
-
-Formally, a concurrent Kleene algebra with tests is a structure
-\begin{align*}
-(K,B,+,\cdot,\|,\overline{\phantom{x}},\phantom{|}^\star,0,1),
-\end{align*}
-where $(K,+,\cdot,\|,\phantom{|}^\star,0,1)$ is a concurrent Kleene
-algebra and $(B,+,\cdot,\overline{\phantom{x}})$ is a Boolean subalgebra such that
-\begin{align*}
-(p\|q)(x\|y) &= px\|qy,\\ % THESE RULES MAKE NO SENSE!
-(x\|y)(p\|q) &= xp\|yq,\\
-p\|q &= pq,\\
-pp &= p,\\
-p &\le 1.
-\end{align*}
-Just as in KAT, negation is only defined for elements of the boolean
-subalgebra.
 

### 2013-10-29 Paper/paper.tex
Started re-working paper

@@
 \DeclareMathOperator{\guar}{guar}
+\DeclareMathOperator{\letters}{letters}
 \DeclareMathOperator{\wlp}{wlp}
@@
 
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Define a rely-guarantee algebra as a structure
+A rely-guarantee algebra is a structure
 $(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
-lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
+distributive lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
 $(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
@@
 it is impossible to construct a finite model demonstrating that
-(\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg2}).
+(\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg3}).
 
@@
 
-We can expand our rely-guarantee algebras to have residuals
-axiomatised by Galois connections for both parallel and sequential
-composition, i.e.
+If we expand our rely-guarantee algebras to have residuals axiomatised
+by Galois connections for both parallel and sequential composition,
+i.e.
 \begin{align}
@@
 xy \le z &\iff x \le z \leftarrow y, \label{res2}\\
-x\|y \le z &\iff y \le x/z \label{res3}.
+x\|y \le z &\iff y \le x/z \label{res3},
 \end{align}
-For sequential composition, (\ref{res1}) and (\ref{res2}) essentially
-provide a link between weakest-precondition style semantics and hoare
+then for sequential composition, (\ref{res1}) and (\ref{res2}) provide
+a link between weakest-precondition style semantics and hoare
 logic. As mentioned in Section \ref{sec:KA}, the triple $\{p\} c
@@
 
-By using residuals quintuples can be encoded in the following way,
+By using these residuals quintuples can be encoded in the following way,
 which is equivalent to the encoding in Equation (\ref{quin}).
@@
 placed in parallel with interference from $r$, and guaranteeing
-interference at most $g$, goes from $p$ to $q$---which is essentially
-a generic specification for a concurrent program.
+interference at most $g$, goes from $p$ to $q$---a generic
+specification for a concurrent program.
 
@@
 While the algebra in section \ref{sec:RG} is adequate for deriving the
-standard inference rules, the equality is too strong to capture many
+standard inference rules, it's equality is too strong to capture many
 interesting statements about concurrent programs. Consider the
@@
 be equal under all possible interference from any arbitrary $z$. At
-first glance this might seem to preclude any reasoning about
-interference in a fine-grained way using only algebra. This is not the
-case, but breaking such inherent compositionality in just the right way to
-capture interesting properties of interference requires some extra work.
+first sight, this might seem to preclude any reasoning about
+interference in a fine-grained way using purely algebra. This is not
+the case, but breaking such inherent compositionality in just the
+right way to capture interesting properties of concurrency requires
+extra work.
 
-The most obvious way we can acheive this is by simply expanding our
-rely-guarantee algebra with an arbitrary function $f$. And redefining our quintuples
-as,
+The most obvious way we can acheive this is by expanding our
+rely-guarantee algebra with an additional function $\pi$ and redefining
+our quintuples as,
 \begin{align*}
-r, g \vdash \{p\} c \{q\} \iff f(p(r\|c)) \le f(q) \land c \le g.
+r, g \vdash \{p\} c \{q\} \iff \pi(p(r\|c)) \le \pi(q) \land c \le g.
 \end{align*}
+Since for any operator $\bullet$, it is not required that
+\begin{align*}
+\pi(x) = \pi(y) \implies \pi(x \bullet z) = \pi(y \bullet z),
+\end{align*}
+we can `break' compositionality in just the right way, provided we
+choose the correct properties for $\pi$. For example, if we consider
+the following seven axioms
+\begin{align}
+\pi(\pi(x)) &= \pi(x), \label{pi1}\\
+x \le y \implies \pi(x) &\le \pi(y), \label{pi2}\\
+\pi(x) &\le x, \label{pi3}\\
+\pi(x^*) &\le \pi(\pi(x)^*), \label{pi4}\\
+\pi(xy) &\le \pi(\pi(x)\pi(y)), \label{pi5}\\
+\pi(x \sqcap y) &= \pi(x) \sqcap \pi(y). \label{pi6}
+\end{align}
 
- but we can work around this by
-using an \emph{interior operator}. In a rely-guarantee algebra an
-interior operator is any function $f : K \to K$ which satisfies
-\begin{align*}
-f(f(x)) &= f(x) \tag{$f$ is idempotent}\\
-x \le y \implies f(x) &\le f(y) \tag{$f$ is isotone}\\
-f(x) &\le x \tag{$f$ is coextensive}
-\end{align*}
+Axioms (\ref{pi1}) to (\ref{pi3}) show that $\pi$ is an interior
+operator, while (\ref{pi4}) and (\ref{pi5}) allow us to introduce
+$\pi$ as needed around sequential composition and finite iteration,
+which allows us to substitute under those operators as required---we
+only want to break compositionality for $\|$.
 
-We extend our rely-guarantee algebra to
-$(K,RG,+,\cdot,\|,^\star,[\--],0,1)$ by adding a single interior
-operator $[\--]$, which additionally satisfies
+The above axioms indicate that we could define $\pi$ as a function
+$\lambda x.\; x \sqcap c$ where $c$ satisfies
 \begin{align}
-[xy] &= [x][y], \label{rgimult}\\
-[x + y] &= [x] + [y],\\
-[x \sqcap y] &= [x] \sqcap [y],\\
-[x^\star] &= [x]^\star,\\
-[0] &= 0 \label{rgi0}
+x^* \sqcap c &\le (x \sqcap c)^* \sqcap c, \label{con1}\\
+xy \sqcap c &\le (x \sqcap c)(y \sqcap c) \sqcap c. \label{con2}
 \end{align}
-Intuitively $[x]$ can be thought of as the \emph{execution} of a
-program $x$. While $x$ contains all information about what the process
-or program \emph{could} do, $[x]$ only contains information about what
-it \emph{actually} does. Note that (\ref{rgimult}) -- (\ref{rgi0}) do
-not mention parallel composition, such that in general $[x] = [y]
-\centernot\implies [x\|z] = [y\|z]$. We amend our jones-quintuples to
-involve this operator as shown below.
-\begin{align*}
-r,g \vdash \{p\} c \{q\} \iff p[r\|c] \le q \land c \le g.
-\end{align*}
+
+TODO: Define algebra w/ c.
+
 All the rules in Figure \ref{fig:rgrules} can be derived with this
-encoding, and moreover their proofs remain the same, mutatis
+algebra, and moreover their proofs remain the same, mutatis
 mutandis.
 
-\newpage
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Algebra for Separation Logic}
-
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-As usual, the product of two languages X and Y over an alphabet
-$\Sigma$ is defined as $XY = \{x\frown y.\; x \in X \land y \in Y\}$.
+We consider languages where the alphabet contains state pairs of the
+form $(\sigma_1,\sigma_2) \in \Sigma \times \Sigma$. A word in such a
+language is \emph{consistent} if every such pair in a word has the
+same first state as the previous transition's second state. For
+example, $(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is consistent, while
+$(\sigma_1,\sigma_2)(\sigma_3,\sigma_3)$ is inconsistent. Sets of
+consistent words are essentially \emph{Aczel
+  traces}~\cite{boer_formal_1999}, but lack the usual process
+labels. We denote the set of all consistent words by $C$.
+
+As usual, the product of two languages X and Y is defined as
+\begin{align*}
+XY = \{x\frown y.\; x \in X \land y \in Y\}.
+\end{align*}
 The language product is associative, and distributes over arbitrary
@@
 \begin{align*}
-X \sha Y = \bigcup\{xs \sha ys.\; x \in X \land x \in Y\}.
+X \sha Y = \bigcup\{x \sha y.\; x \in X \land x \in Y\}.
 \end{align*}
@@
 treatment of the shuffle product see~\cite{shufflethings}. The
-above properties are enough to show that $(\mathcal{P}(\Sigma)^*),
+above properties are enough to show that $(\mathcal{P}((\Sigma\times\Sigma)^*),
 \cup, \cdot,\sha, \emptyset, \{\epsilon\})$ forms a weak trioid.
@@
 The rely-guarantee elements in this model are sets containing all the
-words which can be built from some set of actions in $\Sigma$. We can
-consider a function $\rely : \mathcal{P}(\Sigma) \to
-\mathcal{P}(\Sigma)^*$ which takes a set of actions and returns all
-the words that can be built from them. This function is an upper
-adjoint for $\guar$, the function which returns all the letters in a
-language.
-\begin{align*}
-\guar(X) \le Y \iff X \le \rely(Y)
-\end{align*}
+words which can be built from some set of state pairs in
+$\Sigma\times\Sigma$. We define a function $\langle R\rangle$ which
+lifts a relation $R$ to a language containing length 1 words for each
+pair in $R$. The set of rely-guaratee elements RG is then defined as
+$\{r.\; \exists R. r = \langle R\rangle^*\}$. We can show that
+$(\mathcal{P}((\Sigma\times\Sigma)^*), RG, \cup, \cdot,\sha,
+\emptyset, \{\epsilon\}, C)$ is a rely-guarantee algebra.
 
-Properties (\ref{rg1}) and (\ref{rg2}) from Section \ref{sec:RG} can
-mostly be derived from basic properties of $\rely$. Namely,
-\begin{align*}
-\rely(X \cup Y) &= \rely(X) \| \rely(Y),\\
-\qquad \rely(X \cap Y) &= \rely(X) \cap \rely(Y).
-\end{align*}
-The proofs of (\ref{rg3}) and (\ref{rg4}) are somewhat more involved,
-but ultimately rather mechanical, so we do not show them here. We
-define the set $RG$ of rely-guarantee elements as the range of the
-$\rely$ function and therefore are able to show that $(\mathcal{P}(\Sigma)^*), RG, \cup,
-\cdot,\sha, \emptyset, \{\epsilon\})$ is a rely-guarantee algebra.
-
-To make our model more concrete we consider an alphabet comprising
-relations ($R,S,T,\dots$) over a state space $\Sigma$. A single
-relation represents a single atomic action of a program. Assuming
-$\Sigma = \{\sigma_1,\sigma_2,\sigma_3\}$, an example of a word in
-$(\Sigma^2)^*$ is shown below.
+To demonstrate how this model works, consider the graphical
+representation of a language shown below.
 
@@
   \node [right of=s13, node distance=0.75cm] (r1) {};
-  \node [below of=r1, node distance=0.5cm] (t4) {R};
 
@@
   \node [right of=s23, node distance=0.75cm] (r2) {};
-  \node [below of=r2, node distance=0.5cm] (t4) {S};
 
@@
   \node [right of=s33, node distance=0.75cm] (r3) {};
-  \node [below of=r3, node distance=0.5cm] (t4) {T};
 
@@
   \node [right of=s43, node distance=0.75cm] (r4) {};
-  \node [below of=r4, node distance=0.5cm] (t4) {U};
-
-  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
-  \node [elem, above of=s52] (s51) {$\sigma_1$};
-  \node [elem, below of=s52] (s53) {$\sigma_3$};
 
@@
   \path [line] (s32) -- (s43);
-  \path [line] (s43) -- (s52);
 
@@
   \path [rel] (s23) -- (s32);
-  \path [rel] (s32) -- (s42);
-  \path [rel] (s42) -- (s51);
-  \path [rel] (s41) -- (s52);
 \end{tikzpicture}
 \end{center}
-
-This word only has a single valid \emph{trace}, which is
-$\sigma_1\rightarrow\sigma_1\rightarrow\sigma_2\rightarrow\sigma_3\rightarrow\sigma_2$.
-Such state transition traces are commonly called \emph{Aczel
-  traces}~\cite{Aczel}. Note that unlike much of the literature involving
-Aczel-style traces for concurrency, we omit any kind of process or
-environment labelling from our transitions. We define an operator
-$\tr(X)$ which returns all such traces in a language $X$. $\tr$ is
-both idempotent and isotone.
-
-Note that we cannot use the shuffle operator on directly traces. For
-example, imagine we shuffled the single transition $\sigma_2 \to
-\sigma_3$ into the above word, we could end up with a word as below.
+The language contains the following six words
+\begin{align*}
+(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
+(\sigma_1,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3),\\
+(\sigma_2,\sigma_2)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3), \qquad
+(\sigma_1,\sigma_1)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),\\
+(\sigma_1,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3), \qquad
+(\sigma_2,\sigma_2)(\sigma_3,\sigma_2)(\sigma_2,\sigma_3),
+\end{align*}
+but only the first,
+$(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is
+consistent. This word is highlighted with solid arrows in the
+diagram above. Now imagine we place the single state pair $(\sigma_2,
+\sigma_3)$ into the above language, we might end up with a language as
+below:
 
@@
   \node [right of=i3, node distance=0.75cm] (ri) {};
-  \node [below of=ri, node distance=0.5cm] (ti) {V};
 
@@
   \path [line] (s32) -- (s43);
-  \path [line] (s43) -- (s52);
   \path [line] (i2) -- (s23);
 
-  \path [rel] (s11) -- (i2);
+  \path [line] (s11) -- (i2);
   \path [line] (s12) -- (i2);
   \path [line] (s23) -- (s32);
-  \path [line] (s32) -- (s42);
-  \path [line] (s42) -- (s51);
-  \path [rel] (s41) -- (s52);
 \end{tikzpicture}
@@
 
-By performing this shuffle action, we no longer have a trace from
-$\sigma_1$ to $\sigma_2$, but instead two traces from $\sigma_2$ to
-$\sigma_2$ and $\sigma_3$. This subtle interplay between shuffle and
-$\tr$ mimics how Aczel-traces are typically generated from an
-operational semantics in much existing rely-guarantee literature. This
-operational approach is not desireable from an algebraic perspective
-though, as we would rather see programs as mathematical objects in
-their own right---a more denotational approach.
+By performing this shuffle action, we no longer have a consistent word
+from $\sigma_1$ to $\sigma_3$, but instead a consistent word from
+$\sigma_2$ to $\sigma_3$ and $\sigma_1$ to $\sigma_3$. The new
+consistent word was constructed by making a previously inconsistent
+word consistent by `filling in the gaps' in that word. This shows that
+the shuffle operator can take two inconsistent words and shuffle them
+together in such a way that it generates many consistent words. If we
+only considered consistent words, we would be unable to define such a
+shuffle operator, and would instead have to rely on some operational
+semantics to generate traces instead, which would be less
+algebraic. This also clearly demonstrates why shuffle is not well
+defined on the subalgebra we get by intersecting with $C$.
+
+\newpage
 
@@
 
-Next, we define the following language for a word $w = R_0R_1\dots
-R_n$. The language $w^\udagger$ contains all strings $S_0S_1\dots S_n$
-such that for all $m \le n$, $S_m \subseteq R_m$. Effectively this
-gives us all the words which are less that $w$ pointwise. Again this
-operation is lifted to languages in the same way as before. We define
-the set of \emph{$smp$-closed} languages by combining both these
-closure operators into a single closure operator, $X^\ddagger$, which
-satisfies the properties
-\begin{align}
-(X^\ddagger \sha Y)^\ddagger &= (X \sha Y)^\ddagger,\\
-(X^\ddagger Y)^\ddagger = (&XY)^\ddagger = (XY^\ddagger)^\ddagger,\\
-(X \cup Y)^\ddagger &= (X^\ddagger \cup Y^\ddagger),\\
-(X^\ddagger \cap Y^\ddagger)^\ddagger &= (X^\ddagger \cap Y^\ddagger),\\
-\tr(X^\ddagger) &\subseteq X^\ddagger. \label{trcoext}
-\end{align}
-These five properties essentially allow us to `lift' all our operators
-from above to act on $smp$-closed languages in such a way that they
-still form a weak trioid. Henceforth, we will assume all our languages are
-$smp$-closed, and all operators have been appropriately lifted.
-
-Equation (\ref{trcoext}) shows that $\tr$ is now an interior operator
-on the set of $smp$-closed languages.
-
-We now consider how tests are represented in this model, we define
-$\langle R\rangle$ as the singleton language containing just the
-relation $R$. Tests are defined as traces of the form $\langle
-R\rangle^\ddagger$ where $R \subseteq I$. We denote the set of all
-tests as $B$, and let it be ranged over by $P,Q,S$. From the
-definition of $smp$-closure $\langle I\rangle^\ddagger =
-\{\epsilon\}^\ddagger$, thus the greatest test is $1$, while the least
-test is $\langle \emptyset\rangle^\ddagger$, $0$. As such our tests
-form a boolean algebra between $1$ and $0$, much they would in
-Kozen's Kleene algebra with tests.
-
-\begin{center}
-\begin{tikzpicture}[x=1cm,auto]
-  \node (center) {};
-  \node [elem] (s12) {$\sigma_2$};
-  \node [elem, above of=s12] (s11) {$\sigma_1$};
-  \node [elem, below of=s12] (s13) {$\sigma_3$};
-  \node [right of=s13, node distance=0.75cm] (r1) {};
-  \node [below of=r1, node distance=0.5cm] (t4) {$R$};
-
-  \node [elem, right of=s12, node distance=1.5cm] (s22) {$\sigma_2$};
-  \node [elem, above of=s22] (s21) {$\sigma_1$};
-  \node [elem, below of=s22] (s23) {$\sigma_3$};
-  \node [right of=s23, node distance=0.75cm] (r2) {};
-  \node [below of=r2, node distance=0.5cm] (t4) {$R$};
-
-  \node [elem, right of=s22, node distance=1.5cm] (s32) {$\sigma_2$};
-  \node [elem, above of=s32] (s31) {$\sigma_1$};
-  \node [elem, below of=s32] (s33) {$\sigma_3$};
-
-  \node [elem, right of=s32, node distance=1.5cm] (s42) {$\sigma_2$};
-  \node [elem, above of=s42] (s41) {$\sigma_1$};
-  \node [elem, below of=s42] (s43) {$\sigma_3$};
-  \node [right of=s43, node distance=0.75cm] (r4) {};
-  \node [below of=r4, node distance=0.5cm] (t4) {$R$};
-
-  \node [elem, right of=s42, node distance=1.5cm] (s52) {$\sigma_2$};
-  \node [elem, above of=s52] (s51) {$\sigma_1$};
-  \node [elem, below of=s52] (s53) {$\sigma_3$};
-  \node [right of=s53, node distance=0.75cm] (r5) {};
-  \node [below of=r5, node distance=0.5cm] (t5) {$\lnot R$};
-
-  \node [elem, right of=s52, node distance=1.5cm] (s62) {$\sigma_2$};
-  \node [elem, above of=s62] (s61) {$\sigma_1$};
-  \node [elem, below of=s62] (s63) {$\sigma_3$};
-
-  \path [line] (s11) -- (s21);
-  \path [line] (s12) -- (s22);
-  \path [line] (s21) -- (s31);
-  \path [line] (s22) -- (s32);
-
-  \path [rel] (s42) -- (s52);
-  \path [rel] (s41) -- (s51);
-  \path [rel] (s53) -- (s63);
-\end{tikzpicture}
-\end{center}
-
-A key difference however is that tests only behave as tests when under
-$\tr$. This is illustrated above---$\tr(\langle R\rangle^\ddagger\langle
-R\rangle^\ddagger) = \tr(\langle R\rangle^\ddagger)$ while $\tr(\langle
-R\rangle^\ddagger\langle \lnot R\rangle^\ddagger) = 0$.
-
 \newpage

### 2013-11-05 Paper/paper.tex
More work on assignment
Examples...

@@
 \DeclareMathOperator{\wlp}{wlp}
+\DeclareMathOperator{\test}{test}
+\DeclareMathOperator{\eval}{eval}
 
@@
   We investigate equational laws for the derivation of rely-guarantee
-style, Owicki-Gries style and separation logic style inference and
-refinement rules in the minimalistic setting of semirings. The
-algebraic level is linked with concrete programming models based on
-languages and execution traces. The approach has been implemented in
-Isabelle/HOL where the link between algebras and models supports
-reasoning about both the control and data flow of concurrent
-programs. This is illustrated by simple program verification examples.
+  style inference and refinement rules in the minimalistic setting of
+  semirings. This algebraic level is linked with concrete programming
+  models based on languages and execution traces. The approach has
+  been implemented in Isabelle/HOL where the link between algebras and
+  models supports reasoning about both the control and data flow of
+  concurrent programs. This is illustrated by simple program
+  verification examples.
 \end{abstract}
@@
 
+In the trace model, assignment is defined as
+\begin{align*}
+  x := e = \bigsqcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
+\end{align*}
+where $x \leftarrow v$ denotes the atomic command which assigns the value $v$ to $x$.
+
 \newpage

### 2013-11-07 Paper/paper.tex
Started work on FINDP program

@@
   models supports reasoning about both the control and data flow of
-  concurrent programs. This is illustrated by simple program
-  verification examples.
+  concurrent programs at varying levels of abstraction. Finally, we
+  illustrate our approach via simple program verification examples.
 \end{abstract}

### 2013-11-11 Paper/paper.tex
Updates to paper

@@
 \DeclareMathOperator{\eval}{eval}
+\DeclareMathOperator{\Id}{Id}
+\DeclareMathOperator{\edn}{end}
 
@@
 $\cdot$ have different units, which, in the presence of the exchange
-law implies that $\cdot$ is equal to $\|$.
+law, implies that $\cdot$ is equal to $\|$.
 
@@
 once, we write $x^+$. One can similarly axiomatise a parallel star
-operation, but in practice this seems to be a less useful operation,
-although one could see it as an operator that spawns an arbitrary but
-finite number of processes.
+operation.
 
@@
 assignment rule can be derived.
-
-One might be tempted to add a test subalgbra to concurrent Kleene
-algebra in the style of KAT. In such an algebra, can the Kozen-style
-encoding of hoare triples above can be used to derive Owicki-Gries
-style concurrency rules? The answer seems to be no---an algebraic
-approach imposes strong compositionality requirements (a point we will
-return to later), so it should come as no suprise that a
-non-compositional method such as Owicki-Gries cannot be so easily be
-done algebraically.
 
@@
 
-TODO: Define algebra w/ c.
+As such we redefine our rely-guarantee algebra from section
+\ref{sec:RG} as a structure $(K,RG,+,\sqcap,\cdot,\|,^\star,0,1,c)$
+which, in addition to the rules in section \ref{sec:RG} satisfies both
+(\ref{con1}) and (\ref{con2}).
 
@@
 $\Sigma\times\Sigma$. We define a function $\langle R\rangle$ which
-lifts a relation $R$ to a language containing length 1 words for each
-pair in $R$. The set of rely-guaratee elements RG is then defined as
-$\{r.\; \exists R. r = \langle R\rangle^*\}$. We can show that
+lifts a relation $R$ to a language containing words of length one for
+each pair in $R$. The set of rely-guaratee elements RG is then defined
+as $\{r.\; \exists R. r = \langle R\rangle^*\}$. We can show that
 $(\mathcal{P}((\Sigma\times\Sigma)^*), RG, \cup, \cdot,\sha,
 \emptyset, \{\epsilon\}, C)$ is a rely-guarantee algebra.
+
+Since $\langle R\rangle$ is atomic, it satisfies several useful properties, such as,
+\begin{align*}
+\langle R\rangle^\star \| \langle S\rangle &= \langle R\rangle^\star; \langle S\rangle; \langle R\rangle^\star\\
+\langle R\rangle^\star \| \langle S\rangle^\star &= (\langle R\rangle^\star; \langle S\rangle^\star)^\star
+\end{align*}
 
@@
 \end{align*}
-but only the first,
+where only the first,
 $(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is
 consistent. This word is highlighted with solid arrows in the
-diagram above. Now imagine we place the single state pair $(\sigma_2,
+diagram above. Now if we shuffle the single state pair $(\sigma_2,
 \sigma_3)$ into the above language, we might end up with a language as
@@
 from $\sigma_1$ to $\sigma_3$, but instead a consistent word from
-$\sigma_2$ to $\sigma_3$ and $\sigma_1$ to $\sigma_3$. The new
-consistent word was constructed by making a previously inconsistent
-word consistent by `filling in the gaps' in that word. This shows that
-the shuffle operator can take two inconsistent words and shuffle them
-together in such a way that it generates many consistent words. If we
-only considered consistent words, we would be unable to define such a
-shuffle operator, and would instead have to rely on some operational
-semantics to generate traces instead, which would be less
-algebraic. This also clearly demonstrates why shuffle is not well
-defined on the subalgebra we get by intersecting with $C$.
-
-\newpage
+$\sigma_2$ to $\sigma_3$ and $\sigma_1$ to $\sigma_3$. These new
+consistent words were constructed from previously inconsistent
+words---the shuffle operator can take two inconsistent words and
+shuffle them together in such a way to generate many consistent
+words. If we only considered consistent words, \'a la Aczel traces, we
+would be unable to define such a shuffle operator directly on the
+traces themselves, and would instead have to rely on some operational
+semantics to generate traces.
 
 Following~\cite{brookes_full_1993} and~\cite{dingel_refinement_2002},
-we inductively generate the stutter/mumble language $w^\dagger$ for a
-word $w$ in a language over $\Sigma^2$ as follows: Assume $R,S,T \in
-\Sigma^2$ and $u,v,w \in (\Sigma^2)^*$. First, $Iw \in
-w^\dagger$. Secondly, if $uRv \in x^\dagger$ then $uIRv \in
-x^\dagger$ and $uRIv \in x^\dagger$ (\emph{stuttering}). Thirdly, if
-$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$
-(\emph{mumbling}). The stutter/mumble closure of a language is
-thus defined by lifting this operation to languages.
+we inductively generate the mumble language $w^\dagger$ for a word $w$
+in a language over $\Sigma^2$ as follows: Assume $R,S,T \in \Sigma^2$
+and $u,v,w \in (\Sigma^2)^*$. First, $w \in w^\dagger$. Secondly, if
+$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$. This operation is
+lifted to languages in the obvious way as
 \begin{align*}
-X^\dagger = \bigcup\{x^\dagger.\; x \in X\}
+X^\dagger = \bigcup\{x^\dagger.\; x \in X\}.
 \end{align*}
+Stuttering is represented as a rely condition $\langle \Id\rangle^\star$
+where $\Id$ is the identity relation. Two languages $X$ and $Y$ are
+equal under stuttering if $\langle \Id\rangle^\star \| X =_\pi \langle
+\Id\rangle^\star \| Y$.
+
+Stuttering and mumbling are important for defining \emph{tests}. A
+test is any language $P$ where $P \le \langle \Id\rangle$. We
+write $\test(P)$ for $\langle \Id_P\rangle$. Without any
+stuttering and mumbling the traces $\test(P); \test(Q)$ and
+$\test(P \cap Q)$ are incomparable, as all words in the former have
+length two, while all the words in the latter have length one. With mumbling we have that
+\begin{align*}
+\test(P \cap Q) \le_\pi \test(P); \test(Q)
+\end{align*}
+as the longer words in $\test(P);\test(Q)$ can be mumbled down into
+the shorter words of $\test(P\cap Q)$, whereas with stuttering we
+have the opposite,
+\begin{align*}
+\langle I\rangle^\star \| (\test(P);\test(Q)) \le_\pi \langle I\rangle^\star \| \test(P \cap Q).
+\end{align*}
+For the rest of the paper, we assume that all languages are
+implicitly mumble closed.
+
+Using tests, we can encoding common program statements such as if
+statements and while loops as in Kozen's Kleene algebra with tests.
+\begin{align*}
+\text{if } P \text{ \{ } X \text { \} else \{ } Y \text{ \}} &= \test(P); X + \test(- P); Y\\
+\text{while } P \text{ \{ } X \text{ \}} &= (\test(P);X)^\star;\test(- P)
+\end{align*}
+Finally, we define the operator $\edn(P)$ which is contains all the
+words which end in a state satisfying $P$. Some useful properties of $\edn$ include
+\begin{align*}
+\edn(P); \test{Q} &\le_\pi \edn(P \cap Q),\\
+\test(P) &\le \edn(Q),\\
+\text{range}(\Id_P \circ R) \le P \implies \edn(P); \langle R\rangle^\star &\le_\pi \edn(P).
+\end{align*}
+
+In the trace model, assignment is defined as
+\begin{align*}
+  x := e = \bigcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
+\end{align*}
+where $x \leftarrow v$ denotes the atomic command which assigns the
+value $v$ to $x$. The $\eval$ function evaluates an expression $e$ in
+the state $\sigma$. Using this definition of assignment we can derive
+the assignment rule
+\begin{align*}
+&\text{unchanged}(\text{vars}(e)) \cap \text{preserves}(P) \cap \text{preserves}(P[x/e]),\\
+&\text{unchanged}(- {x})\\
+&\vdash \{\edn(P)\} x := e \{\edn(P[x/e])\}.
+\end{align*}
+
+The rely condition states that the environment is not allowed to
+modify any of the variables used when evaluating $e$, i.e. those
+variables must remain unchanged, and also that the environment must
+preserve the precondition and postcondition of the assignment
+statement. In turn, the assignment statement itself guarantees that it
+leaves every variable other than $x$ unchanged. Preserves and unchanged are defined as
+\begin{align*}
+\text{preserves}(P) &= \langle \{(\sigma,\sigma').\; P(\sigma) \implies P(\sigma')\}\rangle^\star,\\
+\text{unchanged}(X) &= \langle \{(\sigma,\sigma').\; \forall v\in X.\; \sigma(v) = \sigma'(v)\}\rangle^\star.
+\end{align*}
+We also defined two futher rely conditions, increasing and decreasing,
+which are defined much like unchanged except they only require that variables
+increase or decrease, rather than stay the same.
 
@@
 
-In the trace model, assignment is defined as
+Figure \ref{fig:findp} shows the classic FINDP program, which has been
+used by numerous authors including Owicki~\cite{owicki},
+Jones~\cite{jones}, de Roever~\cite{derover} and
+Hayes~\cite{hayes}. The program finds the least element of an array
+satisfying the predicate $P$. The index of the first element
+satisfying $p$ is placed in the variable $f$. If there are no elements
+of the array satisfying $P$ then $f$ will be set to the length of the
+array. The program has two subprograms, $A$ and $B$, which run in
+parallel, one of which searches the even indices while the other
+searches the odd indices. A speedup over a sequential implementation
+is acheived as $A$ will terminate when $B$ finds an element of the
+array satisfying $P$ which is less than $i_A$.
+
+\begin{figure}
+\[
+\begin{aligned}
+&f_A := \text{len}(\text{array});\\
+&f_B := \text{len}(\text{array});\\
+&\left(
+\begin{aligned}
+&i_A = 0\\
+&\text{while } i_A < f_A \land i_A < f_B \text{ \{ }\\
+&\qquad\text{if } P(\text{array}[i_A]) \text{ \{}\\
+&\qquad\qquad f_A := i_A\\
+&\qquad\text{\} else \{}\\
+&\qquad\qquad i_A := i_A + 2\\
+&\qquad\text{\}}\\
+&\text{\}}
+\end{aligned}
+\middle\|
+\begin{aligned}
+&i_B = 1\\
+&\text{while } i_B < f_A \land i_B < f_B \text{ \{}\\
+&\qquad\text{if } P(\text{array}[i_B]) \text{ \{}\\
+&\qquad\qquad f_B := i_B\\
+&\qquad\text{\} else \{}\\
+&\qquad\qquad i_B := i_B + 2\\
+&\qquad\text{\}}\\
+&\text{\}}
+\end{aligned}
+\right);\\
+&f = \text{min}(f_A,f_B)
+\end{aligned}
+\]
+\caption{FINDP Program}
+\label{fig:findp}
+\end{figure}
+
+To prove the correctness of FINDP, we must show that
 \begin{align*}
-  x := e = \bigsqcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
+\text{FINDP} \le_\pi \text{ends}\{\sigma.\; \text{leastp}(\sigma, f)\} + \text{ends}\{\sigma.\; \sigma(f) = \text{len}(\text{array})\}.
 \end{align*}
-where $x \leftarrow v$ denotes the atomic command which assigns the value $v$ to $x$.
+where
+\begin{align*}
+\text{leastp}(\sigma, x) =  (\forall v<\sigma(x). \lnot P(\sigma(\text{array}[v])))\land P(\sigma(\text{array}[\sigma(x)])).
+\end{align*}
+The interesting part of this proof is the verification
 

### 2013-11-11 Paper/paper.tex
Started on introduction

@@
 
-\title{Algebraic Principles for Concurrency Verification Tools}
+\title{Algebraic Principles for Rely-Guarantee Style Concurrency Verification Tools}
 
-\author{Alasdair Armstrong \and Victor B.~F.~Gomes\thanks{Sponsored by CNPq, Brazil}  \and Georg Struth}
+\author{Alasdair Armstrong \and Victor B.~F.~Gomes \and Georg Struth}
 
@@
 \begin{abstract}
-  We investigate equational laws for the derivation of rely-guarantee
-  style inference and refinement rules in the minimalistic setting of
-  semirings. This algebraic level is linked with concrete programming
-  models based on languages and execution traces. The approach has
-  been implemented in Isabelle/HOL where the link between algebras and
-  models supports reasoning about both the control and data flow of
-  concurrent programs at varying levels of abstraction. Finally, we
-  illustrate our approach via simple program verification examples.
+  We provide simple equational principles for deriving
+  rely-guarantee-style inference rules and refinement laws based on
+  idempotent semirings. We link the algebraic layer with concrete
+  models of programs based on languages and execution traces. We have
+  implemented the approach in Isabelle/HOL as a lightweight
+  concurrency verification tool that supports reasoning about the
+  control and data flow of concurrent programs with shared variables
+  at different levels of abstraction. This is illustrated on two
+  simple verification examples.
 \end{abstract}
@@
 
-x
+We aim to develop tools which support reasoning about concurrent
+programs at varying levels of abstraction. At the highest level,
+abstract algebra provides a framework for the prototyping of new
+inference and refinement rules, whereas at the lowest level concrete
+models of program stores allow for fine-grained reasoning about
+program data flow. We have implemented this approach is the
+Isabelle/HOL interactive theorem prover. Isabelle/HOL not only
+provides us with a high degree of confidence in the correctness of our
+theorems, but also supports the construction of custom proof tactics
+and tools for program verification tasks. Our approach differs from
+existing frameworks for rely-guarantee reasoning in
+Isabelle~\cite{nieto}, both by making the link between concrete
+programs and algebras explicit, but also by allowing arbitrary nested
+parallelism.
 
-x
+For sequential programs the applicability of algebra, and Kleene
+algebra in particular, has been known for over a decade. Kleene
+algebra provides operations for non-deterministic choice, sequential
+composition and finite iteration, in addition to skip and abort. With
+appropriate extensions, Kleene algebras support Hoare-style program
+verification, and allow the derivation of program equivalences and
+refinement rules. Furthermore, Kleene algebra has been applied in many
+applications, such as compiler
+optimisation~\cite{kozen_certification_2000}, program construction,
+transformation and termination analysis, and static analysis.
 
-x
+Algebra has also been applied in the study of separation logic, for
+example, as the separation algebras in~\cite{ohearn}, or in algebraic
+separation logic.
 
-x
+\fbox{Algebraic Separation Logic}
 
-x
+\fbox{Concurrent Kleene Algebra}
 
-x
+Rely-guarantee is a so called \emph{compositional} method, where the
+verification of a large program is reduced the independent
+verification of individual subprograms, with the interference from
+other subprograms being abstractly captured by \emph{rely} and
+\emph{guarantee} conditions. This form of compositionality is,
+however, weaker than the strong notion of compositionality inherent in
+algebra, where full congruence rules exist for every operator.
 
-x
+In this paper, we investigate algebraic principles for rely-guarantee
+style reasoning. This has been explored previously, for example
+in~\cite{cka}, but we extend this approach by considering a minimal
+set of rules sufficient to derive the standard rely-guarantee
+inference rules. Furthermore, we explore how the algebraic
+compositionality of this approach can be broken in just the right way,
+so as to capture the ideas of abstract interference conditions in
+rely-guarantee reasoning.
 
-x
+The verification and construction of actual concurrent programs often
+requires complex semantics involving synchronisation, termination
+analysis and complicated data structures.
 
-x
+\fbox{We verify simple programs}
 
-x
+\fbox{Coherent framework for programs}
 
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-x
-
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Semirings and Kleene Algebra}
+\section{Algebraic Preliminaries}
 \label{sec:KA}
 
-A \emph{semiring} is an structure
-$(D,+,\cdot,0,1)$ satisfying the following laws,
-\begin{align*}
-  x(y + z) &= xy + xz, &\qquad (x + y)z &= xz + yz,\\
-  (x + y) + z &= x + (y + z), &\qquad (xy)z &= x(yz),\\
-  x + y &= y + x, &\qquad 0 + x &= x,\\
-  1x &= x, &\qquad x1 &= x,\\
-  0x &= 0, &\qquad x0 &= 0.
-\end{align*}
-A semiring is \emph{idempotent} if $x + x = x$, and \emph{weak} if $x0
-= 0$ does not hold. The natural partial order $\le$ is defined as $x
-\le y $ iff $x + y = y$. In an idempotent semiring $(D,+,0)$ therefore
-forms a join semilattice with least element $0$. Idempotent semirings
-are also called \emph{dioids}. In some contexts, it is also useful to
-add a meet operation $\sqcap$, such that $(D,+,\sqcap)$ is a lattice.
+\fbox{define weak trioid!}
 
-In the context of programs, one typically thinks of $\cdot$ as being
-sequential composition, $+$ as non-deterministic choice, $0$ as the
-abortive action and $1$ as skip. For concurrent software, it is
-natural to extend the notion of a dioid to that of a \emph{trioid}
-with operators for both sequential and parallel composition. Formally,
-a trioid is a structure $(D,+,\cdot,\|,0,1)$ where $(D,+,\cdot,0,1)$
-is a dioid and,
-\begin{align*}
-  x\|(y + z) &= x\|y + x\|z, &\qquad x\|y &= y\|x,\\
-  (x\|y)\|z &= x\|(y\|z), &\qquad x\|1 &= x,\\
-  0\|x &= x.
-\end{align*}
+Rely-guarantee algebras, which are introduced in the following
+section, are based on dioids and Kleene algebras. A \emph{semiring} is
+a structure $(S,+,\cdot,0,1)$ such that $(S,+,0)$ is a commutative
+monoid, $(S,\cdot, 1)$ is a monoid and the distributivity laws $x\cdot
+(y+z)=x\cdot z + y \cdot z$ and $(x+y)\cdot z = x\cdot z+y\cdot z$ as
+well as the annihilation laws $x\cdot 0=0$ and $0\cdot x=0$ hold. A
+\emph{dioid} is a semiring in which addition is idempotent:
+$x+x=x$. Hence $(S,+,0)$ forms a join semilattice with least element
+$0$ and partial order defined, as usual, as $x\le y\Leftrightarrow
+x+y=y$. The operations of addition and multiplication are isotone with
+respect to the order, that is, $x \le y $ implies $z+x\le z+y$,
+$z\cdot x \le z\cdot y$ and $x\cdot z \le y\cdot z$. A dioid is
+\emph{commutative} if multiplication is: $x\cdot y = y \cdot x$.
 
-In a trioid there are no laws governing the interaction between
-sequential and parallel composition. One law which has been considered
-in previous papers is the \emph{exchange} law~\cite{hoare_concurrent_2011,hoare_locality_2011}
-\begin{align*}
-  (w \| x) \cdot (y \| z) \le (w \cdot z) \| (x \cdot y).
-\end{align*}
+In the context of sequential programs, one typically thinks of $\cdot$
+as sequential composition, $+$ as nondeterministic choice, $0$ as the
+abortive action and $1$ as skip.  In this context it is essential that
+multiplication is not commutative.  More formally, it is well known
+that (regular) languages with language union as $+$ and language
+product as $\cdot$, the empty language as $0$ and the empty word
+language $\{\varepsilon\}$ as $1$ form dioids. Another model is formed
+by binary relations with the union of relations as $+$, the product of
+relations as $\cdot$, the empty relation as $0$ and the identity
+relation as $1$. A model of commutative dioids is formed by sets of
+(finite) multisets or Parikh vectors with multiset addition as
+multiplication.
 
-The exchange law is initially appealing as it provides a way to derive
-many properties of parallel composition from a single law. For example
-one can show that $\|$ must be commutative, or that $x\cdot y \le
-x\|y$. However, there are many interesting models which do not satisfy
-this law. First, one could consider fair parallel composition
-$x\parallel_f y$ in a model with infinite, or non-terminating
-programs. In this model, $x \cdot y \not\leq x \parallel_f y$ in the
-case where $x$ is non-terminating. A second example would be
-predicate-transformer style models for separation logic, where $\|$ is
-interpreted as separating conjunction. In such models, $\|$ and
-$\cdot$ have different units, which, in the presence of the exchange
-law, implies that $\cdot$ is equal to $\|$.
+It is well known that commutative dioids can be used for modelling the
+interaction between concurrent composition and nondeterministic
+choice. The following definition serves as a basis for models of
+concurrency in which sequential and concurrent composition
+interact.\fbox{should we write $;$ instead of $\cdot$?}
 
-A Kleene algebra~\cite{kozen_completeness_1994} expands a dioid with a star operation satisfying the laws
-\begin{align*}
-  1 + xx^* &\le x^*, &\qquad z + xy \le y &\implies x^*z \le y,\\
-  1 + x^*x &\le x^*, &\qquad z + yx \le y &\implies zx^* \le y.
-\end{align*}
+A \emph{trioid} is a structure $(S,+,\cdot,||,0,1)$ such that
+$(S,+,\cdot,0,1)$ is a dioid and $(S,\cdot,||,0,1)$ a commutative
+dioid. In a trioid there is no interaction between the sequential
+composition $\cdot$ and the parallel composition $||$. On the one
+hand, Gischer has shown that trioids are sound and complete for the
+equational theory of series-parallel pomset languages~\cite{Gischer},
+which form a well studied model of true concurrency. On the other
+hand, he has also obtained a completeness result with respect to a
+notion of pomset subsumption for trioids with the additional
+\emph{interchange axiom} $(w \| x) \cdot (y \| z) \le (w \cdot y) \|
+(x \cdot z)$ and it is well known that this additional axiom also
+holds for (regular) languages in which $||$ is interpreted as the
+shuffle or interleaving operation~\cite{Gischer}.
 
-For programs, the star represents finite iteration of a program $x$
-zero or more times. For iteration where $x$ must be executed at least
-once, we write $x^+$. One can similarly axiomatise a parallel star
-operation.
+Formally, the \emph{shuffle} $\sha$ of two finite words is defined
+inductively as
+\begin{equation*}
+\epsilon \sha s = \{s\}, \qquad s \sha \epsilon = \{s\},\qquad
+as \sha bt = a(s \sha bt) \cup b(as \sha t),
+\end{equation*}
+which is then lifted to the shuffle product of languages $X$ and $Y$
+as
+\begin{equation*}
+X \sha Y = \{x \sha y:\; x \in X \land x \in Y\}.
+\end{equation*}
+\fbox{I would use $||$ here}
 
-A Kleene algebra expanded with a parallel composition operator such
-that $(K,+,\cdot,\|,0,1)$ is a trioid in which the exchange law holds
-is called a \emph{concurrent Kleene algebra}~\cite{hoare_concurrent_2011}.
+For logics of programs, notions of iteration or recursion are
+essential. A \emph{Kleene algebra} is a dioid expaned with a star
+operation which satisfies the \emph{left unfold axiom} $1 + x\cdot
+x^*\le x^*$ and the \emph{left} and \emph{right induction axiom} $z+x
+\cdot y \le y\Rightarrow x^*\cdot z \le y$ and $z + y\cdot x \le y
+z\cdot \Rightarrow x^* \le y$. It follows that $1+x\cdot x^* = x^*$
+and that the right unfold axiom $1+x^*\cdot x=x^*$ is derivable as
+well. Thus iteration $x^\ast$ is modelled as the least fixpoint of the
+function $\lambda y. 1+x\cdot y$, which is the same as the least
+fixpoint of $\lambda y.1+ y\cdot x$. A \emph{commutative Kleene
+  algebra} is a Kleene algebra in which multiplication is commutative.
 
-In a (concurrent) Kleene algebra Hoare triples can be encoded \`a la
-Tarlecki as $\{x\}y\{z\} \iff xy \le z$. Assume residuals axiomatised
-by Galois connections
-\begin{align*}
- x \le z \leftarrow y \iff xy \le z \iff y \le x \rightarrow z,
-\end{align*}
-exist, as in Pratt's \emph{action
-  algebra}~\cite{pratt_action_1990}. Then the Hoare triple
-$\{x\}y\{z\}$ is equivalent to $x \le z \leftarrow y$, where $z
-\leftarrow y$ can be seen as $\wlp(y,z)$.
+It is well known that (regular) languages form Kleene algebras and
+that (regular) sets of multisets form commutative Kleene algebras. In
+fact, Kleene algebras are complete with respect to the equational
+theory of regular languages as well as the equational theory of binary
+relations with the reflexive transitive closure operation as the
+star~\cite{Kozen}. Moreover, commutative Kleene algebras are
+complete with respect to the equational theory of regular languages
+over multisets~\cite{Conway}. It follows that equations in
+(commutative) Kleene algebras are decidable.
 
-For sequential programs, Hoare logic can also be encoded algebraically by
-expanding a Kleene algebra with a Boolean test subalgebra. A
-\emph{Kleene algebra with tests} (KAT)~\cite{kozen_kleene_1997} is structure
-$(K,B,+,\cdot,\phantom{|}^\star,\overline{\phantom{x}},0,1)$ where
-$(K,+,\cdot,\phantom{|}^\star,0,1)$ is a Kleene algebra and
-$(B,+,\cdot,\overline{\phantom{x}},0,1)$ is a Boolean algebra such
-that $B \subseteq K$. The $\overline{\phantom{x}}$ operator is only
-defined on elements of the Boolean subalgebra. We denote arbitrary
-elements of $K$ by $x,y,z$ and tests by $p,q,r$.
+A \emph{bi-Kleene algebra} is a structure
+$(K,+,\cdot,||,0,1,^*,^{(*)})$ such that $(K,+,\cdot,0,1,^*)$ is a
+Kleene algebra and $(K,+,||,0,1,^{(*)})$ a commutative Kleene
+algebra. Bi-Kleene algebras are sound and complete with respect to the
+equational theory of regular series-parallel pomset languages, and the
+equational theory is again decidable~\cite{LaurenceStruth}. A
+\emph{concurrent Kleene algebra} is a bi-Kleene algebra which
+satisfies the interchange law~\cite{Hoareetal}. It can be shown that
+shuffle languages and regular series-parallel pomset languages with a
+suitable notion of pomset subsumption form concurrent Kleene algebras.
 
-In KAT Hoare triples are defined in the following (equivalent) ways
-\begin{align*}
-\{p\}x\{q\} \iff px \le xq \iff px = pxq \iff px\overline{q} = 0.
-\end{align*}
-Using these encodings all the rules of Hoare logic excluding the
-assignment rule can be derived.
+In some contexts, it is also useful to add a meet operation $\sqcap$,
+such that $(D,+,\sqcap)$ is a lattice. \fbox{expand}
 
-\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Generalised Hoare Logics  in Kleene Algebra}
+
+It is well known that the inference rules for sequential Hoare logic
+(except the assignment axiom) can be derived in expansions of Kleene
+algebras. One approach is as follows~\cite{MoellerStruth}. Suppose a
+suitable Boolean algebra $B$ of \emph{tests} has been embedded into a
+Kleene algebra $K$ such that $0$ and $1$ are the minimal and maximal
+element of $B$, $+$ corresponds to join and $\cdot$ to
+meet. Complements $-$ are defined only on $B$. Suppose further that a
+\emph{backward diamond operator} $\langle x|p$ has been defined for
+each $x\in K$ and $p\in B$, which models the set of all states to
+which each terminating execution of program $x$ may lead from states
+$q$. Finally suppose that a \emph{forward box operator} $|x]p$ has
+been defined which models the (largest) set of states from which every
+terminating execution of $x$ must end up in states $q$ and that boxes
+and diamonds are adjoints of the Galois connection
+\begin{equation*}
+  \langle x|p \le q \Leftrightarrow p \le |x]q,
+\end{equation*}
+for all $x\in K$ and $p,q\in B$. It is then evident from the above
+explanations that validity of a Hoare triple $\vdash \{p\}x\{q\}$ can
+be encoded as $\langle x|p \le q$ and the weakest liberal precodition
+operator $\mathsf{wlp}(x,q)$ as $|x]p$. Hence the relationship between
+the proof theory and the semantics of Hoare logic is captured by the
+Galois connection $ \vdash\{p\}x\{q\} \Leftrightarrow p \le
+\mathsf{wlp}(x,q)$. It has been shown that the relational semantics of
+sequential while-programs can be encoded in these \emph{modal Kleene
+  algebras} and that the inference rules of Hoare logic can be
+derived~\cite{MoellerStruth}.
+
+In the context of concurrency, this relational approach is no longer
+appropriate. Here, the following approach by
+Tarlecki~\cite{tarlecki_language_1985} can be used instead. One can
+now encode validity of a Hoare triple as
+\begin{equation*}
+  \vdash \{x\}y\{z\} \Leftrightarrow x \cdot y \le z
+\end{equation*}
+for arbitrary elements of a Kleene algebra $K$. Nevertheless all the
+rules of sequential Hoare logic except the assignment axiom can still
+be derived in this setting~\cite{Hoareetal}.  Tarlecki's motivating
+explanations carry over to the algebraic approach.
+
+As an example we show the derivation of a generalised while
+rule. Suppose $x\cdot t\cdot y\le x$. Then $x\cdot (t\cdot y)^\ast \le
+x$ by the right induction axiom of Kleene algebra and therefore
+$x\cdot (t\cdot y)^\ast t'\le x\cdot t'$ for arbitrary element $t'$ by
+isotonicity of multiplication. This derives the while rule
+\begin{equation*}
+  \frac{\{x\cdot t\}y\{x\}}{\{x\}(t\cdot y)^\ast\{t'\cdot x\}}
+\end{equation*}
+for a generalised while loop $(t\cdot y)^\ast\cdot t'$, which
+specialises to the conventional rule when $t$ and $t'$ are, in some
+sense, complements.
+
+The correspondence to a wlp-style semantics, as in model Kleene
+algebra, now requires a generalisation of the Galois connection for
+boxes and diamonds to multiplication and an upper adjoint in the form
+of residuation.  This can be achieved in the context of \emph{action
+  algebras}~\cite{pratt}, which expand Kleene algebras by operations
+of left and right residuation defined by the Galois connections
+\begin{equation*}
+  xy\le z\Leftrightarrow x\le z\leftarrow y,\qquad xy\le z \Leftrightarrow y\le z\rightarrow x.
+\end{equation*}
+\fbox{please check that I didn't confuse these arrows} These
+residuals, and now even the Kleene star can be axiomatised
+equationally in action algebras. For a comprehensive list of the
+properties of action algebras and their most important models
+see~\cite{archive}, including the language and the relational
+model. In analogy to the development in model Kleene algebra we can
+now stipulate $ \wlp(x,y)=y\leftarrow x$ and obtain the Galois
+connection
+\begin{equation*}
+  \vdash \{x\}y\{z\} \Leftrightarrow x\le \wlp(y,z)
+\end{equation*}
+with the characteristic properties $\vdash \{\wlp(y,z)\}y\{z\}$ and $x\le
+\wlp(y,z)\Rightarrow \ \vdash \{x\}y\{z\}$. Moreover, if the action algebra is
+also a quantale, and infinite sums exist, it follows that $
+\wlp(y,z)=\sum\{x:\ \vdash \{x\}y\{z\}\}$.  It is obvious that this definition
+makes sense in all models of action algebras and
+quantales. Intuitively, suppose $p$ stands for the set of all
+behaviours of a system, for instance the set of all execution traces,
+that end in state $p$, and likewise for $q$. Then $\{p\}x\{q\}$ states
+that all executions ending in $p$ can be extended by $x$ to executions
+ending in $q$. $\wlp(x,q)$ is the most general behaviour, that is the
+set of all executions $p$ after which all executions of $x$ must end
+in $q$.
+
+In the context of concurrent Kleene algebra, even without the parallel
+star, additional inference rules can be derived, in particular 
+\begin{equation*}
+  \frac{\vdash\{x\}y\{z\}}{\vdash\{w||x\}y\{w||z\}}\qquad \frac{\vdash\{x_1\}y_1\{z_1\}\quad\vdash\{x_2\}y_2\{z_2\}}{\vdash\{x_1||x_2\}y_1||y_2\{z_1||z_2\}}
+\end{equation*}
+which correspond to the frame rule and the concurrency rule of
+concurrent separation logic~\cite{} when $||$ is interpreted as
+separating conjunction. In fact, the second inference rule is, in any
+bi-Kleene algebra equivalent to the interchange axiom and the frame
+rule is derivable from this axiom as well~\cite{Hoareetal}. The
+precise connection between concurrent separation logic, a variant of
+concurrent Kleene algebra and a resource transformer model has been
+elaborated in~\cite{localitypaper}.
+
+A residuation for concurrent composition can be considered as well:
+\begin{equation*}
+x\|y \le z \Leftrightarrow y \le x/z.
+\end{equation*}
+The residual $x/z$ represents the weakest program such that when
+placed in parallel with $x$, the parallel composition behaves as $z$.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Algebra for Rely-Guarantee}
+\section{A  Rely-Guarantee Algebra}
 \label{sec:RG}
+
+In this section we show how bi-Kleene algebras can be expanded into a
+simple algebra that supports the derivation of rely-guarantee-style
+inference rules. This development does \emph{not} use the interchange
+law for several reasons. First, this law fails for fair parallel
+composition $x\parallel_f y$ in a model with possibly infinite, or
+non-terminating programs. In this model, $x \cdot y \not\leq
+x \parallel_f y$ in the case whenever $x$ is non-terminating. Second,
+it is not needed for deriving the usual concurrency laws. Third,
+whereas it may be assumed in the models outlined in Section *** it
+fails for the Aczel trace model in Section ***. \fbox{is that true?}
+
+\fbox{this needs to be reworked. two letters for a set are not good}
+
+\fbox{ the set RG doesn't need to be named in the algebra I think.}
+
+\fbox{we can just start with a bi-KA and say that elements are relies/guarantees if they satisfy properties}
+
+\fbox{what is a weak trioid?}
+
+\fbox{say that we don't consider the parallel star}
+
+\fbox{should we call relies and guarantees more generically
+  ``conditions'' can call them $c$?}
 
@@
 By convention, we use $r$ and $g$ to refer to elements of $RG$, and
-$x,y,z$ for arbitrary elements of $K$. Some elementary consequences of these rules are as follows
-\begin{align*}
-1 &\le r,\\
-r^\star = rr &= r = r\|r,\\
-r\|x^+ &= (r\|x)^+.
-\end{align*}
-
-%% Could use better explanations.
+$x,y,z$ for arbitrary elements of $K$.
 
@@
 except it deals with finite iteration.
+
+Some elementary consequences of these rules are as follows
+\begin{align*}
+1 &\le r,\\
+r^\star = rr &= r = r\|r,\\
+r\|x^+ &= (r\|x)^+.
+\end{align*}
+
+%% Could use better explanations.
 
@@
 inference rules can be derived.
-
-If we expand our rely-guarantee algebras to have residuals axiomatised
-by Galois connections for both parallel and sequential composition,
-i.e.
-\begin{align}
-xy \le z &\iff y \le x \rightarrow z, \label{res1}\\
-xy \le z &\iff x \le z \leftarrow y, \label{res2}\\
-x\|y \le z &\iff y \le x/z \label{res3},
-\end{align}
-then for sequential composition, (\ref{res1}) and (\ref{res2}) provide
-a link between weakest-precondition style semantics and hoare
-logic. As mentioned in Section \ref{sec:KA}, the triple $\{p\} c
-\{q\}$ can be encoded as $pc \le q$, while $q \leftarrow c$ represents
-$wlp(c,q)$. Equation (\ref{res2}) can thus be written as $\{p\} c
-\{q\} \iff p \le wlp(c,q)$. The parallel residual $x/z$ can be
-understood similarly as follows; it is the weakest program such that
-when placed in parallel with $x$, the composition behaves as $z$.
 
@@
 
+CNPq EPSRC
+
 \bibliography{paper}{}

### 2013-11-12 Paper/paper.tex
Almost finished paper

@@
 \DeclareMathOperator{\letters}{letters}
-\DeclareMathOperator{\wlp}{wlp}
-\DeclareMathOperator{\test}{test}
-\DeclareMathOperator{\eval}{eval}
-\DeclareMathOperator{\Id}{Id}
-\DeclareMathOperator{\edn}{end}
+
+\newcommand{\wlp}{\mathsf{wlp}}
+
+\newcommand{\test}{\mathsf{test}}
+\newcommand{\eval}{\mathsf{eval}}
+\newcommand{\Id}{\mathsf{Id}}
+\newcommand{\edn}{\mathsf{end}}
+\newcommand{\unchanged}{\mathsf{unchanged}}
+\newcommand{\preserves}{\mathsf{preserves}}
 
@@
 
-We aim to develop tools which support reasoning about concurrent
-programs at varying levels of abstraction. At the highest level,
-abstract algebra provides a framework for the prototyping of new
-inference and refinement rules, whereas at the lowest level concrete
-models of program stores allow for fine-grained reasoning about
-program data flow. We have implemented this approach is the
-Isabelle/HOL interactive theorem prover. Isabelle/HOL not only
-provides us with a high degree of confidence in the correctness of our
-theorems, but also supports the construction of custom proof tactics
-and tools for program verification tasks. Our approach differs from
-existing frameworks for rely-guarantee reasoning in
-Isabelle~\cite{nieto}, both by making the link between concrete
-programs and algebras explicit, but also by allowing arbitrary nested
-parallelism.
+We aim to develop lightweight formal methods which support
+rely-guarantee style reasoning about concurrent programs at varying
+levels of abstraction. At the most abstract level, we use algebras as
+a framework for reasoning about the control flow of programs as well
+as for deriving inference rules and refinement laws. At the lowest
+level, concrete models of program stores support fine-grained
+reasoning about program data flow and inteference. We have implemented
+this approach is the Isabelle/HOL interactive theorem proving
+environment. Isabelle not only provides us with a high degree of
+confidence in the correctness of our development and tool prototype,
+it also supports the construction of custom proof tactics and
+procedures for program verification and refinement tasks. Our approach
+differs from existing frameworks for rely-guarantee reasoning in
+Isabelle~\cite{nieto_rely-guarantee_2003}, both by making the link
+between concrete programs and algebras explicit, and by allowing
+arbitrary nested parallelism.
 
-For sequential programs the applicability of algebra, and Kleene
+For sequential programs, the applicability of algebra, and Kleene
 algebra in particular, has been known for over a decade. Kleene
@@
 verification, and allow the derivation of program equivalences and
-refinement rules. Furthermore, Kleene algebra has been applied in many
-applications, such as compiler
-optimisation~\cite{kozen_certification_2000}, program construction,
-transformation and termination analysis, and static analysis.
+refinement rules~\cite{kozen_kleene_1997,hoare_concurrent_2011}. Kleene algebras have
+been applied in applications including compiler
+optimisation~\cite{kozen_certification_2000}, program
+construction~\cite{berghammer_automated_2010}, transformation and
+termination analysis~\cite{desharnais_algebraic_2011}, and static
+analysis~\cite{fernandes_describing_2007}. Formalisations and tools are
+available in interactive theorem provers such as Coq~\cite{pous_kleene_2013} and
+Isabelle~\cite{armstrong_kleene_2013,armstrong_program_2013,armstrong_algebras_2014}.
 
-Algebra has also been applied in the study of separation logic, for
-example, as the separation algebras in~\cite{ohearn}, or in algebraic
-separation logic.
+Algebra also plays a role in the study of separation logic, for
+example, as the separation algebras in~\cite{calcagno_local_2007}, and in algebraic
+separation logic~\cite{dang_algebraic_2011}. More recently,
+concurrent Kleene algebras have given an algebraic account of some
+aspects of concurrent separation logic~\cite{hoare_concurrent_2011,hoare_locality_2011} and
+the first steps towards an algebraic description of rely-guarantee
+based reasoning have been undertaken~\cite{hoare_concurrent_2011}.
 
-\fbox{Algebraic Separation Logic}
-
-\fbox{Concurrent Kleene Algebra}
-
-Rely-guarantee is a so called \emph{compositional} method, where the
-verification of a large program is reduced the independent
-verification of individual subprograms, with the interference from
-other subprograms being abstractly captured by \emph{rely} and
-\emph{guarantee} conditions. This form of compositionality is,
-however, weaker than the strong notion of compositionality inherent in
-algebra, where full congruence rules exist for every operator.
+One main characteristic of the rely-guarantee method is
+compositionality. The verification of large concurrent programs is
+reduced to the independent verification of individual subprograms,
+where the effect of interferences from other subprograms are captured
+by \emph{rely} and \emph{guarantee} conditions, which constrain the
+behaviour of subprograms. This form of compositionality is, however,
+weaker than the strong notion of compositionality inherent in algebra,
+where full congruence rules exist for every operator and refinement is
+usually based on monotonicity.
 
 In this paper, we investigate algebraic principles for rely-guarantee
-style reasoning. This has been explored previously, for example
-in~\cite{cka}, but we extend this approach by considering a minimal
-set of rules sufficient to derive the standard rely-guarantee
-inference rules. Furthermore, we explore how the algebraic
-compositionality of this approach can be broken in just the right way,
-so as to capture the ideas of abstract interference conditions in
-rely-guarantee reasoning.
+style reasoning. First step have already been taken, for example,
+in~\cite{hoare_concurrent_2011}, but we extend this approach by extracting a minimal set
+of axioms for relies and guarantees sufficient to derive the standard
+rely-guarantee inference rules. Furthermore, we explore how the
+algebraic compositionality of this approach can be broken in just the
+right way, so as to capture the ideas of abstract interference
+conditions in rely-guarantee reasoning.
 
@@
 requires complex semantics involving synchronisation, termination
-analysis and complicated data structures.
-
-\fbox{We verify simple programs}
-
-\fbox{Coherent framework for programs}
+analysis and intricate data structures. For simplicity, in this paper
+we eschew such models in favor of relatively simple language and trace
+based semantics. Despite the simplicity of this model, we demonstrate
+and evaluate our prototypical verification tool implemented in
+Isabelle by verifiying simple two examples from the literature. Beyond
+that our approach provides a coherent framework from which more
+complex and detailed models can be integrated in the future.
 
@@
 \label{sec:KA}
-
-\fbox{define weak trioid!}
 
@@
 abortive action and $1$ as skip. In this context it is essential that
-multiplication is not commutative.  More formally, it is well known
-that (regular) languages with language union as $+$ and language
-product as $\cdot$, the empty language as $0$ and the empty word
-language $\{\varepsilon\}$ as $1$ form dioids. Another model is formed
-by binary relations with the union of relations as $+$, the product of
+multiplication is not commutative. Often we use $;$ in place of
+$\cdot$ when discussing programs. More formally, it is well known that
+(regular) languages with language union as $+$ and language product as
+$\cdot$, the empty language as $0$ and the empty word language
+$\{\varepsilon\}$ as $1$ form dioids. Another model is formed by
+binary relations with the union of relations as $+$, the product of
 relations as $\cdot$, the empty relation as $0$ and the identity
@@
 concurrency in which sequential and concurrent composition
-interact.\fbox{should we write $;$ instead of $\cdot$?}
+interact.
 
@@
 hand, Gischer has shown that trioids are sound and complete for the
-equational theory of series-parallel pomset languages~\cite{Gischer},
+equational theory of series-parallel pomset languages~\cite{gischer_equational_1988},
 which form a well studied model of true concurrency. On the other
@@
 holds for (regular) languages in which $||$ is interpreted as the
-shuffle or interleaving operation~\cite{Gischer}.
+shuffle or interleaving operation~\cite{gischer_shuffle_1981}.
 
-Formally, the \emph{shuffle} $\sha$ of two finite words is defined
+Formally, the \emph{shuffle} $\|$ of two finite words is defined
 inductively as
 \begin{equation*}
-\epsilon \sha s = \{s\}, \qquad s \sha \epsilon = \{s\},\qquad
-as \sha bt = a(s \sha bt) \cup b(as \sha t),
+\epsilon \| s = \{s\}, \qquad s \| \epsilon = \{s\},\qquad
+as \| bt = a(s \| bt) \cup b(as \| t),
 \end{equation*}
@@
 \begin{equation*}
-X \sha Y = \{x \sha y:\; x \in X \land x \in Y\}.
+X \| Y = \{x \| y:\; x \in X \land x \in Y\}.
 \end{equation*}
-\fbox{I would use $||$ here}
 
@@
 relations with the reflexive transitive closure operation as the
-star~\cite{Kozen}. Moreover, commutative Kleene algebras are
-complete with respect to the equational theory of regular languages
-over multisets~\cite{Conway}. It follows that equations in
-(commutative) Kleene algebras are decidable.
+star~\cite{kozen_completeness_1994}. Moreover, commutative Kleene
+algebras are complete with respect to the equational theory of regular
+languages over multisets~\cite{conway_regular_1971}. It follows that
+equations in (commutative) Kleene algebras are decidable.
 
 A \emph{bi-Kleene algebra} is a structure
-$(K,+,\cdot,||,0,1,^*,^{(*)})$ such that $(K,+,\cdot,0,1,^*)$ is a
-Kleene algebra and $(K,+,||,0,1,^{(*)})$ a commutative Kleene
+$(K,+,\cdot,||,0,1,\hphantom{}^*,\hphantom{}^{(*)})$ where
+$(K,+,\cdot,0,1,\hphantom{}^*)$ is a Kleene algebra and
+$(K,+,||,0,1,\hphantom{}^{(*)})$ is a commutative Kleene
 algebra. Bi-Kleene algebras are sound and complete with respect to the
@@
 \emph{concurrent Kleene algebra} is a bi-Kleene algebra which
-satisfies the interchange law~\cite{Hoareetal}. It can be shown that
+satisfies the interchange law~\cite{hoare_concurrent_2011}. It can be shown that
 shuffle languages and regular series-parallel pomset languages with a
@@
 In some contexts, it is also useful to add a meet operation $\sqcap$,
-such that $(D,+,\sqcap)$ is a lattice. \fbox{expand}
+such that $(D,+,\sqcap)$ is a distributive lattice. This is
+particularly needed in the context of refinement, where we typically
+want to represent specifications as well as programs.
 
@@
 (except the assignment axiom) can be derived in expansions of Kleene
-algebras. One approach is as follows~\cite{MoellerStruth}. Suppose a
+algebras. One approach is as follows~\cite{moller_algebras_2006}. Suppose a
 suitable Boolean algebra $B$ of \emph{tests} has been embedded into a
@@
   algebras} and that the inference rules of Hoare logic can be
-derived~\cite{MoellerStruth}.
+derived~\cite{moller_algebras_2006}.
 
@@
 rules of sequential Hoare logic except the assignment axiom can still
-be derived in this setting~\cite{Hoareetal}.  Tarlecki's motivating
+be derived in this setting~\cite{hoare_concurrent_2011}. Tarlecki's motivating
 explanations carry over to the algebraic approach.
@@
 \begin{equation*}
-  \frac{\{x\cdot t\}y\{x\}}{\{x\}(t\cdot y)^\ast\{t'\cdot x\}}
+  \frac{\vdash \{x\cdot t\}y\{x\}}{\vdash \{x\}(t\cdot y)^\ast\{t'\cdot x\}}
 \end{equation*}
@@
 of residuation.  This can be achieved in the context of \emph{action
-  algebras}~\cite{pratt}, which expand Kleene algebras by operations
+  algebras}~\cite{pratt_action_1990}, which expand Kleene algebras by operations
 of left and right residuation defined by the Galois connections
 \begin{equation*}
-  xy\le z\Leftrightarrow x\le z\leftarrow y,\qquad xy\le z \Leftrightarrow y\le z\rightarrow x.
+  x\cdot y\le z\Leftrightarrow x\le z\leftarrow y,\qquad x\cdot y\le z \Leftrightarrow y\le x\rightarrow z.
 \end{equation*}
-\fbox{please check that I didn't confuse these arrows} These
-residuals, and now even the Kleene star can be axiomatised
+These residuals, and now even the Kleene star can be axiomatised
 equationally in action algebras. For a comprehensive list of the
 properties of action algebras and their most important models
-see~\cite{archive}, including the language and the relational
+see~\cite{armstrong_kleene_2013}, including the language and the relational
 model. In analogy to the development in model Kleene algebra we can
@@
 bi-Kleene algebra equivalent to the interchange axiom and the frame
-rule is derivable from this axiom as well~\cite{Hoareetal}. The
+rule is derivable from this axiom as well~\cite{hoare_concurrent_2011}. The
 precise connection between concurrent separation logic, a variant of
 concurrent Kleene algebra and a resource transformer model has been
-elaborated in~\cite{localitypaper}.
+elaborated in~\cite{hoare_locality_2011}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{A  Rely-Guarantee Algebra}
@@
 composition $x\parallel_f y$ in a model with possibly infinite, or
-non-terminating programs. In this model, $x \cdot y \not\leq
-x \parallel_f y$ in the case whenever $x$ is non-terminating. Second,
-it is not needed for deriving the usual concurrency laws. Third,
-whereas it may be assumed in the models outlined in Section *** it
-fails for the Aczel trace model in Section ***. \fbox{is that true?}
-
-\fbox{this needs to be reworked. two letters for a set are not good}
-
-\fbox{ the set RG doesn't need to be named in the algebra I think.}
-
-\fbox{we can just start with a bi-KA and say that elements are relies/guarantees if they satisfy properties}
-
-\fbox{what is a weak trioid?}
-
-\fbox{say that we don't consider the parallel star}
-
-\fbox{should we call relies and guarantees more generically
-  ``conditions'' can call them $c$?}
+non-terminating programs. In this model, $x \cdot y \not\leq x
+\parallel_f y$ in the case whenever $x$ is non-terminating. Secondly,
+it is not needed for deriving the usual laws or rely-guarantee.
 
 A rely-guarantee algebra is a structure
-$(K,RG,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
-distributive lattice, $(K,+,\cdot,\|,0,1)$ is a weak trioid and
-$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. $RG$ is a distinguished
-subset of relys and guarantees which satisfy the following axioms
+$(K,C,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
+distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
+$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. Note that we do not
+consider the parallel star. $C$ is a distinguished subset of rely and
+guarantee conditions or \emph{interference constraints} which satisfy
+the following axioms
 \begin{align}
@@
 \end{align}
-By convention, we use $r$ and $g$ to refer to elements of $RG$, and
-$x,y,z$ for arbitrary elements of $K$.
+By convention, we use $r$ and $g$ to refer to elements of $C$,
+depending on whether they are used as relies or guarantees, and
+$x,y,z$ for arbitrary elements of $K$. The operators $\|$ and $\sqcap$
+must be closed with respect to $C$.
 
@@
 
-Some elementary consequences of these rules are as follows
-\begin{align*}
-1 &\le r,\\
-r^\star = rr &= r = r\|r,\\
-r\|x^+ &= (r\|x)^+.
-\end{align*}
-
-%% Could use better explanations.
+Some elementary consequences of these rules are
+\begin{equation*}
+1 \le r,\qquad
+r^\star = rr = r = r\|r,\qquad
+r\|x^+ = (r\|x)^+.
+\end{equation*}
 
@@
 show this, we use Isabelle's
-\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample generator
-to construct models which violate each of these axioms, yet satisfy
-all others. The models thus constructed for (\ref{rg1}) and
-(\ref{rg3}) are shown in Figures \ref{fig:rg1} and \ref{fig:rg3}
-respectively. The model constructed for (\ref{rg2}) is just the two
-element rely-guarantee algebra with $0$ and $1$.
+\emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample
+generator to construct models which violate each particular axiom
+while satisfying all others.
 
-\begin{figure}[H]
-\centering
-\begin{minipage}{0.24\textwidth}
-\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
-  \node (center) {};
-  \node [elem] (r1) at (90:1) {$r_1$};
-  \node [elem] (one) at (200:1) {$1$};
-  \node [elem] (r2) at (20:1) {$r_2$};
-  \node [elem] (zero) at (-90:1) {$0$};
-
-  \path [line] (r1) -- (r2);
-  \path [line] (r1) -- (one);
-  \path [line] (r1) -- (zero);
-  \path [line] (r2) -- (one);
-  \path [line] (r2) -- (zero);
-  \path [line] (one) -- (zero);
-\end{tikzpicture}
-\end{minipage}
-\begin{minipage}{0.24\textwidth}
-\begin{align*}
-r_1 \| r_1 &= r_1\\
-r_1 \| r_2 &= r_1\\
-r_2 \| r_1 &= r_1\\
-r_2 \| r_2 &= r_1
-\end{align*}
-\end{minipage}
-\begin{minipage}{0.24\textwidth}
-\begin{align*}
-r_1r_1 &= r_1\\
-r_1r_2 &= r_1\\
-r_2r_1 &= r_1\\
-r_2r_2 &= r_2
-\end{align*}
-\end{minipage}
-\begin{minipage}{0.24\textwidth}
-\begin{align*}
-r_1^\star &= r_1\\
-r_2^\star &= r_2
-\end{align*}
-\end{minipage}
-\caption{4 element counterexample for $r \in RG \implies r\|r \le r$}
-\label{fig:rg1}
-\end{figure}
-
-Axiom (\ref{rg4}) can be derived from (\ref{rg3}) in all finite
-models. This is because any finite $K$ is complete, and in a complete
-setting fixpoint laws can be used to show $r\|x^+ \le (r\|x)^+$. Thus
-it is impossible to construct a finite model demonstrating that
-(\ref{rg4}) is independent from (\ref{rg1}) -- (\ref{rg3}).
-
-\begin{figure}[t]
-\centering
-\begin{minipage}{0.24\textwidth}
-\begin{tikzpicture}[x=1.5cm,y=1.5cm,auto]
-  \node (center) {};
-  \node [elem] (r1) at (90:1) {$r_1$};
-  \node [elem] (one) at (0:1) {$1$};
-  \node [elem] (zero) at (-90:1) {$0$};
-
-  \path [line] (r1) -- (one);
-  \path [line] (r1) -- (zero);
-  \path [line] (r1) -- (zero);
-  \path [line] (one) -- (zero);
-\end{tikzpicture}
-\end{minipage}
-\begin{minipage}{0.24\textwidth}
-\begin{align*}
-r_1 \| r_1 &= r_1\\
-r_1 r_1 &= r_1\\
-r_1^\star &= r_1
-\end{align*}
-\end{minipage}
-\caption{3 element counterexample for $r \in RG \implies r\|xy = (r\|x)(r\|y)$}
-\label{fig:rg3}
-\end{figure}
+Axiom (\ref{rg4}) can be derived from (\ref{rg3}) by a simple
+inductive argument in quantales where the plus operation can be
+defined as a sum of powers $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$
+and $x^{i+1}=x\cdot x^i$. In first-order Kleene algebras, however, it
+is independent, but it is impossible to find a counterexample with Nitpick because all such counterexamples are finite, but all finite Kleene algebras are a forteriori quantales.
 
@@
 \end{align}
-This encoding allows us to think in terms of program refinement, as
+This encoding allows us to think in terms of program refinement~\cite{hayes_refining_2013}, as
 $r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
@@
 \end{figure}
-
-% Or should this be weak?
 
@@
   traces}~\cite{boer_formal_1999}, but lack the usual process
-labels. We denote the set of all consistent words by $C$.
+labels. We denote the set of all consistent words by $T$.
 
-As usual, the product of two languages X and Y is defined as
-\begin{align*}
-XY = \{x\frown y.\; x \in X \land y \in Y\}.
-\end{align*}
-The language product is associative, and distributes over arbitrary
-joins from both the left and the right, i.e.
-\begin{align*}
-X(\bigcup Y) = \bigcup\{XZ.\; Z \in Y\} \qquad \text{and} \qquad (\bigcup Y)X = \bigcup\{ZX.\; Z \in Y\}.
-\end{align*}
-The shuffle $\sha$ on a pair
-of finite words is defined inductively as
-\begin{align*}
-\epsilon \sha s = \{s\}, \qquad s \sha \epsilon = \{s\},\\
-as \sha bt = a(s \sha bt) \cup b(as \sha t),
-\end{align*}
-which is then lifted to languages X and Y as the shuffle product,
-\begin{align*}
-X \sha Y = \bigcup\{x \sha y.\; x \in X \land x \in Y\}.
-\end{align*}
-The shuffle product is associative, commutative, and distributes over
-arbitrary joins. Both products share the same unit, $\{\epsilon\}$ and
-zero, $\emptyset$. Proving properties of shuffle is suprisingly
-tricky (especially if one considers infinite words). For a in-depth
-treatment of the shuffle product see~\cite{shufflethings}. The
-above properties are enough to show that $(\mathcal{P}((\Sigma\times\Sigma)^*),
-\cup, \cdot,\sha, \emptyset, \{\epsilon\})$ forms a weak trioid.
+Sequential composition in this model is language product, as per
+usual. Concurrent composition is the shuffle product defined in
+section \ref{sec:KA}. The shuffle product is associative, commutative,
+and distributes over arbitrary joins. Both products share the same
+unit, $\{\epsilon\}$ and zero, $\emptyset$. Proving properties of
+shuffle is suprisingly tricky (especially if one considers infinite
+words). For a in-depth treatment of the shuffle product
+see~\cite{mateescu_shuffle-like_1997}.
+
+\begin{lem}
+$(\mathcal{P}((\Sigma\times\Sigma)^*), \cup, \cdot,\sha, \emptyset,
+  \{\epsilon\})$ forms a trioid.
+\end{lem}
 
@@
 lifts a relation $R$ to a language containing words of length one for
-each pair in $R$. The set of rely-guaratee elements RG is then defined
-as $\{r.\; \exists R. r = \langle R\rangle^*\}$. We can show that
-$(\mathcal{P}((\Sigma\times\Sigma)^*), RG, \cup, \cdot,\sha,
-\emptyset, \{\epsilon\}, C)$ is a rely-guarantee algebra.
+each pair in $R$. The set of rely-guaratee conditions $C$ is then defined
+as $\{r.\; \exists R. r = \langle R\rangle^*\}$.
 
+\begin{lem}
+$(\mathcal{P}((\Sigma\times\Sigma)^*), C, \cup, \cdot,\sha,
+  \emptyset, \{\epsilon\}, T)$ is a rely-guarantee algebra.
+\end{lem}
 Since $\langle R\rangle$ is atomic, it satisfies several useful properties, such as,
@@
 we inductively generate the mumble language $w^\dagger$ for a word $w$
-in a language over $\Sigma^2$ as follows: Assume $R,S,T \in \Sigma^2$
+in a language over $\Sigma^2$ as follows: Assume $R,S \in \Sigma^2$
 and $u,v,w \in (\Sigma^2)^*$. First, $w \in w^\dagger$. Secondly, if
@@
 \begin{align*}
-\text{if } P \text{ \{ } X \text { \} else \{ } Y \text{ \}} &= \test(P); X + \test(- P); Y\\
-\text{while } P \text{ \{ } X \text{ \}} &= (\test(P);X)^\star;\test(- P)
+\mathsf{if } P \text{ \{ } X \text { \} \textsf{else} \{ } Y \text{ \}} &= \test(P); X + \test(- P); Y\\
+\mathsf{while } P \text{ \{ } X \text{ \}} &= (\test(P);X)^\star;\test(- P)
 \end{align*}
@@
 \begin{align*}
-&\text{unchanged}(\text{vars}(e)) \cap \text{preserves}(P) \cap \text{preserves}(P[x/e]),\\
-&\text{unchanged}(- {x})\\
+&\unchanged(\mathsf{vars}(e)) \cap \preserves(P) \cap \preserves(P[x/e]),\\
+&\unchanged(- \{x\})\\
 &\vdash \{\edn(P)\} x := e \{\edn(P[x/e])\}.
@@
 
-The rely condition states that the environment is not allowed to
-modify any of the variables used when evaluating $e$, i.e. those
-variables must remain unchanged, and also that the environment must
-preserve the precondition and postcondition of the assignment
-statement. In turn, the assignment statement itself guarantees that it
-leaves every variable other than $x$ unchanged. Preserves and unchanged are defined as
+The rely condition states the following: First, the environment is not
+allowed to modify any of the variables used when evaluating $e$,
+i.e. those variables must remain unchanged. Second, that the
+environment must preserve the precondition. Third, that postcondition
+of the assignment statement is also preserved. In turn, the assignment
+statement itself guarantees that it leaves every variable other than
+$x$ unchanged. Preserves and unchanged are defined as
 \begin{align*}
-\text{preserves}(P) &= \langle \{(\sigma,\sigma').\; P(\sigma) \implies P(\sigma')\}\rangle^\star,\\
-\text{unchanged}(X) &= \langle \{(\sigma,\sigma').\; \forall v\in X.\; \sigma(v) = \sigma'(v)\}\rangle^\star.
+\preserves(P) &= \langle \{(\sigma,\sigma').\; P(\sigma) \implies P(\sigma')\}\rangle^\star,\\
+\unchanged(X) &= \langle \{(\sigma,\sigma').\; \forall v\in X.\; \sigma(v) = \sigma'(v)\}\rangle^\star.
 \end{align*}
 We also defined two futher rely conditions, increasing and decreasing,
-which are defined much like unchanged except they only require that variables
-increase or decrease, rather than stay the same.
+which are defined much like unchanged except they only require that
+variables increase or decrease, rather than stay the same. Note that
+we can easily define other useful assignment rules---if we know things
+about $P$ and $e$, we can make stronger guarantees about what $x := e$
+can do. For example the assignment $x := x - 2$ can also guarantee
+that $x$ will always decrease.
 
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Example}
+\section{Examples}
+
+\fbox{Simple assignment example}
 
 Figure \ref{fig:findp} shows the classic FINDP program, which has been
-used by numerous authors including Owicki~\cite{owicki},
-Jones~\cite{jones}, de Roever~\cite{derover} and
-Hayes~\cite{hayes}. The program finds the least element of an array
+used by numerous authors including Owicki~\cite{owicki_axiomatic_1975},
+Jones~\cite{jones_development_1981}, de Roever~\cite{de_roever_concurrency_2001} and
+Hayes~\cite{hayes_refining_2013}. The program finds the least element of an array
 satisfying the predicate $P$. The index of the first element
@@
 array satisfying $P$ which is less than $i_A$.
-
 \begin{figure}
@@
 \begin{aligned}
-&f_A := \text{len}(\text{array});\\
-&f_B := \text{len}(\text{array});\\
+&f_A := \mathsf{len}(\mathsf{array});\\
+&f_B := \mathsf{len}(\mathsf{array});\\
 &\left(
@@
 &i_A = 0\\
-&\text{while } i_A < f_A \land i_A < f_B \text{ \{ }\\
-&\qquad\text{if } P(\text{array}[i_A]) \text{ \{}\\
+&\text{\textsf{while} } i_A < f_A \land i_A < f_B \text{ \{ }\\
+&\qquad\text{\textsf{if} } P(\mathsf{array}[i_A]) \text{ \{}\\
 &\qquad\qquad f_A := i_A\\
-&\qquad\text{\} else \{}\\
+&\qquad\text{\} \textsf{else} \{}\\
 &\qquad\qquad i_A := i_A + 2\\
@@
 &i_B = 1\\
-&\text{while } i_B < f_A \land i_B < f_B \text{ \{}\\
-&\qquad\text{if } P(\text{array}[i_B]) \text{ \{}\\
+&\text{\textsf{while} } i_B < f_A \land i_B < f_B \text{ \{}\\
+&\qquad\text{\textsf{if} } P(\mathsf{array}[i_B]) \text{ \{}\\
 &\qquad\qquad f_B := i_B\\
-&\qquad\text{\} else \{}\\
+&\qquad\text{\} \textsf{else} \{}\\
 &\qquad\qquad i_B := i_B + 2\\
@@
 
+Here, we sketch the main outline of the correctness proof, and comment
+on its implementation in Isabelle. We don't attempt to give a fully
+detailed proof, as this has been done many times previously.
+
 To prove the correctness of FINDP, we must show that
 \begin{align*}
-\text{FINDP} \le_\pi \text{ends}\{\sigma.\; \text{leastp}(\sigma, f)\} + \text{ends}\{\sigma.\; \sigma(f) = \text{len}(\text{array})\}.
+\text{FINDP} \le_\pi \edn(\mathsf{leastP}(f)) + \edn(f = \mathsf{len}(\text{array})).
 \end{align*}
-where
-\begin{align*}
-\text{leastp}(\sigma, x) =  (\forall v<\sigma(x). \lnot P(\sigma(\text{array}[v])))\land P(\sigma(\text{array}[\sigma(x)])).
-\end{align*}
-The interesting part of this proof is the verification
+where $\mathsf{leastP(f)}$ is the set of states where $f$ is the least
+index satisfying $P$, and $f = len(array)$ is the set of states where
+$f$ is the length of the array. In other words, either we find the
+least element, or no elements in the array satisfy $P$, in which case
+$f$ remains the same as the length of the array.
 
-\newpage
+To prove the parallel part of the program, subprogram $A$ guarantees
+that it does not modify any of the variables used by subprogram $B$,
+except for $f_A$, which it guarantees will only ever
+decrease. Subprogram $B$ makes effectively the same guarantee to
+$A$. Under these interference constraints we then prove that $A$ or
+$B$ will find the lowest even or odd index which satisfies $P$
+respectively---or they do not find it, in which case $f_A$ or $f_B$
+will remain equal to the length of the array.
+
+Despite the seemingly straightforward nature of this proof, this turns
+out to be suprising difficult to prove in Isabelle. Each atomic step
+needs to be shown to satisfy the guarantee of its containing
+subprogram, as well as any goals relating to it's pre and post
+conditions. This invariably leads to a proliferation of many numerous
+proof goals, even for such a small program. More work needs to be done
+to manage the complexity of such proofs within interactive theorem
+provers. Existing work on combining rely-guarantee with separation
+logic~\cite{vafeiadis_modular_2008} may proof useful here. Our language model is
+sufficiently generic such that arbitrary models of stores may be used,
+including those common in separation logic, which have already been
+implemented in Isabelle.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-11-12 Paper/paper.tex
Almost finished paper 2

@@
 \newtheorem{lem}{Lemma}
+\newtheorem{thm}{Theorem}
 
@@
 \cdot y \le y\Rightarrow x^*\cdot z \le y$ and $z + y\cdot x \le y
-z\cdot \Rightarrow x^* \le y$. It follows that $1+x\cdot x^* = x^*$
+\Rightarrow z\cdot x^* \le y$. It follows that $1+x\cdot x^* = x^*$
 and that the right unfold axiom $1+x^*\cdot x=x^*$ is derivable as
@@
 A rely-guarantee algebra is a structure
-$(K,C,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
+$(K,I,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
 distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
 $(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. Note that we do not
-consider the parallel star. $C$ is a distinguished subset of rely and
+consider the parallel star. $I$ is a distinguished subset of rely and
 guarantee conditions or \emph{interference constraints} which satisfy
@@
 \end{align}
-By convention, we use $r$ and $g$ to refer to elements of $C$,
+By convention, we use $r$ and $g$ to refer to elements of $I$,
 depending on whether they are used as relies or guarantees, and
 $x,y,z$ for arbitrary elements of $K$. The operators $\|$ and $\sqcap$
-must be closed with respect to $C$.
+must be closed with respect to $I$.
 
-Axiom (\ref{rg1}) states that interference from a rely condition being
-run twice in parallel is no different from just the interference from
-that rely condition begin run once in parallel. Axiom (\ref{rg2})
-states that interference from a single rely condition is less than
-interference from itself and another rely condition. Axiom (\ref{rg3})
-allows a rely or guarantee to be split across sequential
+Axiom (\ref{rg1}) states that interference from a constraint being run
+twice in parallel is no different from just the interference from that
+constraint begin run once in parallel. Axiom (\ref{rg2}) states that
+interference from a single constraint is less than interference from
+itself and another interference constraint. Axiom (\ref{rg3}) allows
+an interference constraint to be split across sequential
 programs. Axiom (\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent,
@@
 
-Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent. To
-show this, we use Isabelle's
+\begin{thm}
+  Axioms (\ref{rg1}), (\ref{rg2}) and (\ref{rg3}) are independent.
+\end{thm}
+\begin{proof}
+  To show this, we use Isabelle's
   \emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample
@@
   while satisfying all others.
+\end{proof}
 
+\begin{thm}
 Axiom (\ref{rg4}) can be derived from (\ref{rg3}) by a simple
-inductive argument in quantales where the plus operation can be
-defined as a sum of powers $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$
-and $x^{i+1}=x\cdot x^i$. In first-order Kleene algebras, however, it
-is independent, but it is impossible to find a counterexample with Nitpick because all such counterexamples are finite, but all finite Kleene algebras are a forteriori quantales.
+inductive argument in a quantale where $\|$ distributes over arbitrary suprema.
+\end{thm}
+\begin{proof}
+  In a quantale $x^+$ can be defined as a sum of powers
+  $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$ and $x^{i+1}=x\cdot x^i$. By
+  induction on w$i$ we get $r\|x^i = (r\|x)^i$, hence
+  \begin{align*}
+    r\|x^+ = r\|\sum_{i\ge 1} x^i = \sum_{i\ge 1} r\|x^i = \sum_{i\ge 1} (r\|x)^i = (r\|x)^+.
+  \end{align*}
+\end{proof}
+
+In first-order Kleene algebras (\ref{rg3}) and (\ref{rg4}) are
+independent, but it is impossible to find a counterexample with
+Nitpick because it generates only finite counterexamples, and all
+finite Kleene algebras are a forteriori quantales.
 
@@
 \end{align}
+
+\begin{thm}
 With this encoding we can derive the standard rely-guarantee inference
-rules, as shown in Figure \ref{fig:rgrules}. Thus equations
-(\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
-rules, somehow represent a minimal set of axioms from which these
-inference rules can be derived.
+rules, as shown in Figure \ref{fig:rgrules}.
+\end{thm}
 
-By using these residuals quintuples can be encoded in the following way,
-which is equivalent to the encoding in Equation (\ref{quin}).
-\begin{align}
-r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
-\end{align}
-This encoding allows us to think in terms of program refinement~\cite{hayes_refining_2013}, as
-$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
-placed in parallel with interference from $r$, and guaranteeing
-interference at most $g$, goes from $p$ to $q$---a generic
-specification for a concurrent program.
-
-\begin{figure}[tbh]
+\begin{figure}[hbt]
 \centering
@@
 
+Thus equations
+(\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
+rules, somehow represent a minimalistic set of axioms from which these
+inference rules can be derived.
+
+If we add residuals to our algebra quintuples can be encoded in the
+following way, which is equivalent to the encoding in Equation
+(\ref{quin}).
+\begin{align}
+r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
+\end{align}
+This encoding allows us to think in terms of program refinement, as in~\cite{hayes_refining_2013}, since
+$r/(p \rightarrow q) \sqcap g$ defines the weakest program that when
+placed in parallel with interference from $r$, and guaranteeing
+interference at most $g$, goes from $p$ to $q$---a generic
+specification for a concurrent program.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 the case, but breaking such inherent compositionality in just the
-right way to capture interesting properties of concurrency requires
+right way to capture interesting properties of interference requires
 extra work.
@@
 \begin{align*}
-r, g \vdash \{p\} c \{q\} \iff \pi(p(r\|c)) \le \pi(q) \land c \le g.
+r, g \vdash \{p\} x \{q\} \iff p(r\|c) \le_\pi q \land x \le g.
 \end{align*}
-Since for any operator $\bullet$, it is not required that
+Where $x \le_\pi y$ is $\pi(x) \le \pi(y)$. Since for any operator
+$\bullet$, it is not required that
 \begin{align*}
@@
 we can `break' compositionality in just the right way, provided we
-choose the correct properties for $\pi$. For example, if we consider
-the following seven axioms
+choose the correct properties for $\pi$. However, if we simply listed
+all properties that $\pi$ needs to satisfy we would end up with an
+overly long list of axioms to add to our algebra. Instead we define
+the function $\pi$ as $\lambda x.\; x \sqcap c$ where $c$ is the
+greatest $consistent$ element such that
 \begin{align}
-\pi(\pi(x)) &= \pi(x), \label{pi1}\\
-x \le y \implies \pi(x) &\le \pi(y), \label{pi2}\\
-\pi(x) &\le x, \label{pi3}\\
-\pi(x^*) &\le \pi(\pi(x)^*), \label{pi4}\\
-\pi(xy) &\le \pi(\pi(x)\pi(y)), \label{pi5}\\
-\pi(x \sqcap y) &= \pi(x) \sqcap \pi(y). \label{pi6}
+x^* &\le_\pi \pi(x)^*, \label{con1}\\
+xy &\le_\pi \pi(x)\pi(y). \label{con2}\\
+z + x\cdot y \le_\pi y &\implies x^*\cdot z \le_\pi y\label{con3}\\
+z + y\cdot x \le_\pi y &\implies z\cdot x^* \le_\pi y\label{con4}
 \end{align}
 
-Axioms (\ref{pi1}) to (\ref{pi3}) show that $\pi$ is an interior
-operator, while (\ref{pi4}) and (\ref{pi5}) allow us to introduce
-$\pi$ as needed around sequential composition and finite iteration,
-which allows us to substitute under those operators as required---we
-only want to break compositionality for $\|$.
-
-The above axioms indicate that we could define $\pi$ as a function
-$\lambda x.\; x \sqcap c$ where $c$ satisfies
-\begin{align}
-x^* \sqcap c &\le (x \sqcap c)^* \sqcap c, \label{con1}\\
-xy \sqcap c &\le (x \sqcap c)(y \sqcap c) \sqcap c. \label{con2}
-\end{align}
-
+For any operator $\bullet$, we write $x \bullet_\pi y$ for the
+operator $\pi(x\bullet y)$, and we write $x^\pi$ for
+$\pi(x^\star)$.
+\begin{thm}
+Axioms (\ref{con1}) to (\ref{con4}) induce a Kleene subalgebra.
+\begin{align*}
+(\{x\in K.\; \pi(x) = x\}, +_\pi, \cdot_\pi, \hphantom{}^\pi,0,1)
+\end{align*}
+\end{thm}
 As such we redefine our rely-guarantee algebra from section
-\ref{sec:RG} as a structure $(K,RG,+,\sqcap,\cdot,\|,^\star,0,1,c)$
+\ref{sec:RG} as a structure $(K,RG,+,\sqcap,\cdot,\|,\hphantom{}^\star,0,1,c)$
 which, in addition to the rules in section \ref{sec:RG} satisfies both
-(\ref{con1}) and (\ref{con2}).
-
+(\ref{con1}) to (\ref{con4}).
+\begin{thm}
 All the rules in Figure \ref{fig:rgrules} can be derived with this
@@
 mutandis.
+\end{thm}
 
@@
 
-CNPq EPSRC
+We have...
+
+\paragraph*{Acknowledgements} This work was supported by funding from
+CNPq Brazil and the EPSRC.
+
+\newpage
 

### 2013-11-15 Paper/paper.tex
Finished and submitted paper

@@
 
-We aim to develop lightweight formal methods which support
-rely-guarantee style reasoning about concurrent programs at varying
-levels of abstraction. At the most abstract level, we use algebras as
-a framework for reasoning about the control flow of programs as well
-as for deriving inference rules and refinement laws. At the lowest
-level, concrete models of program stores support fine-grained
-reasoning about program data flow and inteference. We have implemented
-this approach is the Isabelle/HOL interactive theorem proving
-environment. Isabelle not only provides us with a high degree of
-confidence in the correctness of our development and tool prototype,
-it also supports the construction of custom proof tactics and
-procedures for program verification and refinement tasks. Our approach
-differs from existing frameworks for rely-guarantee reasoning in
-Isabelle~\cite{nieto_rely-guarantee_2003}, both by making the link
-between concrete programs and algebras explicit, and by allowing
-arbitrary nested parallelism.
+Extensions of Hoare logics are becoming increasingly important for the
+verification and development of concurrent and multiprocessor
+programs. One of the most popular extensions is Jones' rely-guarantee
+method~\cite{jones_development_1981}.  A main benefit of this method is
+compositionality: the verification of large concurrent programs can be
+reduced to the independent verification of individual subprograms. The
+effect of interactions or interference between subprograms is captured
+by \emph{rely} and \emph{guarantee} conditions. Rely conditions describe
+the effect of the environment on an individual subprogram. Guarantee conditions,
+in turn, describe the effect of an individual subprogram on the
+environment. By constraining a subprogram by a rely condition, the global
+effect of interactions is captured locally.
+
+To make this method applicable to concrete program development and
+verification tasks, its integration into tools is essential. To
+capture the flexibility of the method, a number of features seem
+desirable. First, we need to implement solid mathematical models for
+fine-grained program behaviour. Second, we would like an abstract
+layer at which inference rules and refinement laws can be derived
+easily. Third, a high degree of proof automation is mandatory for the
+analysis of concrete programs. In the context of the rely-guarantee
+method, tools with these important features are currently missing.
+
+This paper presents a novel approach for providing such a tool
+integration in the interactive theorem proving environment
+Isabelle/HOL. At the most abstract level, we use algebras to reason
+about the control flow of programs as well as for deriving inference
+rules and refinement laws. At the most concrete level, detailed models
+of program stores support fine-grained reasoning about program data
+flow and interference. These models are then linked with the
+algebras. Isabelle allows us to implement these layers in a modular
+way and relate them formally with one another. It not only provides us
+with a high degree of confidence in the correctness of our
+development, it also supports the construction of custom proof tactics
+and procedures for program verification and refinement tasks.
 
 For sequential programs, the applicability of algebra, and Kleene
-algebra in particular, has been known for over a decade. Kleene
-algebra provides operations for non-deterministic choice, sequential
+algebra in particular, has been known for decades. Kleene algebra
+provides operations for non-deterministic choice, sequential
 composition and finite iteration, in addition to skip and abort. With
-appropriate extensions, Kleene algebras support Hoare-style program
-verification, and allow the derivation of program equivalences and
-refinement rules~\cite{kozen_kleene_1997,hoare_concurrent_2011}. Kleene algebras have
-been applied in applications including compiler
-optimisation~\cite{kozen_certification_2000}, program
-construction~\cite{berghammer_automated_2010}, transformation and
-termination analysis~\cite{desharnais_algebraic_2011}, and static
-analysis~\cite{fernandes_describing_2007}. Formalisations and tools are
-available in interactive theorem provers such as Coq~\cite{pous_kleene_2013} and
-Isabelle~\cite{armstrong_kleene_2013,armstrong_program_2013,armstrong_algebras_2014}.
+appropriate extensions, Kleene algebras support Hoare-style
+verification of sequential programs, and allow the derivation of
+program equivalences and refinement
+rules~\cite{kozen_kleene_1997,hoare_concurrent_2011}. Kleene algebras
+have been used in applications including compiler optimisation,
+program construction, transformation and termination analysis, and
+static analysis. Formalisations and tools are available in interactive
+theorem provers such as Coq~\cite{pous_kleene_2013} and
+Isabelle~\cite{armstrong_kleene_2013,armstrong_program_2013,armstrong_algebras_2014}. A
+first step towards an algebraic description of rely-guarantee based
+reasoning has recently been undertaken~\cite{hoare_concurrent_2011}.
 
-Algebra also plays a role in the study of separation logic, for
-example, as the separation algebras in~\cite{calcagno_local_2007}, and in algebraic
-separation logic~\cite{dang_algebraic_2011}. More recently,
-concurrent Kleene algebras have given an algebraic account of some
-aspects of concurrent separation logic~\cite{hoare_concurrent_2011,hoare_locality_2011} and
-the first steps towards an algebraic description of rely-guarantee
-based reasoning have been undertaken~\cite{hoare_concurrent_2011}.
+The main contributions of this paper are as follows. First, we
+investigate algebraic principles for rely-guarantee style
+reasoning. Starting from~\cite{hoare_concurrent_2011} we extract a basic
+minimal set of axioms for rely and guarantee conditions which suffice
+to derive the standard rely-guarantee inference rules. These axioms
+provide valuable insights into the conceptual and operational role of
+these constraints. However, algebra is inherently compositional, so it
+turns out that these axioms do not fully capture the semantics of
+interference in execution traces. We therefore explore how the
+compositionality of these axioms can be broken in the right way, so as to capture the
+intended trace semantics.
 
-One main characteristic of the rely-guarantee method is
-compositionality. The verification of large concurrent programs is
-reduced to the independent verification of individual subprograms,
-where the effect of interferences from other subprograms are captured
-by \emph{rely} and \emph{guarantee} conditions, which constrain the
-behaviour of subprograms. This form of compositionality is, however,
-weaker than the strong notion of compositionality inherent in algebra,
-where full congruence rules exist for every operator and refinement is
-usually based on monotonicity.
+Second, we link our rely-guarantee algebras with a simple trace based
+semantics which so far is restricted to finite executions and
+disregards termination and synchronisation. Despite the simplicity of
+this model, we demonstrate and evaluate our prototypical verification
+tool implemented in Isabelle by verifying two simple examples from
+the literature. Beyond that our approach provides a coherent framework
+from which more complex and detailed models can be implemented in the
+future.
 
-In this paper, we investigate algebraic principles for rely-guarantee
-style reasoning. First step have already been taken, for example,
-in~\cite{hoare_concurrent_2011}, but we extend this approach by extracting a minimal set
-of axioms for relies and guarantees sufficient to derive the standard
-rely-guarantee inference rules. Furthermore, we explore how the
-algebraic compositionality of this approach can be broken in just the
-right way, so as to capture the ideas of abstract interference
-conditions in rely-guarantee reasoning.
+Third, we derive the usual inference rules of the rely-guarantee
+method with the exception of assignment axioms directly from the
+algebra, and obtain assignment axioms from our models. Our
+formalisation in Isabelle allows us to reason seamlessly across these
+layers, which includes the data flow and the control flow of concurrent
+programs.
 
-The verification and construction of actual concurrent programs often
-requires complex semantics involving synchronisation, termination
-analysis and intricate data structures. For simplicity, in this paper
-we eschew such models in favor of relatively simple language and trace
-based semantics. Despite the simplicity of this model, we demonstrate
-and evaluate our prototypical verification tool implemented in
-Isabelle by verifiying simple two examples from the literature. Beyond
-that our approach provides a coherent framework from which more
-complex and detailed models can be integrated in the future.
+Taken together, our Isabelle implementation constitutes a tool
+prototype for the verification and construction of concurrent
+programs. We illustrate the tool with two simple examples from the
+literature. The complete Isabelle code can be found
+online\footnote{\url{www.dcs.shef.ac.uk/$\sim$alasdair/rg}}. A
+previous Isabelle implementation of rely-guarantee reasoning is due to
+Prensa Nieto~\cite{nieto_rely-guarantee_2003}. Our implementation
+differs both by making the link between concrete programs and algebras
+explicit, which increases modularity, and by allowing arbitrary nested parallelism.
 
@@
 abortive action and $1$ as skip. In this context it is essential that
-multiplication is not commutative. Often we use $;$ in place of
-$\cdot$ when discussing programs. More formally, it is well known that
-(regular) languages with language union as $+$ and language product as
-$\cdot$, the empty language as $0$ and the empty word language
-$\{\varepsilon\}$ as $1$ form dioids. Another model is formed by
-binary relations with the union of relations as $+$, the product of
+multiplication is not commutative. Often we use $;$ for sequential
+composition when discussing programs. More formally, it is well known
+that (regular) languages with language union as $+$, language
+product as $\cdot$, the empty language as $0$ and the empty word
+language $\{\varepsilon\}$ as $1$ form dioids. Another model is formed
+by binary relations with the union of relations as $+$, the product of
 relations as $\cdot$, the empty relation as $0$ and the identity
@@
 A \emph{trioid} is a structure $(S,+,\cdot,||,0,1)$ such that
-$(S,+,\cdot,0,1)$ is a dioid and $(S,\cdot,||,0,1)$ a commutative
+$(S,+,\cdot,0,1)$ is a dioid and $(S,+,||,0,1)$ a commutative
 dioid. In a trioid there is no interaction between the sequential
@@
 Formally, the \emph{shuffle} $\|$ of two finite words is defined
-inductively as
-\begin{equation*}
-\epsilon \| s = \{s\}, \qquad s \| \epsilon = \{s\},\qquad
-as \| bt = a(s \| bt) \cup b(as \| t),
-\end{equation*}
-which is then lifted to the shuffle product of languages $X$ and $Y$
-as
-\begin{equation*}
-X \| Y = \{x \| y:\; x \in X \land x \in Y\}.
-\end{equation*}
+inductively as $\epsilon \| s = \{s\}$, $s \| \epsilon = \{s\}$, and
+$as \| bt = a(s \| bt) \cup b(as \| t)$, which is then lifted to the
+shuffle product of languages $X$ and $Y$ as $X \| Y = \{x \| y:\; x
+\in X \land x \in Y\}$.
 
-For logics of programs, notions of iteration or recursion are
-essential. A \emph{Kleene algebra} is a dioid expaned with a star
-operation which satisfies the \emph{left unfold axiom} $1 + x\cdot
-x^*\le x^*$ and the \emph{left} and \emph{right induction axiom} $z+x
-\cdot y \le y\Rightarrow x^*\cdot z \le y$ and $z + y\cdot x \le y
-\Rightarrow z\cdot x^* \le y$. It follows that $1+x\cdot x^* = x^*$
-and that the right unfold axiom $1+x^*\cdot x=x^*$ is derivable as
-well. Thus iteration $x^\ast$ is modelled as the least fixpoint of the
-function $\lambda y. 1+x\cdot y$, which is the same as the least
-fixpoint of $\lambda y.1+ y\cdot x$. A \emph{commutative Kleene
-  algebra} is a Kleene algebra in which multiplication is commutative.
+For programming, notions of iteration are essential. A \emph{Kleene
+  algebra} is a dioid expanded with a star operation which satisfies
+both the \emph{left unfold axiom} $1 + x\cdot x^\star\le x^\star$ and
+\emph{left} and \emph{right induction axioms} $z+x \cdot y \le
+y\Rightarrow x^\star\cdot z \le y$ and $z + y\cdot x \le y \Rightarrow
+z\cdot x^\star \le y$. It follows that $1+x\cdot x^\star = x^\star$
+and that the right unfold axiom $1+x^\star\cdot x\le x^\star$ is
+derivable as well. Thus iteration $x^\ast$ is modelled as the least
+fixpoint of the function $\lambda y. 1+x\cdot y$, which is the same as
+the least fixpoint of $\lambda y.1+ y\cdot x$. A \emph{commutative
+  Kleene algebra} is a Kleene algebra in which multiplication is
+commutative.
 
@@
 A \emph{bi-Kleene algebra} is a structure
-$(K,+,\cdot,||,0,1,\hphantom{}^*,\hphantom{}^{(*)})$ where
-$(K,+,\cdot,0,1,\hphantom{}^*)$ is a Kleene algebra and
-$(K,+,||,0,1,\hphantom{}^{(*)})$ is a commutative Kleene
+$(K,+,\cdot,||,0,1,\hphantom{}^\star,\hphantom{}^{(\star)})$ where
+$(K,+,\cdot,0,1,\hphantom{}^\star)$ is a Kleene algebra and
+$(K,+,||,0,1,\hphantom{}^{(\star)})$ is a commutative Kleene
 algebra. Bi-Kleene algebras are sound and complete with respect to the
 equational theory of regular series-parallel pomset languages, and the
-equational theory is again decidable~\cite{LaurenceStruth}. A
+equational theory is again decidable~\cite{laurence_completeness_2013}. A
 \emph{concurrent Kleene algebra} is a bi-Kleene algebra which
@@
 
-In some contexts, it is also useful to add a meet operation $\sqcap$,
-such that $(D,+,\sqcap)$ is a distributive lattice. This is
+In some contexts, it is also useful to add a meet operation $\sqcap$ to a bi-Kleene algebra,
+such that $(K,+,\sqcap)$ is a distributive lattice. This is
 particularly needed in the context of refinement, where we typically
@@
 
-It is well known that the inference rules for sequential Hoare logic
+It is well known that the inference rules of sequential Hoare logic
 (except the assignment axiom) can be derived in expansions of Kleene
-algebras. One approach is as follows~\cite{moller_algebras_2006}. Suppose a
-suitable Boolean algebra $B$ of \emph{tests} has been embedded into a
-Kleene algebra $K$ such that $0$ and $1$ are the minimal and maximal
-element of $B$, $+$ corresponds to join and $\cdot$ to
-meet. Complements $-$ are defined only on $B$. Suppose further that a
-\emph{backward diamond operator} $\langle x|p$ has been defined for
-each $x\in K$ and $p\in B$, which models the set of all states to
-which each terminating execution of program $x$ may lead from states
-$q$. Finally suppose that a \emph{forward box operator} $|x]p$ has
-been defined which models the (largest) set of states from which every
-terminating execution of $x$ must end up in states $q$ and that boxes
-and diamonds are adjoints of the Galois connection
-\begin{equation*}
-  \langle x|p \le q \Leftrightarrow p \le |x]q,
-\end{equation*}
-for all $x\in K$ and $p,q\in B$. It is then evident from the above
-explanations that validity of a Hoare triple $\vdash \{p\}x\{q\}$ can
-be encoded as $\langle x|p \le q$ and the weakest liberal precodition
-operator $\mathsf{wlp}(x,q)$ as $|x]p$. Hence the relationship between
-the proof theory and the semantics of Hoare logic is captured by the
-Galois connection $ \vdash\{p\}x\{q\} \Leftrightarrow p \le
-\mathsf{wlp}(x,q)$. It has been shown that the relational semantics of
-sequential while-programs can be encoded in these \emph{modal Kleene
-  algebras} and that the inference rules of Hoare logic can be
+algebras. One approach is as
+follows~\cite{moller_algebras_2006}. Suppose a suitable Boolean
+algebra $B$ of \emph{tests} has been embedded into a Kleene algebra
+$K$ such that $0$ and $1$ are the minimal and maximal element of $B$,
+$+$ corresponds to join and $\cdot$ to meet. Complements $-$ are
+defined only on $B$. Suppose further that a \emph{backward diamond
+  operator} $\langle x|p$ has been defined for each $x\in K$ and $p\in
+B$, which models the set of all states to which each terminating
+execution of program $x$ may lead from states $q$. Finally suppose
+that a \emph{forward box operator} $|x]p$ has been defined which
+models the (largest) set of states from which every terminating
+execution of $x$ must end in states $p$ and that boxes and diamonds
+are adjoints of the Galois connection $\langle x|p \le q
+\Leftrightarrow p \le |x]q$, for all $x\in K$ and $p,q\in B$. It is
+then evident from the above explanations that validity of a Hoare
+triple $\vdash \{p\}x\{q\}$ can be encoded as $\langle x|p \le q$ and
+the weakest liberal precondition operator $\mathsf{wlp}(x,q)$ as
+$|x]p$. Hence the relationship between the proof theory and the
+semantics of Hoare logic is captured by the Galois connection $
+\vdash\{p\}x\{q\} \Leftrightarrow p \le \mathsf{wlp}(x,q)$. It has
+been shown that the relational semantics of sequential while-programs
+can be encoded in these \emph{modal Kleene algebras} and that the
+inference rules of Hoare logic can be
 derived~\cite{moller_algebras_2006}.
@@
 In the context of concurrency, this relational approach is no longer
-appropriate. Here, the following approach by
+appropriate; the following approach by
 Tarlecki~\cite{tarlecki_language_1985} can be used instead. One can
@@
 \end{equation*}
-for arbitrary elements of a Kleene algebra $K$. Nevertheless all the
+for arbitrary elements of a Kleene algebra. Nevertheless all the
 rules of sequential Hoare logic except the assignment axiom can still
-be derived in this setting~\cite{hoare_concurrent_2011}. Tarlecki's motivating
+be derived~\cite{hoare_concurrent_2011}. Tarlecki's motivating
 explanations carry over to the algebraic approach.
@@
 
-The correspondence to a wlp-style semantics, as in model Kleene
+The correspondence to a wlp-style semantics, as in modal Kleene
 algebra, now requires a generalisation of the Galois connection for
@@
 \end{equation*}
-These residuals, and now even the Kleene star can be axiomatised
+These residuals, and now even the Kleene star, can be axiomatised
 equationally in action algebras. For a comprehensive list of the
@@
 see~\cite{armstrong_kleene_2013}, including the language and the relational
-model. In analogy to the development in model Kleene algebra we can
+model. In analogy to the development in modal Kleene algebra we can
 now stipulate $ \wlp(x,y)=y\leftarrow x$ and obtain the Galois
@@
 \end{equation*}
-with the characteristic properties $\vdash \{\wlp(y,z)\}y\{z\}$ and $x\le
-\wlp(y,z)\Rightarrow \ \vdash \{x\}y\{z\}$. Moreover, if the action algebra is
-also a quantale, and infinite sums exist, it follows that $
-\wlp(y,z)=\sum\{x:\ \vdash \{x\}y\{z\}\}$.  It is obvious that this definition
-makes sense in all models of action algebras and
+with $\vdash \{\wlp(y,z)\}y\{z\}$ and $x\le \wlp(y,z)\Rightarrow \
+\vdash \{x\}y\{z\}$ as characteristic properties. Moreover, if the
+action algebra is also a quantale, and infinite sums exist, it follows
+that $ \wlp(y,z)=\sum\{x:\ \vdash \{x\}y\{z\}\}$.  It is obvious that
+this definition makes sense in all models of action algebras and
 quantales. Intuitively, suppose $p$ stands for the set of all
@@
 in $q$.
-
-In the context of concurrent Kleene algebra, even without the parallel
-star, additional inference rules can be derived, in particular
-\begin{equation*}
-  \frac{\vdash\{x\}y\{z\}}{\vdash\{w||x\}y\{w||z\}}\qquad \frac{\vdash\{x_1\}y_1\{z_1\}\quad\vdash\{x_2\}y_2\{z_2\}}{\vdash\{x_1||x_2\}y_1||y_2\{z_1||z_2\}}
-\end{equation*}
-which correspond to the frame rule and the concurrency rule of
-concurrent separation logic~\cite{} when $||$ is interpreted as
-separating conjunction. In fact, the second inference rule is, in any
-bi-Kleene algebra equivalent to the interchange axiom and the frame
-rule is derivable from this axiom as well~\cite{hoare_concurrent_2011}. The
-precise connection between concurrent separation logic, a variant of
-concurrent Kleene algebra and a resource transformer model has been
-elaborated in~\cite{hoare_locality_2011}.
 
@@
 
-In this section we show how bi-Kleene algebras can be expanded into a
-simple algebra that supports the derivation of rely-guarantee-style
+We now show how bi-Kleene algebras can be expanded into a
+simple algebra that supports the derivation of rely-guarantee style
 inference rules. This development does \emph{not} use the interchange
 law for several reasons. First, this law fails for fair parallel
-composition $x\parallel_f y$ in a model with possibly infinite, or
+composition $x\parallel_f y$ in models with possibly infinite, or
 non-terminating programs. In this model, $x \cdot y \not\leq x
-\parallel_f y$ in the case whenever $x$ is non-terminating. Secondly,
-it is not needed for deriving the usual laws or rely-guarantee.
+\parallel_f y$ whenever $x$ is non-terminating. Secondly, it is not
+needed for deriving the usual rules of rely-guarantee.
 
 A rely-guarantee algebra is a structure
-$(K,I,+,\sqcap,\cdot,\|,^\star,0,1)$, where $(K,+,\sqcap)$ is a
-distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
-$(K,+,\cdot,^\star,0,1)$ is a Kleene algebra. Note that we do not
-consider the parallel star. $I$ is a distinguished subset of rely and
-guarantee conditions or \emph{interference constraints} which satisfy
-the following axioms
+$(K,I,+,\sqcap,\cdot,\|,\hphantom{}^\star,0,1)$, where $(K,+,\sqcap)$
+is a distributive lattice, $(K,+,\cdot,\|,0,1)$ is a trioid and
+$(K,+,\sqcap,\cdot,\|,\hphantom{}^\star,0,1)$ is a bi-Kleene algebra
+where we do not consider the parallel star. $I$ is a distinguished
+subset of rely and guarantee conditions or \emph{interference
+  constraints} which satisfy the following axioms
 \begin{align}
@@
 r &\le r\|r', \label{rg2}\\
-r\|xy &= (r\|x)(r\|y), \label{rg3}\\
+r\|(x\cdot y) &= (r\|x)\cdot(r\|y), \label{rg3}\\
 r\|x^+ &\le (r\|x)^+ \label{rg4}.
@@
 depending on whether they are used as relies or guarantees, and
-$x,y,z$ for arbitrary elements of $K$. The operators $\|$ and $\sqcap$
+$x,y,z$ for arbitrary elements of $K$. The operations $\|$ and $\sqcap$
 must be closed with respect to $I$.
 
-Axiom (\ref{rg1}) states that interference from a constraint being run
-twice in parallel is no different from just the interference from that
-constraint begin run once in parallel. Axiom (\ref{rg2}) states that
-interference from a single constraint is less than interference from
-itself and another interference constraint. Axiom (\ref{rg3}) allows
-an interference constraint to be split across sequential
-programs. Axiom (\ref{rg4}) is similar to Axiom (\ref{rg3}) in intent,
-except it deals with finite iteration.
+The general idea is to constrain a program by a rely condition by
+executing the two in parallel. Axiom (\ref{rg1}) states that
+interference from a constraint being run twice in parallel is no
+different from just the interference from that constraint begin run
+once in parallel. Axiom (\ref{rg2}) states that interference from a
+single constraint is less than interference from itself and another
+interference constraint. Axiom (\ref{rg3}) allows an interference
+constraint to be split across sequential programs. Axiom (\ref{rg4})
+is similar to Axiom (\ref{rg3}) in intent, except it deals with finite
+iteration.
 
@@
 1 \le r,\qquad
-r^\star = rr = r = r\|r,\qquad
+r^\star = r\cdot r = r = r\|r,\qquad
 r\|x^+ = (r\|x)^+.
@@
 \begin{proof}
-  To show this, we use Isabelle's
+  We have used Isabelle's
   \emph{Nitpick}~\cite{blanchette_nitpick:_2010} counterexample
   generator to construct models which violate each particular axiom
-  while satisfying all others.
+  while satisfying all others. \qed
 \end{proof}
@@
 \begin{thm}
-Axiom (\ref{rg4}) can be derived from (\ref{rg3}) by a simple
-inductive argument in a quantale where $\|$ distributes over arbitrary suprema.
+  Axiom (\ref{rg3}) implies (\ref{rg4}) in a quantale
+  where $\|$ distributes over arbitrary suprema.
 \end{thm}
@@
   $x^+=\sum_{i\ge 1} x^i$ where $x^1=x$ and $x^{i+1}=x\cdot x^i$. By
-  induction on w$i$ we get $r\|x^i = (r\|x)^i$, hence
+  induction on $i$ we get $r\|x^i = (r\|x)^i$, hence
   \begin{align*}
@@
   \end{align*}
+  \qed
 \end{proof}
@@
 
-Jones quintuples can be encodeded in this setting as
+Jones quintuples can be encoded in this setting as
 \begin{align}
-r, g \vdash \{p\} c \{q\} \iff p(r\|c) \le q \land c \le g. \label{quin}
+r, g \vdash \{p\} x \{q\} \iff p\cdot(r\|x) \le q \land x \le g. \label{quin}
 \end{align}
+This means that program $x$ when constrained by a rely $r$, and
+executed after $p$, behaves as $q$. Moreover, all behaviours of $x$
+are included in its guarantee.
 
 \begin{thm}
-With this encoding we can derive the standard rely-guarantee inference
-rules, as shown in Figure \ref{fig:rgrules}.
+  The standard rely-guarantee inference rules can be derived with the
+  above encoding, as shown in Figure \ref{fig:rgrules}.
 \end{thm}
@@
 \begin{prooftree}
-\RightLabel{Skip}
-\AxiomC{$pr \le p$}
+\RightLabel{\ Skip}
+\AxiomC{$p\cdot r \le p$}
 \UnaryInfC{$r, g \vdash \{p\}1\{p\}$}
@@
 \begin{prooftree}
-\RightLabel{Weakening}
+\RightLabel{\ Weakening}
 \AxiomC{$r' \le r$}
@@
 \begin{prooftree}
-\RightLabel{Sequential}
+\RightLabel{\ Sequential}
 \AxiomC{$r, g \vdash \{p\}x\{q\}$}
 \AxiomC{$r, g \vdash \{q\}y\{s\}$}
-\BinaryInfC{$r, g \vdash \{q\}xy\{s\}$}
+\BinaryInfC{$r, g \vdash \{q\}x\cdot y\{s\}$}
 \end{prooftree}
@@
 \begin{prooftree}
-\RightLabel{Parallel}
+\RightLabel{\ Parallel}
 \AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
@@
 \begin{prooftree}
-\RightLabel{Choice}
+\RightLabel{\ Choice}
 \AxiomC{$r, g \vdash \{p\}x\{q\}$}
@@
 \begin{prooftree}
-\RightLabel{Star}
-\AxiomC{$pr \le p$}
+\RightLabel{\ Star}
+\AxiomC{$p\cdot r \le p$}
 \AxiomC{$r, g \vdash \{p\}x\{p\}$}
@@
 
-Thus equations
-(\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
-rules, somehow represent a minimalistic set of axioms from which these
+Thus (\ref{rg1}) to (\ref{rg4}), which are all necessary to derive these
+rules, represent a minimal set of axioms from which these
 inference rules can be derived.
@@
 
-While the algebra in section \ref{sec:RG} is adequate for deriving the
-standard inference rules, it's equality is too strong to capture many
+While the algebra in the previous section is adequate for deriving the
+standard inference rules, its equality is too strong to capture many
 interesting statements about concurrent programs. Consider the
@@
 This can be read as follows; if $x$ and $y$ are equal, then they must
-be equal under all possible interference from any arbitrary $z$. At
-first sight, this might seem to preclude any reasoning about
-interference in a fine-grained way using purely algebra. This is not
-the case, but breaking such inherent compositionality in just the
-right way to capture interesting properties of interference requires
-extra work.
+be equal under all possible interferences from an arbitrary $z$. At
+first, this might seem to preclude any fine-grained reasoning about
+interference using purely algebra. This is not the case, but breaking
+such inherent compositionality in just the right way to capture
+interesting properties of interference requires extra work.
 
-The most obvious way we can acheive this is by expanding our
-rely-guarantee algebra with an additional function $\pi$ and redefining
-our quintuples as,
+The obvious way of achieving this is to expand our rely-guarantee
+algebra with an additional function $\pi : K \to K$ and redefining our
+quintuples as,
 \begin{align*}
-r, g \vdash \{p\} x \{q\} \iff p(r\|c) \le_\pi q \land x \le g.
+r, g \vdash \{p\} x \{q\} \iff p\cdot (r\|c) \le_\pi q \land x \le g.
 \end{align*}
 Where $x \le_\pi y$ is $\pi(x) \le \pi(y)$. Since for any operator
-$\bullet$, it is not required that
+$\bullet$ it is not required that
 \begin{align*}
@@
 \end{align*}
-we can `break' compositionality in just the right way, provided we
-choose the correct properties for $\pi$. However, if we simply listed
-all properties that $\pi$ needs to satisfy we would end up with an
-overly long list of axioms to add to our algebra. Instead we define
-the function $\pi$ as $\lambda x.\; x \sqcap c$ where $c$ is the
-greatest $consistent$ element such that
+we can break compositionality in just the right way, provided we chose
+appropriate properties for $\pi$. These properties are extracted from
+properties of the trace model, which will be explained in detail in
+the next section. Many of those can be derived from the fact that, in
+our model, $\pi = \lambda x.\; x \sqcap c$, where $c$ is healthiness
+condition which filters out ill-defined traces. We do not list these
+properties here. In addition $\pi$ must satisfy the properties
 \begin{align}
-x^* &\le_\pi \pi(x)^*, \label{con1}\\
-xy &\le_\pi \pi(x)\pi(y). \label{con2}\\
-z + x\cdot y \le_\pi y &\implies x^*\cdot z \le_\pi y\label{con3}\\
-z + y\cdot x \le_\pi y &\implies z\cdot x^* \le_\pi y\label{con4}
+x^\star &\le_\pi \pi(x)^\star, \label{con1}\\
+x\cdot y &\le_\pi \pi(x)\cdot \pi(y), \label{con2}\\
+z + x\cdot y \le_\pi y &\implies x^\star\cdot z \le_\pi y\label{con3},\\
+z + y\cdot x \le_\pi y &\implies z\cdot x^\star \le_\pi y\label{con4}.
 \end{align}
@@
 \begin{thm}
-Axioms (\ref{con1}) to (\ref{con4}) induce a Kleene subalgebra.
-\begin{align*}
-(\{x\in K.\; \pi(x) = x\}, +_\pi, \cdot_\pi, \hphantom{}^\pi,0,1)
-\end{align*}
+$(\pi(K), +_\pi, \cdot_\pi, \hphantom{}^\pi,0,1)$ is a Kleene algebra.
 \end{thm}
-As such we redefine our rely-guarantee algebra from section
-\ref{sec:RG} as a structure $(K,RG,+,\sqcap,\cdot,\|,\hphantom{}^\star,0,1,c)$
-which, in addition to the rules in section \ref{sec:RG} satisfies both
-(\ref{con1}) to (\ref{con4}).
+\begin{proof}
+  It can be shown that $\pi$ is a retraction, that is, $\pi^2 =
+  \pi$. Therefore, $x\in \pi(K)$ iff $\pi(x) = x$. This condition can
+  then be used to check the closure conditions for all operations. \qed
+\end{proof}
+
+We redefine our rely-guarantee algebra as a structure
+$(K,I,+,\sqcap,\cdot,\|,\hphantom{}^\star,\pi,0,1)$ which, in addition
+to the rules in Section \ref{sec:RG}, satisfies (\ref{con1}) to
+(\ref{con4}).
 \begin{thm}
-All the rules in Figure \ref{fig:rgrules} can be derived with this
-algebra, and moreover their proofs remain the same, mutatis
-mutandis.
+  All rules in Figure \ref{fig:rgrules} can be derived in this
+  algebra.
 \end{thm}
+Moreover their proofs remain the same, mutatis mutandis.
 
@@
 
-In this section, we construct a finite language model satisfying the
-axioms in Section \ref{sec:RG}. Restricting our attention to finite
-languages means we do not need to concern ourselves with non-algebraic
+We now construct a finite language model satisfying the axioms in
+Section \ref{sec:RG} and \ref{sec:INT}. Restricting our attention to
+finite languages means we do not need to concern ourselves with
 termination side-conditions, nor do we need to worry about additional
 restrictions on parallel composition, e.g. fairness. However, all the
-results in this section can easily be adapted to potentially
-infinite languages, and our Isabelle/HOL formalisation includes these more
-general definitions by using coinductively defined lazy lists to
-represent words, and having a weakly-fair shuffle operator for such
-infinite languages.
+results in this section can be adapted to potentially infinite
+languages, and our Isabelle/HOL formalisation already includes general
+definitions by using coinductively defined lazy lists to represent
+words, and having a weakly-fair shuffle operator for such infinite
+languages.
 
 We consider languages where the alphabet contains state pairs of the
-form $(\sigma_1,\sigma_2) \in \Sigma \times \Sigma$. A word in such a
-language is \emph{consistent} if every such pair in a word has the
-same first state as the previous transition's second state. For
-example, $(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is consistent, while
+form $(\sigma_1,\sigma_2) \in \Sigma^2$. A word in such a language is
+\emph{consistent} if every such pair in a word has the same first
+state as the previous transition's second state. For example,
+$(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is consistent, while
 $(\sigma_1,\sigma_2)(\sigma_3,\sigma_3)$ is inconsistent. Sets of
@@
   traces}~\cite{boer_formal_1999}, but lack the usual process
-labels. We denote the set of all consistent words by $T$.
+labels. We denote the set of all consistent words by $C$ and define
+the function $\pi$ from the previous section as $\lambda X.\; X \cap
+C$ in our model.
 
@@
 usual. Concurrent composition is the shuffle product defined in
-section \ref{sec:KA}. The shuffle product is associative, commutative,
+Section \ref{sec:KA}. The shuffle product is associative, commutative,
 and distributes over arbitrary joins. Both products share the same
-unit, $\{\epsilon\}$ and zero, $\emptyset$. Proving properties of
-shuffle is suprisingly tricky (especially if one considers infinite
-words). For a in-depth treatment of the shuffle product
-see~\cite{mateescu_shuffle-like_1997}.
+unit, $\{\epsilon\}$ and zero, $\emptyset$. In Isabelle proving
+properties of shuffle is surprisingly tricky (especially if one
+considers infinite words). For a in-depth treatment of the shuffle
+product see~\cite{mateescu_shuffle-like_1997}.
 
-\begin{lem}
-$(\mathcal{P}((\Sigma\times\Sigma)^*), \cup, \cdot,\sha, \emptyset,
+\begin{thm}
+$(\mathcal{P}((\Sigma^2)^\star), \cup, \cdot,\|, \emptyset,
   \{\epsilon\})$ forms a trioid.
-\end{lem}
+\end{thm}
 
@@
 words which can be built from some set of state pairs in
-$\Sigma\times\Sigma$. We define a function $\langle R\rangle$ which
+$\Sigma^2$. We define a function $\langle R\rangle$ which
 lifts a relation $R$ to a language containing words of length one for
-each pair in $R$. The set of rely-guaratee conditions $C$ is then defined
-as $\{r.\; \exists R. r = \langle R\rangle^*\}$.
+each pair in $R$. The set of rely-guarantee conditions $I$ is then defined
+as $\{r.\; \exists R. r = \langle R\rangle^\star\}$.
 
-\begin{lem}
-$(\mathcal{P}((\Sigma\times\Sigma)^*), C, \cup, \cdot,\sha,
-  \emptyset, \{\epsilon\}, T)$ is a rely-guarantee algebra.
-\end{lem}
+\begin{thm}
+$(\mathcal{P}((\Sigma^2)^\star), I, \cup, \cdot,\|,
+  \hphantom{}^\star,\pi,\emptyset, \{\epsilon\})$ is a rely-guarantee algebra.
+\end{thm}
 Since $\langle R\rangle$ is atomic, it satisfies several useful properties, such as,
 \begin{align*}
-\langle R\rangle^\star \| \langle S\rangle &= \langle R\rangle^\star; \langle S\rangle; \langle R\rangle^\star\\
-\langle R\rangle^\star \| \langle S\rangle^\star &= (\langle R\rangle^\star; \langle S\rangle^\star)^\star
+\langle R\rangle^\star \| \langle S\rangle = \langle R\rangle^\star; \langle S\rangle; \langle R\rangle^\star,
+\qquad \langle R\rangle^\star \| \langle S\rangle^\star = (\langle R\rangle^\star; \langle S\rangle^\star)^\star.
 \end{align*}
@@
 consistent words were constructed from previously inconsistent
-words---the shuffle operator can take two inconsistent words and
-shuffle them together in such a way to generate many consistent
-words. If we only considered consistent words, \'a la Aczel traces, we
-would be unable to define such a shuffle operator directly on the
-traces themselves, and would instead have to rely on some operational
-semantics to generate traces.
+words---the shuffle operator can generate many consistent words from
+two inconsistent words. If we only considered consistent words, \`a la
+Aczel traces, we would be unable to define such a shuffle operator
+directly on the traces themselves, and would instead have to rely on
+some operational semantics to generate traces.
 
-Following~\cite{brookes_full_1993} and~\cite{dingel_refinement_2002},
-we inductively generate the mumble language $w^\dagger$ for a word $w$
-in a language over $\Sigma^2$ as follows: Assume $R,S \in \Sigma^2$
-and $u,v,w \in (\Sigma^2)^*$. First, $w \in w^\dagger$. Secondly, if
-$uRSv \in w^\dagger$ then $u(R;S)v \in w^\dagger$. This operation is
-lifted to languages in the obvious way as
+\section{Enriching the Model}
+
+To model and verify programs we need additional concepts such as tests
+and assignment axioms. A \emph{test} is any language $P$ where $P \le
+\langle \Id\rangle$. We write $\test(P)$ for $\langle
+\Id_P\rangle$. In Kleene algebra the sequential composition of two
+tests should be equal to their intersection. However, the traces
+$\test(P); \test(Q)$ and $\test(P \cap Q)$ are incomparable, as all
+words in the former have length two, while all the words in the latter
+have length one. To overcome this problem, we use the concepts of
+\emph{stuttering} and \emph{mumbling},
+following~\cite{brookes_full_1993}
+and~\cite{dingel_refinement_2002}. We inductively generate the
+\emph{mumble language} $w^\dagger$ for a word $w$ in a language over
+$\Sigma^2$ as follows: Assume $\sigma_1,\sigma_2,\sigma_3 \in \Sigma$
+and $u,v,w \in (\Sigma^2)^\star$. First, $w \in w^\dagger$. Secondly,
+if $u(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)v \in w^\dagger$ then
+$u(\sigma_1,\sigma_3)v \in w^\dagger$. This operation is lifted to
+languages in the obvious way as
 \begin{align*}
@@
 
-Stuttering and mumbling are important for defining \emph{tests}. A
-test is any language $P$ where $P \le \langle \Id\rangle$. We
-write $\test(P)$ for $\langle \Id_P\rangle$. Without any
-stuttering and mumbling the traces $\test(P); \test(Q)$ and
-$\test(P \cap Q)$ are incomparable, as all words in the former have
-length two, while all the words in the latter have length one. With mumbling we have that
+With mumbling we now have that
 \begin{align*}
@@
 as the longer words in $\test(P);\test(Q)$ can be mumbled down into
-the shorter words of $\test(P\cap Q)$, whereas with stuttering we
-have the opposite,
+the shorter words of $\test(P\cap Q)$, whereas stuttering gives us the opposite direction,
 \begin{align*}
-\langle I\rangle^\star \| (\test(P);\test(Q)) \le_\pi \langle I\rangle^\star \| \test(P \cap Q).
+\langle \Id\rangle^\star \| (\test(P);\test(Q)) \le_\pi \langle \Id\rangle^\star \| \test(P \cap Q).
 \end{align*}
-For the rest of the paper, we assume that all languages are
+We henceforth assume that all languages are
 implicitly mumble closed.
 
-Using tests, we can encoding common program statements such as if
-statements and while loops as in Kozen's Kleene algebra with tests.
+Using tests, we can encode if statements and while loops
 \begin{align*}
-\mathsf{if } P \text{ \{ } X \text { \} \textsf{else} \{ } Y \text{ \}} &= \test(P); X + \test(- P); Y\\
-\mathsf{while } P \text{ \{ } X \text{ \}} &= (\test(P);X)^\star;\test(- P)
+\mathsf{if }\;\; P \text{ \{ } X \text { \} \textsf{else} \{ } Y \text{ \}} &= \test(P); X + \test(- P); Y,\\
+\mathsf{while}\;\; P \text{ \{ } X \text{ \}} &= (\test(P);X)^\star;\test(- P).
 \end{align*}
-Finally, we define the operator $\edn(P)$ which is contains all the
+Next, we define the operator $\edn(P)$ which contains all the
 words which end in a state satisfying $P$. Some useful properties of $\edn$ include
+\begin{gather*}
+\edn(P); \test{Q} \le_\pi \edn(P \cap Q), \qquad
+\test(P) \le \edn(Q),\\
+\text{range}(\Id_P \circ R) \le P \implies \edn(P); \langle R\rangle^\star \le_\pi \edn(P).
+\end{gather*}
+
+In this model, assignment is defined as
 \begin{align*}
-\edn(P); \test{Q} &\le_\pi \edn(P \cap Q),\\
-\test(P) &\le \edn(Q),\\
-\text{range}(\Id_P \circ R) \le P \implies \edn(P); \langle R\rangle^\star &\le_\pi \edn(P).
-\end{align*}
-
-In the trace model, assignment is defined as
-\begin{align*}
-  x := e = \bigcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
+  x := e \ =\ \bigcup v.\,\test\{\sigma.\,\eval(\sigma,e) = v\} \cdot x \leftarrow v
 \end{align*}
@@
 value $v$ to $x$. The $\eval$ function evaluates an expression $e$ in
-the state $\sigma$. Using this definition of assignment we can derive
-the assignment rule
+the state $\sigma$. Using this definition we derive the assignment
+rule
 \begin{align*}
@@
 &\unchanged(- \{x\})\\
-&\vdash \{\edn(P)\} x := e \{\edn(P[x/e])\}.
+&\vdash \{\edn(P)\}\ x := e\ \{\edn(P[x/e])\}.
 \end{align*}
@@
 allowed to modify any of the variables used when evaluating $e$,
-i.e. those variables must remain unchanged. Second, that the
-environment must preserve the precondition. Third, that postcondition
+i.e. those variables must remain unchanged. Second, the
+environment must preserve the precondition. Third, the postcondition
 of the assignment statement is also preserved. In turn, the assignment
@@
 which are defined much like unchanged except they only require that
-variables increase or decrease, rather than stay the same. Note that
-we can easily define other useful assignment rules---if we know things
-about $P$ and $e$, we can make stronger guarantees about what $x := e$
-can do. For example the assignment $x := x - 2$ can also guarantee
-that $x$ will always decrease.
+variables increase or decrease, rather than stay the same. We can
+easily define other useful assignment rules---if we know properties about
+$P$ and $e$, we can make stronger guarantees about what $x := e$ can
+do. For example the assignment $x := x - 2$ can also guarantee that
+$x$ will always decrease.
 
@@
 \section{Examples}
+\label{sec:Examples}
 
-\fbox{Simple assignment example}
+To demonstrate how the parallel rule behaves, consider the following
+simple statement, which simply assigns two variables in parallel:
+\begin{align*}
+&\langle\Id\rangle^\star, \langle\top\rangle^\star \vdash \{\edn(x = 2 \land y = 2 \land z = 5)\}\\
+&\qquad x := x + 2 \;\|\; y := z\\
+&\{\edn (x = 4 \land y = 5 \land z = 5)\}.
+\end{align*}
+The environment $\langle\Id\rangle^\star$ is only giving us stuttering
+interference. Since we are considering this program in isolation, we
+make no guarantees about how this affects the environment. To apply
+the parallel rule from Figure \ref{fig:rgrules}, we weaken or
+strengthen the interference constrains and pre/postcondition as needed
+to fit the form of the parallel rule.
 
-Figure \ref{fig:findp} shows the classic FINDP program, which has been
-used by numerous authors including Owicki~\cite{owicki_axiomatic_1975},
-Jones~\cite{jones_development_1981}, de Roever~\cite{de_roever_concurrency_2001} and
-Hayes~\cite{hayes_refining_2013}. The program finds the least element of an array
-satisfying the predicate $P$. The index of the first element
-satisfying $p$ is placed in the variable $f$. If there are no elements
-of the array satisfying $P$ then $f$ will be set to the length of the
-array. The program has two subprograms, $A$ and $B$, which run in
-parallel, one of which searches the even indices while the other
-searches the odd indices. A speedup over a sequential implementation
-is acheived as $A$ will terminate when $B$ finds an element of the
-array satisfying $P$ which is less than $i_A$.
+First, we weaken the rely condition to $\unchanged\{x\}\ \sqcap\ \unchanged\{y,z\}$. Second we strengthen the guarantee condition to
+$\unchanged\{y,z\}\ \|\ \unchanged\{x\}$. When we apply the parallel
+rule each assignment's rely will become the other assignment's
+guarantee. Finally, we split the precondition and postcondition into
+$\edn(x = 2)\ \sqcap\ \edn(y = 2 \land z = 5)$ and $\edn(x =
+4)\ \sqcap\ \edn(y = 5 \land z = 5)$ respectively. Upon applying the
+parallel rule, we obtain two trivial goals
+\begin{align*}
+&\langle\unchanged\{x\}\rangle^\star, \langle\unchanged\{y,z\}\rangle^\star \vdash \{\edn(x = 2)\}\;x := x + 2 \;\{\edn (x = 4)\},\\
+&\langle\unchanged\{y,z\}\rangle^\star, \langle\unchanged\{x\}\rangle^\star \vdash \{\edn(y = 2 \land z = 5)\}\\
+&\qquad y := z \\
+&\{\edn (y = 5 \land z = 5)\}.
+\end{align*}
+
+Figure \ref{fig:findp} shows the FINDP program, which has been
+used by numerous authors
+e.g.~\cite{owicki_axiomatic_1975,jones_development_1981,de_roever_concurrency_2001,hayes_refining_2013}. The
+program finds the least element of an array satisfying a predicate
+$P$. The index of the first element satisfying $p$ is placed in the
+variable $f$. If no element of the array satisfies $P$
+then $f$ will be set to the length of the array. The program has two
+subprograms, $A$ and $B$, running in parallel, one of which searches
+the even indices while the other searches the odd indices. A speedup
+over a sequential implementation is achieved as $A$ will terminate
+when $B$ finds an element of the array satisfying $P$ which is less
+than $i_A$.
 \begin{figure}
@@
 
-Here, we sketch the main outline of the correctness proof, and comment
-on its implementation in Isabelle. We don't attempt to give a fully
-detailed proof, as this has been done many times previously.
+Here, we only sketch the correctness proof, and comment on its
+implementation in Isabelle. We do not attempt to give a detailed
+proof, as this has been done many times previously.
 
@@
 \begin{align*}
-\text{FINDP} \le_\pi \edn(\mathsf{leastP}(f)) + \edn(f = \mathsf{len}(\text{array})).
+\text{FINDP} \le_\pi \edn(\mathsf{leastP}(f)) + \edn(f = \mathsf{len}(\text{array})),
 \end{align*}
@@
 
-Despite the seemingly straightforward nature of this proof, this turns
-out to be suprising difficult to prove in Isabelle. Each atomic step
+Despite the seemingly straightforward nature of this proof, it turns
+out to be surprisingly difficult in Isabelle. Each atomic step
 needs to be shown to satisfy the guarantee of its containing
-subprogram, as well as any goals relating to it's pre and post
-conditions. This invariably leads to a proliferation of many numerous
-proof goals, even for such a small program. More work needs to be done
+subprogram, as well as any goals relating to its pre and post
+conditions. This invariably leads to a proliferation of many small
+proof goals, even for such a simple program. More work must be done
 to manage the complexity of such proofs within interactive theorem
-provers. Existing work on combining rely-guarantee with separation
-logic~\cite{vafeiadis_modular_2008} may proof useful here. Our language model is
-sufficiently generic such that arbitrary models of stores may be used,
-including those common in separation logic, which have already been
-implemented in Isabelle.
+provers.
 
@@
 
-We have...
+We have introduced variants of semirings and Kleene algebras intended
+to model rely-guarantee and interference based reasoning. We have
+developed an interleaving model for these algebras which uses familiar
+concepts from traces and language theory. This theory has been
+implemented in the Isabelle/HOL theorem prover, providing a solid
+mathematical basis on which to build a tool for mechanised refinement
+and verification tasks. In line with this aim, we have applied
+our formalisation to two simple examples.
 
-\paragraph*{Acknowledgements} This work was supported by funding from
-CNPq Brazil and the EPSRC.
+This implementation serves as a basis from which further interesting
+aspects of concurrent programs, such as non-termination and
+synchronisation can be explored. As mentioned in Section
+\ref{sec:Model}, some of the work needed to implement this we have
+already done in Isabelle.
 
-\newpage
+Algebra plays an important role in our development. First, it
+allowed us to derive inference rules rapidly and with little proof
+effort. Second, it yields an abstract layer at which many properties
+that would be difficult to prove in concrete models can be verified
+with relative ease by equational reasoning. Third, as pointed out in
+Section \ref{sec:KA}, some fragments of the algebras considered are
+decidable. Therefore, decision procedures for some aspects of
+rely-guarantee reasoning can be implemented in interactive theorem
+proving tools such as Isabelle.
+
+The examples from Section \ref{sec:Examples} confirm previous
+evidence~\cite{nieto_rely-guarantee_2003} that even seemingly
+straightforward concurrency verification tasks can be tedious and complex. It is
+too early to draw informed conclusions, but while part of this complexity
+may be unavoidable, more advanced models and proof automation are
+needed to overcome such difficulties. Existing work on combining
+rely-guarantee with separation logic~\cite{vafeiadis_modular_2008} may
+prove useful here. Our language model is sufficiently generic such
+that arbitrary models of stores may be used, including those common in
+separation logic, which have already been implemented in Isabelle~\cite{klein_mechanised_2012}.
+
+In addition, algebraic approaches to separation logic have already
+been introduced. Examples are the separation algebras
+in~\cite{calcagno_local_2007}, and algebraic separation
+logic~\cite{dang_algebraic_2011}. More recently, concurrent Kleene
+algebras have given an algebraic account of some aspects of concurrent
+separation logic~\cite{hoare_concurrent_2011,hoare_locality_2011}.
+
+\paragraph*{Acknowledgements.} The authors would like to thank Brijesh
+Dongol and Ian Hayes for inspiring discussions on concurrency
+verification and the rely-guarantee method. The first author
+acknowledges funding from an EPSRC doctoral fellowship. The second
+author is supported by CNPq Brazil. The third
+author acknowledges funding by EPSRC grant EP/J003727/1.
 

### 2014-03-04 Paper/paper.tex
Added Aczel traces with process labels
Added environments

@@
 \usepackage{marginnote}
+\usepackage{hyperref}
 
@@
   control and data flow of concurrent programs with shared variables
-  at different levels of abstraction. This is illustrated on two
-  simple verification examples.
+  at different levels of abstraction. This is illustrated on a simple
+  verification example.
 \end{abstract}
@@
 this model, we demonstrate and evaluate our prototypical verification
-tool implemented in Isabelle by verifying two simple examples from
+tool implemented in Isabelle by verifying a simple example from
 the literature. Beyond that our approach provides a coherent framework
@@
 prototype for the verification and construction of concurrent
-programs. We illustrate the tool with two simple examples from the
+programs. We illustrate the tool with a simple example from the
 literature. The complete Isabelle code can be found
-online\footnote{\url{www.dcs.shef.ac.uk/$\sim$alasdair/rg}}. A
+online\footnote{\url{www.dcs.shef.ac.uk/~alasdair/rg}}. A
 previous Isabelle implementation of rely-guarantee reasoning is due to
@@
 
+A (unital) \emph{quantale} is a dioid based on a complete lattice
+where the multiplication distributes over arbitrary suprema. Formally, it is a
+structure $(S,\le,\cdot,1)$ such that $(S,\le)$ is a complete lattice,
+$(S,\cdot,1)$ is a monoid and
+\begin{align*}
+x(\Sigma Y) = \Sigma\{xy|y\in Y\}, \qquad
+(\Sigma X)y = \Sigma\{xy|x\in X\}.
+\end{align*}
+In a quantale the star is the sum of all powers $x^n$. Therefore, all
+quantales are also Kleene algebras.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 B$, which models the set of all states to which each terminating
-execution of program $x$ may lead from states $q$. Finally suppose
+execution of program $x$ may lead from states $p$. Finally suppose
 that a \emph{forward box operator} $|x]p$ has been defined which
@@
 x$ by the right induction axiom of Kleene algebra and therefore
-$x\cdot (t\cdot y)^\ast t'\le x\cdot t'$ for arbitrary element $t'$ by
+$x\cdot (t\cdot y)^\ast\cdot t'\le x\cdot t'$ for arbitrary element $t'$ by
 isotonicity of multiplication. This derives the while rule
 \begin{equation*}
-  \frac{\vdash \{x\cdot t\}y\{x\}}{\vdash \{x\}(t\cdot y)^\ast\{t'\cdot x\}}
+  \frac{\vdash \{x\cdot t\}y\{x\}}{\vdash \{x\}(t\cdot y)^\ast\cdot t'\{x\cdot t'\}}
 \end{equation*}
@@
 executed after $p$, behaves as $q$. Moreover, all behaviours of $x$
-are included in its guarantee.
+are included in its guarantee $q$. Note that this encoding is stronger
+than in traditional rely-guarantee, as $x$ is required to
+unconditionally implement $g$. The algebra could easily be extended
+with an additional operator $f$ such that $f(r,x) \le q$ would encode
+that $x$ implements $q$ only under interference of at most $r$. For
+more complex examples than what we present in section
+\ref{sec:Examples} such an encoding may prove neccessary.
 
@@
 
-\begin{figure}[hbt]
+\begin{figure}[bth]
 \centering
@@
 \RightLabel{\ Parallel}
-\AxiomC{$r_1, g_2 \vdash \{p_1\}x\{q_1\}$}
+\AxiomC{$r_1, g_1 \vdash \{p_1\}x\{q_1\}$}
 \AxiomC{$g_1 \le r_2$}
-\AxiomC{$r_1, g_2 \vdash \{p_2\}y\{q_2\}$}
+\AxiomC{$r_2, g_2 \vdash \{p_2\}y\{q_2\}$}
 \AxiomC{$g_2 \le r_1$}
-\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap q_2\}x\|y\{q_1 \sqcap q_2\}$}
+\QuaternaryInfC{$r_1 \sqcap r_2, g_1 \| g_2 \vdash \{p_1 \sqcap p_2\}x\|y\{q_1 \sqcap q_2\}$}
 \end{prooftree}
@@
 \begin{align}
-r, g \vdash \{p\} c \{q\} \iff c \le r/(p \rightarrow q) \sqcap g \label{refine}.
+r, g \vdash \{p\} x \{q\} \iff x \le r/(p \rightarrow q) \sqcap g \label{refine}.
 \end{align}
@@
 
-The obvious way of achieving this is to expand our rely-guarantee
+A way of achieving this is to expand our rely-guarantee
 algebra with an additional function $\pi : K \to K$ and redefining our
@@
 $(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is consistent, while
-$(\sigma_1,\sigma_2)(\sigma_3,\sigma_3)$ is inconsistent. Sets of
+$(\sigma_1,\sigma_2)(\sigma_3,\sigma_3)$ is consistent only if $\sigma_2 = \sigma_3$. Sets of
 consistent words are essentially \emph{Aczel
@@
 $(\sigma_1,\sigma_1)(\sigma_1,\sigma_2)(\sigma_2,\sigma_3)$ is
-consistent. This word is highlighted with solid arrows in the
-diagram above. Now if we shuffle the single state pair $(\sigma_2,
-\sigma_3)$ into the above language, we might end up with a language as
-below:
+consistent. This word is highlighted with solid arrows in the diagram
+above. Now if we shuffle the single state pair $(\sigma_2, \sigma_3)$
+into the above language, would end up with a language containing the
+words represented in the diagram below:
 
@@
 
-With mumbling we now have that
+Assuming we apply mumbling to both sides of the following equation, we have that
 \begin{align*}
@@
 \edn(P); \test{Q} \le_\pi \edn(P \cap Q), \qquad
-\test(P) \le \edn(Q),\\
+\test(P) \le \edn(P),\\
 \text{range}(\Id_P \circ R) \le P \implies \edn(P); \langle R\rangle^\star \le_\pi \edn(P).
@@
 where $x \leftarrow v$ denotes the atomic command which assigns the
-value $v$ to $x$. The $\eval$ function evaluates an expression $e$ in
+value $v$ to $x$. The $\eval$ function atomically evaluates an expression $e$ in
 the state $\sigma$. Using this definition we derive the assignment
@@
 $f$ is the length of the array. In other words, either we find the
-least element, or no elements in the array satisfy $P$, in which case
-$f$ remains the same as the length of the array.
+least element, or $f$ remains the same as the length of the array, in
+which case no elements in the array satisfy $P$.
 
@@
 and verification tasks. In line with this aim, we have applied
-our formalisation to two simple examples.
+our formalisation to a simple example program.
 
@@
 
-Algebra plays an important role in our development. First, it
-allowed us to derive inference rules rapidly and with little proof
+Algebra plays an important role in our development. First, it allowed
+us to derive inference rules rapidly and with little proof
 effort. Second, it yields an abstract layer at which many properties
@@
 rely-guarantee reasoning can be implemented in interactive theorem
-proving tools such as Isabelle.
+proving tools such as Isabelle. However, we have not yet
+investigated the extent to which such decision procedures would benefit
+our approach.
 
